---
title: "(ST 411/511) Lab 1 - Intro to R"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---
## Hello!
- My office hours
  - Tuesday, 2:00pm - 3:00pm
  - Friday, 11:00am - 12:00pm
  
## Why show up for lab?
- The lab material is going to be directly applicable to homework assignments
- Great time to get assistance with homework
- Reinforcement of course material
- Programming assistance

:::aside
If no one shows I'll be super sad...
:::

## Objectives {transition="zoom" transition-speed="slow"}
- Introduce RStudio
- Install and load R packages
- Produce some simple graphs (histogram, boxplot, dotplot)
- Calculate sample statistics (mean, median, standard deviation, etc.)
- Perform a t-test
- Knit an R Markdown document

## Getting R and RStudio
- If using personal computer
  - Instructions to install are provided in the Lab 1 PDF, step 1
  - Can also use Citrix Workspace
  - Or Posit Cloud
- If using lab computer
  - R and RStudio should both be installed and ready for use
  
## Loading Sleuth3 package {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## install.packages("Sleuth3")
library(Sleuth3) # Load the Sleuth3 package
```

## Viewing a Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
head(case0102)
```

## Viewing a Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0102
```

## Viewing a Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# The View() function allows the dataset to be viewed in its own tab 
View(case0102)

# The help() function will provide additional details (if available)
help(case0102)
help(package = Sleuth3)
```

## Loading ggplot2 package {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## install.packages("ggplot2")
library(ggplot2) # Load the ggplot2 package
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(case0102$Salary)
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(case0102$Salary, geom = "histogram")
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, qplot(Salary, geom = "histogram"))
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Salary, data = case0102, geom = "histogram")
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, qplot(Salary[Sex == "Male"], geom = "histogram"))
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, qplot(Salary[Sex == "Female"], geom = "histogram"))
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, qplot(Salary[Sex == "Female"], 
                     geom = "histogram",
                     breaks = c(3800, 4200, 4600, 5000, 5400, 5800, 6200, 6600)))
```

## Learning About a Function {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# The help() function will provide additional details of a function
help(qplot)
?qplot
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Sex, Salary, data = case0102, geom = "boxplot")
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Sex, Salary, geom = "boxplot", data = case0102)
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Salary, Sex, data = case0102, geom = "boxplot")
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(y = Salary, x = Sex, data = case0102, geom = "boxplot")
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
ggplot(case0102, aes(x = Sex, y = Salary)) + 
  geom_boxplot()
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
ggplot(case0102, aes(x = Sex, y = Salary)) + 
  geom_boxplot(aes(fill = Sex))
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
ggplot(case0102, aes(x = Sex, y = Salary)) + 
  geom_boxplot(aes(fill = Sex)) +
  ggtitle("Starting Salaries")
```

## Plotting in ggplot2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
ggplot(case0102, aes(x = Sex, y = Salary)) + 
  geom_boxplot(aes(fill = Sex)) +
  labs(title = "Starting Salaries",
       y = "Salary (US$)")
```

## Barplots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, boxplot(Salary ~ Sex))
```

## Stem-and-Leaf Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, stem(Salary[Sex == "Male"]))
with(case0102, Salary[Sex == "Male"])
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, summary(Salary[Sex == "Female"]))
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, sd(Salary[Sex == "Female"]))
with(case0102, length(Salary[Sex == "Female"]))
```

## Summary() Function {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, summary(Sex))
```

## Typing of Variable {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0102, is.factor(Sex))
```

## Typing of Variable {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
str(case0102$Sex)
```

## t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# One-sided two-sample equal variance t-test
t.test(Salary ~ Sex, alternative = "less", data = case0102, var.equal = TRUE)
```

## R-Markdown
- Allows text, code, and corresponding output to be compiled in a single document
- Steps to get started are outlined in the Lab 1 PDF

## Homework 1
- I would advise we use the last portion of lab to get started on the computation portions of the homework