---
title: "(ST 411/511) Lab 4 - Tons of Terrific? Tests!"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---
## Homework 2 - Question 1(b)(iv)
- To describe the scope of inference, the following 2 components are needed
  - Causality
    - To assess causality we are looking to see if the *treatments have randomly assigned* (or subjects randomly assigned to treatments)
  - Inference to a larger population
    - We are seeing if we have a *random sample* from the population of interest, if not we can only make inference on the sample collected

## Homework 2 - Question 2
- Question (a): Describe what is meant by the sampling distribution of the sample variance
- Question (b): Displayed a standard normal distribution and asked "Could this distribution be the sampling distribution of the sample variance, when sampling from some population? Why or why not?
- Shiny Time!

## Objectives {transition="zoom" transition-speed="slow"}
- Do a Welch’s t-test
- Perform a Wilcoxon rank-sum test
- Perform a sign test

## Loading packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Dataset
- `case0201` (Finch Beak Data)
  - 178 observations
  - 2 variables
    - `Year`: Year the finch was caught, 1976 or 1978
    - `Depth`: Beak depth of the finch (mm)
    
## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0201
```
    
## Welch's $t$-test {auto-animate="true" auto-animate-easing="ease-in-out"}
- The Welch's $t$-test (also known as the unequal variance two-sample $t$-test) is performed when the equal variance assumption is not reasonable

:::aside
In practice, this test should be performed over the equal-variance version unless you are very certain the population variances are equal
:::

## Welch's $t$-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Default is unequal variance (not assuming equal variance)
t.test(Depth ~ Year, data = case0201)
```

:::aside
In practice, this test should be performed over the equal-variance version unless you are very certain the population variances are equal
:::

## Welch's $t$-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
t.test(Depth ~ Year, data = case0201, var.equal = FALSE,
       alternative = "two.sided")
```

:::aside
In practice, this test should be performed over the equal-variance version unless you are very certain the population variances are equal
:::

## [Comparing the two two-sample t-tests side-by-side]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| scroll: FALSE
t.test(Depth ~ Year, data = case0201, var.equal = TRUE,
       alternative = "two.sided")
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| scroll: FALSE
t.test(Depth ~ Year, data = case0201, var.equal = FALSE,
       alternative = "two.sided")
```
:::

::::

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t_{U}(\mu_{0}) & = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}} \\[0.25cm]
\end{aligned}
$$

:::aside
$\bar{X}$ and $\bar{Y}$ are sample mean estimations for $\mu_{X}$ and $\mu_{Y}$ respectively
:::

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t_{U}(0) = \dfrac{(\bar{X} - \bar{Y})}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}}
\end{aligned}
$$

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
**Welch-Satterthwaite Approximation**: to the null distribution of $t(\mu_{0})$
$$
\nu = \dfrac{\left(\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}\right)^{2}}{\dfrac{s^{4}_{X}}{m^{2}(m-1)} + \dfrac{s^{4}_{Y}}{n^{2}(n-1)}}
$$
Note: $\min(m-1, n-1) \le \nu \le m+n-2$

:::aside
Note: $\nu$ are the degrees of freedom
:::

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
**Confidence Interval:**
$$
(\bar{X} - \bar{Y}) \pm t_{(\nu)(1-\frac{\alpha}{2})}\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}
$$

:::aside
Remember, unless specified otherwise, R will make group 1 the one that comes alphabetically first
:::

## [Behind the Scenes of 't.test' function (Equal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
$$
t_{E}(\mu_{0}) = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s_{p}^{2}}{m} + \dfrac{s_{p}^{2}}{n}}} = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{s_{p}^{2}\left(\dfrac{1}{m} + \dfrac{1}{n}\right)}}
$$

## [Behind the Scenes of 't.test' function (Equal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
$$
\begin{align*}
s_{p}^{2} & = \dfrac{\sum\limits_{i=1}^{m}(X_{i} - \bar{X})^{2} + \sum\limits_{i=1}^{n}(Y_{i} - \bar{Y})^{2}}{(m - 1) + (n - 1)} \\[0.25cm]
& = \dfrac{(m-1)s^{2}_{X} + (n-1)s^{2}_{Y}}{(m+n-2)}
\end{align*}
$$

:::aside
We use the pooled methods if we assume the two groups have equal variance (equal standard deviations)
:::

## [Behind the Scenes of 't.test' function (Equal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
$$
(\bar{X} - \bar{Y}) \pm t_{(m+n-2)(1-\frac{\alpha}{2})}\sqrt{s_{p}^{2}\left(\dfrac{1}{m} + \dfrac{1}{n}\right)}
$$

:::aside
Notice the degrees of freedom = $m + n - 2$
:::

## [Behind the Scenes of 't.test' function (Summary)]{.r-fit-text} {transition="slide" transition-speed="slow"}

- **Unequal Variance (un-pooled)**
  - No assumption of equal variance
  - Degrees of freedom calculated with complicated formula (R will do this for you)
  - `var.equal = FALSE` (R's default)
- **Equal Variance (pooled)**
  - Assumption of equal variance
  - Degrees of freedom: $m + n - 2$ 
    - = Group 1 sample size + Group 2 sample size - 2
  - `var.equal = TRUE`
  
## Wilcoxon Rank-Sum Test
- The two-sample version of the Wilcoxon Signed-Rank Test
- A non-parametric test
  - Meaning that a parametric assumption is not necessary (no normality assumption needed)
- Known as a permutation test
  - Creation of a null distribution using re-sampling techniques on observed data
  
## Dataset
- `case0402` (Cognitive Load)
  - 28 observations
  - 3 variables
    - `Time`: the time (in seconds) that the student worked on the problem
    - `Treatment`: factor variable with two levels ("Modified" and "Conventional")
    - `Censored`: 1 if the individual did not complete the problem in 5 minutes, 0 if they did

## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0402
```

## Rank Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0402$Time
```

## Rank Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
rank(case0402$Time)
```

## Rank Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Side by side
# We can see that values that are tied - the corresponding rank is averaged
cbind(case0402$Time, rank(case0402$Time))
```

## Rank Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# A bit easier to see the averaging of tied ranks when ordered
cbind(sort(case0402$Time), sort(rank(case0402$Time)))
```

## Side-by-Side Boxplot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Where is the conventional upper whisker?
qplot(Treatment, Time, data = case0402, geom = "boxplot")
```

## Facet Histograms {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Time, data = case0402, geom = "histogram") +
  facet_grid(Treatment ~ .)
```

## Facet Histograms {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Time, data = case0402, geom = "histogram") +
  facet_grid(. ~ Treatment)
```

## Facet Histograms {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Time, data = case0402, geom = "histogram") +
  facet_grid(Treatment ~ Censored)
```

## Facet Histograms {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Custom histogram 
ggplot(case0402, aes(x = Time, fill = Treatment)) +
  geom_histogram(color = "black", bins = 30, alpha = 0.7) +  # Custom bin count and transparency
  scale_fill_brewer(palette = "Set2") +  # Use a color palette
  facet_grid(Treatment ~ .) +  # Facet by Treatment
  labs(title = "Facet Histograms by Treatment", 
       x = "Time", 
       y = "Frequency") +
  theme_minimal(base_size = 15) +  # Clean, minimal theme
  theme(
    axis.ticks.y = element_line(),  # Revert to default y-axis ticks
    axis.ticks.length.y = unit(0.15, "cm"),  # Default length for y-axis ticks
    axis.text.y = element_text(size = 10, color = "black"),  # Default size and color for y-axis labels
    strip.text = element_text(size = 14, face = "bold"),  # Bold facet labels
    plot.title = element_text(hjust = 0.5)  # Center plot title
  )
```

## Wilcoxon Rank-Sum Test
$$
H_{0}: \delta = 0 \quad\quad\quad H_{1}: \delta > 0 
$$

- **Research Question:** 
  - Whether the "Modified" group had lower times than the "Conventional" group
- Notice I use $\delta$ notation here because we're not actually comparing mean times like $\mu$ often denotes

## Wilcoxon Rank-Sum Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
wilcox.test(Time ~ Treatment, data = case0402, alternative = "greater",
            exact = FALSE, correct = FALSE)
```

## Wilcoxon Rank-Sum Test {auto-animate="true" auto-animate-easing="ease-in-out"}
- `exact = FALSE`: indicates that the normal approximation is being used
  - When setting to TRUE indicating the want of an exact $p$-value can fail due to permutation being too computationally intensively or if any ties
- `correct = FALSE`: indicates that a continuity correction is not wanted (in the normal approximation to $p$-value)

## Wilcoxon Rank-Sum Test
$$
H_{0}: \delta = 0 \quad\quad\quad H_{1}: \delta < 0 
$$

## Wilcoxon Rank-Sum Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# If we wanted to change which group is denoted at "Group 1"
with(case0402, wilcox.test(x = Time[Treatment == "Modified"],
                          y = Time[Treatment == "Conventional"],
                          alternative = "less", correct = FALSE, exact = FALSE))
```

## [Wilcoxon Rank-Sum Test (Assumptions)]{.r-fit-text}
- If one assumes the only difference between populations is an additive shift then it can test whether that shift is 0

## [Wilcoxon Rank-Sum Test (Assumptions)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
sim.data <- data.frame(x = rnorm(100000, mean = 0, sd = 1))
ggplot(data = sim.data) + 
  geom_histogram(mapping = aes(x = x), fill = "#cc77f7", color = "black", binwidth = 0.05, alpha = 0.50) +
  geom_histogram(mapping = aes(x = x + 2), fill = "#8af777", color = "black", binwidth = 0.05, alpha = 0.50) +
  labs(title = "Only Difference Between Distributions is an Additive Shift",
       x = "Values",
       y = "Frequency",
       caption = expression(paste("Shift difference equal to ", delta, " equal to 2 in this example"))) +
  theme_bw()
```

## [Wilcoxon Rank-Sum Test (Confidence Interval)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
wilcox.test(Time ~ Treatment, data = case0402, exact = FALSE, correct = FALSE,
            conf.int = TRUE)
```

## Sign Test
- A test of  medians (or any other percentile)
  - Under the null hypothesis, the distribution is binomial
  - If we have a null $H_{0}: M = 0$, we can perform a two-sample binomial test with $p_{0}=0.50$
- **Research Question:** Is the left hippocampus volume larger in the unaffected twin than in the affected twin?

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
head(case0202)
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Back at it again calculating differences
diffs <- with(case0202, Unaffected-Affected)
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Back at it again calculating differences
diffs <- with(case0202, Unaffected-Affected)
diffs
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
length(diffs)
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
which(diffs > 0) # How many pairs?
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
length(which(diffs > 0)) # How many of these are greater than 0?
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Performs exact test
binom.test(x = 14, n = 15, p = 0.5, alternative = "greater") 
```

