---
title: "(ST 411/511) Lab 3 - Data Transformations"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide        # Options: 'slide', 'fade', 'convex', 'concave', 'zoom'
    transition-speed: slow
    auto-animate: TRUE
    auto-animate-easing: ease-in-out
    chalkboard: TRUE
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js  # Enable zooming in if needed
      - reveal.js/plugin/highlight/highlight.js  # Enable syntax highlighting for code
---

##### Add something to clarify median interpretation, maybe some algebra or something?

## Objectives {transition="zoom" transition-speed="slow"}
- Do log and logit transformations
- Locate data on a plot
- Perform a t-test with specified data excluded
  
## Loading packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Today's Dataset
- `case0301` (Cloud seeding case)
  - 52 observations
  - 2 variables
    - `Rainfall`: the volume of rainfall in the target area (in acre-feet)
    - `Treatement`: a factor with levels "Unseeded" and "Seeded" indicating whether the clouds were unseeded or seeded.

## Today's Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0301
```

## Today's Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Some exploratory functions when looking at a new dataset
names(case0301)
```

## Today's Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Some exploratory functions when looking at a new dataset
summary(case0301)
```

## Today's Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Some exploratory functions when looking at a new dataset
str(case0301)
```

## t-test Assumptions
- Populations are normally-distributed
- Given there are two populations, they have equal variance
- Observations are statistically independent

:::aside
Answer alert! Homework 3 - Question 5
:::

## [t-test Assumptions (slightly more detailed)]{.r-fit-text}
- Populations are normally-distributed
  - Normality assumption OR if the sample size is large enough CLT will lead us to approximate normality
- Given there are two populations, they have equal variance
  - Equal variance assumption is only needed in the pooled (equal variance) two-sample t-test
- Observations are statistically independent
  - Independence between group (unless paired t-test) is also needed

## Chatting About Boxplots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Treatment, Rainfall, data = case0301, geom = "boxplot")
```

## Chatting About Boxplots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
ggplot(data = case0301, aes(x = Treatment, y = Rainfall)) +
  geom_boxplot(aes(fill = as.factor(Treatment))) +
  labs(title = "Rainfall Volume for Unseeded and Seeded Clouds",
       x = "Treatment Group",
       y = "Rainfall (acre-feet)",
       fill = "Treatment") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = c(0.85, 0.75),
        legend.key.size = unit(2, 'cm'),
        plot.title = element_text(size = 18))
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE          # Hide the code
#| fig-width: 16        # Adjust figure width (aspect ratio)
#| fig-height: 9        # Adjust figure height (aspect ratio)
#| fig-align: center    # Center the plot
#| out.width: "100%"    # Full width of the slide
#| out.height: "90vh"   # 90% of viewport height to avoid scrollbars
#| dev.args: list(res = 300)  # High resolution for clarity

# Generate the plot
ggplot(data = case0301, aes(x = Treatment, y = Rainfall)) +
  geom_boxplot(aes(fill = as.factor(Treatment))) +
  labs(title = "Rainfall Volume for Unseeded and Seeded Clouds",
       x = "Treatment Group",
       y = "Rainfall (acre-feet)",
       fill = "Treatment") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = c(0.85, 0.75),
        legend.key.size = unit(2, 'cm'),
        plot.title = element_text(size = 18))
```

## Normal Distributions
- When we plot the data like we just did, we're looking to assess our assumptions, specifically the normality assumption
  - But, what does data from a Normal distribution look like?

:::aside
Technically, in a confirmatory study, exploratory plotting like this should be done at the end to prevent data-snooping
:::

## Normal Distributions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating (standard) Normal data (100 Observations)
da.data <- data.frame(vals = rnorm(100, mean = 0, sd = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#3fe0a5", col = "white", alpha = 0.8) +
  labs(title = "Histogram of Simulated Standard Normal Data",
       x = "",
       y = "Frequency",
       caption = "100 N(0,1) data points, binwidth = 0.25") +
  xlim(-4, 4) +
  scale_x_continuous(breaks = seq(-4, 4, length = 9)) +
  theme_bw()
```

## Normal Distributions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating (standard) Normal data (100 Observations)
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.10, fill = "#3fe0a5", col = "white", alpha = 0.8) +
  labs(title = "Histogram of Simulated Standard Normal Data",
       x = "",
       y = "Frequency",
       caption = "100 N(0,1) data points, binwidth = 0.10") +
  xlim(-4, 4) +
  scale_x_continuous(breaks = seq(-4, 4, length = 9)) +
  theme_bw()
```

## Normal Distributions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating (standard) Normal data (1000 Observations)
da.data <- data.frame(vals = rnorm(1000, mean = 0, sd = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#3fe0a5", col = "white", alpha = 0.8) +
  labs(title = "Histogram of Simulated Standard Normal Data",
       x = "",
       y = "Frequency",
       caption = "1000 N(0,1) data points, binwidth = 0.25") +
  xlim(-4, 4) +
  scale_x_continuous(breaks = seq(-4, 4, length = 9)) +
  theme_bw()
```

## Normal Distributions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating (standard) Normal data (1000 Observations)
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.10, fill = "#3fe0a5", col = "white", alpha = 0.8) +
  labs(title = "Histogram of Simulated Standard Normal Data",
       x = "",
       y = "Frequency",
       caption = "1000 N(0,1) data points, binwidth = 0.10") +
  xlim(-4, 4) +
  scale_x_continuous(breaks = seq(-4, 4, length = 9)) +
  theme_bw()
```

## Normal Distributions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating (standard) Normal data (10000 Observations)
da.data <- data.frame(vals = rnorm(10000, mean = 0, sd = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#3fe0a5", col = "white", alpha = 0.8) +
  labs(title = "Histogram of Simulated Standard Normal Data",
       x = "",
       y = "Frequency",
       caption = "10,000 N(0,1) data points, binwidth = 0.25") +
  xlim(-4, 4) +
  scale_x_continuous(breaks = seq(-4, 4, length = 9)) +
  theme_bw()
```

## Normal Distributions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating (standard) Normal data (10000 Observations)
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.10, fill = "#3fe0a5", col = "white", alpha = 0.8) +
  labs(title = "Histogram of Simulated Standard Normal Data",
       x = "",
       y = "Frequency",
       caption = "10,000 N(0,1) data points, binwidth = 0.10") +
  xlim(-4, 4) +
  scale_x_continuous(breaks = seq(-4, 4, length = 9)) +
  theme_bw()
```


## Normal Distributions
- A Normal distribution is unimodel and symmetric (very similar to the $t$-distribution)
- What if our data are not unimodel and symmetric?
  - Data transformations are an option
  
## Transforming Data
- A common data transformation is the log transformation
  - Often for right-skewed positive data
- One of the most abused data transformations

:::aside
To statisticians, only *log base e* exists. Also know as the natural log, denoted $\ln(\cdot)$. Why do we (they) use $\log(\cdot)$ notation instead? I have no idea, but don't worry I'll talk to the top dogs
:::

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text}
Testing hypotheses regarding population means on the original data can answer a different question than testing hypotheses regarding population means of transformed data since

$$
E(g(T)) \neq g(E(T))
$$

Transforming inference back to the original scale is very challenging to interpret unless strong assumptions are made

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text}
- When might be a good time to perform a log transformation?
  - If we have right skewed positive data that once transformed has a symmetric distribution
  - Or we are more interested in the results on the actual log scale

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating right-skewed positive data (log-normal data)
set.seed(1)
da.data <- data.frame(vals = rlnorm(10000, mean = 0, sd = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Log-Normal Data)") +
  xlim(0, 20) +
  theme_bw()
```

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Natural log of that data
ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
```

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
library(gridExtra)
p1 <- ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Log-Normal Data)") +
  xlim(0, 20) +
  theme_bw()

p2 <- ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Load the necessary libraries
# install.packages("reshape2")
library(reshape2)

# Data
dat.data <- cbind.data.frame(da.data, log(da.data))
colnames(dat.data) <- c("Y", "log(Y)")

# Reshape the data for ggplot
dat_long <- melt(dat.data, variable.name = "variable", value.name = "value")

# Create the boxplot with customization
ggplot(dat_long, aes(x = variable, y = value, fill = variable)) +
  geom_boxplot(outlier.color = "black", outlier.shape = 16, outlier.size = 1.25) +  # Customize outliers
  scale_fill_manual(values = c("Y" = "#eb34de", "log(Y)" = "#eb34de")) +       # Custom colors
  theme_minimal() +
  theme(
    axis.title = element_blank(),            # Remove x and y labels
    axis.text.x = element_text(size = 14),   # Increase x-axis text size
    axis.text.y = element_text(size = 12),   # Increase y-axis text size
    legend.position = "none",                # Remove legend
    panel.grid.minor = element_blank(),      # Remove minor grid lines
    plot.title = element_text(hjust = 0.5)   # Center the title
  ) +
  labs(title = "Boxplot of Y vs log(Y)")    # Title
```


## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- The right skewed data shown was log-normal so unless you assume that your data is log-normal you won't have completely Normal data once transformed
  - The couple following of slides will illustrate this
  
## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating right-skewed positive data (exponential data)
set.seed(1)
da.data <- data.frame(vals = rexp(10000, rate = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Exponential)") +
  xlim(0, 20) +
  theme_bw()
```

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Natural log of that data
ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
```

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Plotting side-by-side
p3 <- ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Exponential)") +
  xlim(0, 20) +
  theme_bw()
p4 <- ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
grid.arrange(p3, p4, ncol = 2)
```

## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- We can see that since our original data was not log-normal we don't end up with a symmetric distribution
- Notice that just because a distribution is unimodal and symmetric does not mean it’s normal
  - But for the purposes of this class we won't stress about it too much
- Also, if you have negative values you cannot take the natural log
    - Some people shift the data before log-transforming, then interpretation becomes challenging / impossible
    
## [Log-transformation Tangent / Soapbox Ted-Talk]{.r-fit-text .smaller} {auto-animate="true" auto-animate-easing="ease-in-out"}
- Be wary of transformations and if you're unsure you seek out a statistical consultant
- Consulting advertisement
  - Statistical advice to currently enrolled OSU graduate students on an OSU-related research project is offered without charge in weeks 2 through 10 of fall, winter and spring term.
- I'm not saying that transformation should never be used, they just need to be used thoughtfully

:::aside
This slide is brought to you by the OSU statistical consulting services 
:::

## Plotting Log-Transformed Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Treatment, log(Rainfall), data = case0301, geom = "boxplot")
```

:::aside
Homework 3 - Question 4: Hint: log(response variable) when plotting
:::

## [Comparing Box-Plots (Untransformed to Transformed)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
p3 <- ggplot(data = case0301, aes(x = Treatment, y = log(Rainfall))) +
  geom_boxplot(aes(fill = as.factor(Treatment))) +
  labs(title = "Rainfall Volume for Unseeded and Seeded Clouds (Transformed)",
       x = "Treatment Group",
       y = "Rainfall (acre-feet)",
       fill = "Treatment") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = c(0.90, 0.85),
        legend.key.size = unit(1.25, 'cm'),
        plot.title = element_text(size = 12))

p4 <- ggplot(data = case0301, aes(x = Treatment, y = Rainfall)) +
  geom_boxplot(aes(fill = as.factor(Treatment))) +
  labs(title = "Rainfall Volume for Unseeded and Seeded Clouds (Untransformed)",
       x = "Treatment Group",
       y = "Rainfall (acre-feet)",
       fill = "Treatment") +
  coord_flip() +
  theme_classic() +
  theme(legend.position = c(0.90, 0.85),
        legend.key.size = unit(1.25, 'cm'),
        plot.title = element_text(size = 12))
grid.arrange(p3, p4, nrow = 2)
```

## t-test Time! {auto-animate="true" auto-animate-easing="ease-in-out"}
- **Research Question**: Does cloud-seeding increase rainfall?
  - What are our hypotheses?

## t-test Time! {.smaller} 
$$
  H_{0}: \mu_{\text{Seeded}} - \mu_{\text{Unseeded}} = 0 \quad\quad\quad H_{1}: \mu_{\text{Seeded}} - \mu_{\text{Unseeded}} > 0
$$

where

- $\mu_{\text{Seeded}}$ is the population mean in the `Seeded` cloud group
- $\mu_{\text{Unseeded}}$ is the population mean in the `Unseeded` cloud group

## t-test Time! {auto-animate="true" auto-animate-easing="ease-in-out"}
- Notice the hypothesis are set up to match R's default coding order
  - `Seeded` is before `Unseeded` alphabetically 

## t-test Time! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| code-overflow: wrap
t.test(log(Rainfall) ~ Treatment, data = case0301, var.equal = TRUE, 
       alternative = "greater")
```

:::aside
Homework 3 - Question 7 & 8: Hint: log(response variable) when running t-test
:::

## Interpretation
- There is strong evidence that the **median** acre-feet of rain is greater for the seeded than the unseeded days (one-sided $p = 0.007041$ from a two-sample t-test)
- Why "median" and not mean?

## Interpretation
- Recall an earlier warning that "testing hypotheses regarding population means on the original data can answer a different question than testing hypotheses regarding population means of transformed data since"
  - In the case of the log-transformed data having a symmetric distribution, we can work out that the mean of our logged data is equal to the median of the log data

## Interpretation
Once the data has been back-transformed

$$
e^{\bar{X}_{1}} \quad\text{  estimates  }\quad \text{Median}(X_{1}) \\[0.25cm]
e^{\bar{X}_{1} - \bar{X}_{2}} \quad\text{  estimates  }\quad \left(\frac{\text{Median}(X_{1})}{\text{Median}(X_{2})}\right)
$$

:::aside
More information about this process can be found in The Statistical Sleuth in section 3.5.1
:::

## Confidence Interval Construction {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| code-overflow: wrap
t.test(log(Rainfall) ~ Treatment, data = case0301, var.equal = TRUE, 
       alternative = "greater")
```

## Confidence Interval Construction {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| code-overflow: wrap
t.test(log(Rainfall) ~ Treatment, data = case0301, var.equal = TRUE, 
       alternative = "two.sided")
```

:::aside
Not best practice to obtain two-sided confidence interval but for the purposes of this class...
:::

## Confidence Interval Construction {auto-animate="true" auto-animate-easing="ease-in-out"}
Exponentiation of the confidence interval

$$
\Big(0.240865 \,,\, 2.046697\Big) \quad\Rightarrow\quad \Big(e^{0.240865} \,,\, e^{2.046697}\Big) = 
$$

## Confidence Interval Construction {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
point.est <- 5.134187 - 3.990406
exp.point.est <- exp(point.est)
exp.point.est
```

## Confidence Interval Construction {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
CI.Bounds <- c(0.240865, 2.046697)
exp.CI <- exp(CI.Bounds)
exp.CI
```


## Logit Function
$$
\text{logit}(Y) = \ln\left(\frac{Y}{1-Y}\right)
$$

## Visual of Data Transformation
```{r}
#| echo: FALSE
set.seed(1)
df <- data.frame(x = runif(10000), logit = log(runif(10000) / (1-runif(10000))))
p1 <- ggplot(data = df, aes(x = x)) +
  geom_histogram(binwidth = 0.025, fill = "#50e2f2", col = "white", alpha = 0.8) +
  labs(x = "X",
       y = "Frequency",
       title = "Uniform Data") +
  theme_bw()

p2 <- ggplot(data = df, aes(x = logit)) +
  geom_histogram(binwidth = 0.25, fill = "#50e2f2", col = "white", alpha = 0.8) +
  labs(x = "Logit(X)",
       y = "Frequency",
       title = "Exact Same Data, Logit Transformed") +
  theme_bw()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Loading in Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Year 2/Fall 2024/ST 511 (TA)/Labs/Lab 3")
```

## Loading in Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Year 2/Fall 2024/ST 511 (TA)/Labs/Lab 3")
C_dalli <- read.csv("C_dalli.csv")
```

:::aside
One of many options to read in a .csv file (Homework 3 - Question 1)
:::

## Loading in Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
C_dalli
```

## Explortatory Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(as.factor(time_point), pct_cover, data = C_dalli, geom = "boxplot")
```

## Performing logit transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
logit_pct <- with(C_dalli, log((pct_cover) / (100 - pct_cover)))
logit_pct
```

## Performing logit transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(as.factor(time_point), logit_pct, data = C_dalli, geom = "boxplot")
```

## Comparing Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
p10 <- qplot(as.factor(time_point), pct_cover, data = C_dalli, geom = "boxplot") +
  labs(title = "Untransformed",
       x = "Time Point",
       y = "Percent Cover") +
  theme_bw()
p11 <- qplot(as.factor(time_point), logit_pct, data = C_dalli, geom = "boxplot") +
  labs(title = "Transformed",
       x = "Time Point",
       y = "Logit(Percent Cover)") +
  theme_bw()
grid.arrange(p10, p11)
```

## Case Study 3.1.2 - Dioxin Data {auto-animate="true" auto-animate-easing="ease-in-out"}
- 743 Observations
- 2 Variables
  - `Dioxin` :the concentration of TCDD, in parts per trillion
  - `Veteran`: factor variable with two levels, "Vietnam" and "Other", to indicate the type of veteran

## Case Study 3.1.2 - Dioxin Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
head(case0302)
```

## Case Study 3.1.2 - Dioxin Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
qplot(Veteran, Dioxin, data = case0302, geom = "boxplot")
```

## Removing Outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Using identify() function
# Base R plotting
plot(Dioxin ~ Veteran, data = case0302)
with(case0302, identify(x = Veteran, y = Dioxin))
```

## Removing Outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Getting R to tell us
with(case0302, which(Dioxin > 10))
```

## Two-Sample t-test
$$
H_{0}: \mu_{V} - \mu_{N} = 0, \quad\quad\quad H_{1}: \mu_{V} - \mu_{N} > 0
$$

- $\mu_{V}$ represents the population mean dioxin concentrations for Vietnam veterans
- $\mu_{N}$ represents the population mean dioxin concentrations for non-Vietnam veterans

## Two-Sample t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Verify how R is ordering groups
summary(case0302$Veteran)
```

## Two-Sample t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# t-test time!
t.test(Dioxin ~ Veteran, data = case0302, var.equal = TRUE, 
       alternative = "less")
```

## Two-Sample t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# t-test time! (without observation 646)
t.test(Dioxin ~ Veteran, data = case0302, var.equal = TRUE, 
       alternative = "less", subset = -646)
```

## Two-Sample t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# t-test time! (without observations 646 and 645)
t.test(Dioxin ~ Veteran, data = case0302, var.equal = TRUE, 
       alternative = "less", subset = -c(646, 645))
```