---
title: "(ST 411/511) Lab 2 <br> t-test Party"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---
## All Office Hours {.smaller}
<ul class="nonincremental">
  <li>My office hours (Weniger 253)
    <ul>
      <li>Tuesday, 2:00pm - 3:00pm</li>
      <li>Friday, 11:00am - 12:00pm</li>
    </ul>
  </li>
  <li>Other TA office hours (Weniger 253)
    <ul>
      <li>Monday, 3:00pm - 4:00pm</li>
      <li>Tuesday, 12:00pm - 1:00pm</li>
      <li>Wednesday, 2:00pm - 3:00pm</li>
      <li>Friday, 3:00pm - 4:00pm</li>
    </ul>
  </li>
  <li>Lisa's office hours (Kidder 82)
    <ul>
      <li>Monday, 11:00am - 12:00pm</li>
      <li>Wednesday, 11:00am - 12:00pm</li>
    </ul>
  </li>
</ul>
  
## Objectives {transition="zoom" transition-speed="slow"}
- Do a paired t-test with R’s `t.test()` function.
- Calculate the t-statistic and p-value by hand, with R’s assistance.
- Calculate a confidence interval by hand, using R as a calculator.
- Do a two-sample t-test using `t.test()`.
- Read in a data set from a comma-delimited text file

## Lab2.pdf and Lab2.R
- All the relevant code and details pertaining to Lab 2 are expressed within these two documents
  
## Loading packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Notice once installed, we only need to load to use now
library(Sleuth3)
library(ggplot2)
```

## Today's Dataset
- 15 observations
- 2 variables
  - `Uneffected`: volume of left hippocampus of unaffected twin (in $\text{cm}^{3}$)
  - `Affected`: volume of left hippocampus of affected twin (in $\text{cm}^{3}$)

## Today's Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
help(case0202)
```

## Today's Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0202
```

## Explortatory Histogram {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Histogram of differences
qplot(Unaffected - Affected, data = case0202, geom = "histogram")
```

<!-- ## Explortatory Histogram {auto-animate="true" auto-animate-easing="ease-in-out"} -->
<!-- ```{r} -->
<!-- #| echo: TRUE -->
<!-- #| output-location: fragment -->
<!-- # Another version of a histogram -->
<!-- ggplot(data = case0202, aes(x = (Unaffected - Affected))) +  -->
<!--   geom_freqpoly() + -->
<!--   labs(title = "Freqpoly Plot of Differences", -->
<!--        y = "Frequency", -->
<!--        x = "Difference (Unaffected - Affected) in Twins") + -->
<!--   theme_bw() -->
<!-- ``` -->

## Paired t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create a vector of differences (Unaffected - Affected)
diffs <- case0202$Unaffected - case0202$Affected
diffs <- with(case0202, Unaffected - Affected)

# Equivalent ways to do the same thing
```

## Paired t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create a vector of differences (Unaffected - Affected)
diffs
```

## Paired t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Histogram of differences (again)
qplot(diffs, geom = "histogram")
```

<!-- ## Paired t-test {auto-animate="true" auto-animate-easing="ease-in-out"} -->
<!-- ```{r} -->
<!-- #| echo: TRUE -->
<!-- #| output-location: fragment -->
<!-- # Another version of a histogram (again) -->
<!-- ggplot(data = as.data.frame(diffs), aes(x = diffs)) +  -->
<!--   geom_freqpoly() + -->
<!--   labs(title = "Freqpoly Plot of Differences", -->
<!--        y = "Frequency", -->
<!--        x = "Difference (Unaffected - Affected) in Twins") + -->
<!--   theme_bw() -->
<!-- ``` -->

## Paired t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Actually doing the t-test
t.test(diffs)
```

## Paired t-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# What else is being done by default?
t.test(x = diffs, mu = 0, alternative = "two.sided", conf.level = 0.95, paired = FALSE, var.equal = FALSE)
```

## Paired t-test
- Notice the `paired` argument is `FALSE` by default. How can we call it a paired t-test if we don't use the paired argument?
  - A two-sample paired t-test is mathematically equivalent to a one sample t-test on the differences
- Let's perform it another way and see if our results match

## Paired t-test V2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
with(case0202, t.test(Unaffected, Affected, paired = TRUE))
```

## [Comparing the two paired t-tests side-by-side]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
t.test(diffs)
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
with(case0202, t.test(Unaffected, Affected, paired = TRUE))
```
:::

::::

## [Make sure to use `paired' argument when performing paired t-tests]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
t.test(diffs)
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
with(case0202, t.test(Unaffected, Affected))
```
:::

::::

## Paired t-test `by hand'
$$
t(\delta_{0}) = \dfrac{(\bar{X} - \bar{Y}) - \delta_{0}}{\sqrt{\dfrac{s^{2}_{X}}{n} + \dfrac{s^{2}_{Y}}{n} - 2\dfrac{s_{XY}}{n}}} = \dfrac{\bar{D}- \delta_{0}}{\sqrt{\dfrac{s^{2}_{D}}{n}}}
$$

## Paired t-test `by hand'
$$
t(0) = \dfrac{(\bar{X} - \bar{Y})}{\sqrt{\dfrac{s^{2}_{X}}{n} + \dfrac{s^{2}_{Y}}{n} - 2\dfrac{s_{XY}}{n}}} = \dfrac{\bar{D}}{\sqrt{\dfrac{s^{2}_{D}}{n}}}
$$

:::aside
We typically test if the difference is non-zero therefore the $\delta_{0}$ will be erased
:::

## Paired t-test `by hand' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Ybar <- mean(diffs)  #Sample mean of the differences
s <- sd(diffs)       #Sample standard deviation
n <- length(diffs)   #Sample size
```

## Paired t-test `by hand' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Ybar    #Sample mean of the differences
s       #Sample standard deviation
n       #Sample size
```

## Paired t-test `by hand' {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
SE(\bar{D}) = \sqrt{\frac{s_{D}^{2}}{n}}
$$

```{r}
#| echo: TRUE
se_Ybar <- s/sqrt(n)    #Calculate the SE of the sample mean
se_Ybar
```

## Paired t-test `by hand' {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
t\text{-ratio} = \dfrac{\bar{D}}{\sqrt{\dfrac{s^{2}_{D}}{n}}}
$$

```{r}
#| echo: TRUE
t.ratio <- Ybar/se_Ybar
t.ratio
```

## Checking That it Matches R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.ratio
t.test(diffs)
```

## Probability Distribution Visualization {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
tt <- data.frame(t.val = seq(-4, 4, length.out = 100))
theme_set(theme_classic(base_size = 28))
ggplot(tt, aes(x = t.val, y = dt(t.val, 14))) + 
  geom_line(size = 1.1) + 
  ylab("Density") + 
  ggtitle("t Probability Density Function (pdf)")
```

```{r}
theme_set(theme_grey())
```

## Calculating $p$-values {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
pt(3.228928, 14)
```

## Visualization of $p$-value {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
t.dist <- data.frame(x = seq(-4, 4, by = 0.001),
                     y = dt(seq(-4, 4, by = 0.001), df = 14))
Crit.Vals <- c(-1,1)*t.ratio
wrong.tail <- subset(t.dist, x < Crit.Vals[2])
lower.tail <- subset(t.dist, x < Crit.Vals[1])
upper.tail <- subset(t.dist, x > Crit.Vals[2])

ggplot(data = t.dist, aes(x,y)) +
    geom_line(linewidth = 1.25) +
    geom_area(data = wrong.tail, fill = "lightblue") +
  labs(title = "One-Sided Lower p-value Visualization",
       y = "Density",
       x = "t-statistic") +
  theme_bw()
```

## Calculating $p$-values {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
1 - pt(3.228928, 14)
```

## Visualization of $p$-value {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
ggplot(data = t.dist, aes(x,y)) +
    geom_line(linewidth = 1.25) +
    geom_area(data = upper.tail, fill = "lightblue") +
  labs(title = "One-Side Upper p-value Visualization",
       y = "Density",
       x = "t-statistic") +
  theme_bw()
```

## Calculating $p$-values {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
2*(1 - pt(3.228928, 14))

2*(pt(3.228928, 14, lower.tail = FALSE))
```

## Visualization of $p$-value {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
ggplot(data = t.dist, aes(x,y)) +
    geom_line(linewidth = 1.25) +
    geom_area(data = lower.tail, fill = "lightblue") +
    geom_area(data = upper.tail, fill = "lightblue") +
  labs(title = "Two-sided p-value Visualization",
       y = "Density",
       x = "t-statistic") +
  theme_bw()
```

## Checking That it Matches R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.ratio
2*(1 - pt(3.228928, 14))
t.test(diffs)
```

## [Calculating 95% Confidence Interval `by hand']{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\bar{D} \pm t_{n-1}\left(1-\frac{\alpha}{2}\right)\sqrt{\dfrac{s^{2}_{D}}{n}}
$$


## [Calculating 95% Confidence Interval `by hand']{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Ybar - qt(0.975, df = 14)*se_Ybar
Ybar + qt(0.975, df = 14)*se_Ybar
```

## [Calculating 95% Confidence Interval `by hand']{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Another way
Ybar + c(-1,1)*qt(0.975, df = 14)*se_Ybar
```


## Checking That it Matches R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.ratio
2*(1 - pt(3.228928, 14))
Ybar - qt(0.975, df = 14)*se_Ybar
Ybar + qt(0.975, df = 14)*se_Ybar
t.test(diffs)
```

## Case Study 2.1.1. - Two-Sample $t$-test
- 178 observations
- 2 variables
  - `Year`: Year the finch was caught, 1976 or 1978
  - `Depth`: Beak depth of the finch (mm)
  
## Explortatory Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(factor(Year), Depth, data = case0201, geom = "boxplot")
```

## Explortatory Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
t.test(Depth ~ Year, data = case0201, var.equal = TRUE)
```

## One-Sided Alternative Hypothesis
$$
H_{0}: \mu_{1} - \mu_{2} = 0, \quad\quad\quad H_{1}: \mu_{1} - \mu_{2} < 0
$$

## One-Sided $t$-test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
t.test(Depth ~ Year, data = case0201, var.equal = TRUE, alternative = "less")
```

## Loading in Data for Homework 2 {.smaller}
```{mermaid}
flowchart TD
A{RStudio} -->|Recommended| B{R Script}
B -->|"Environment Panel (Top Right Panel)"| C1[Import Dataset Button]
--> C2("From Text (base)...") 
--> C3(Wa-La)
B --> D1["File Button (Top Left)"] 
--> D2(Import Dataset) 
--> D3("From Text (base)...") 
--> D4(Ta-Da)
B -->|Directly Coding it| E1[Set Working Directory] 
--> E2("'read.csv' function") 
--> E3(Woo-Hoo)
```

## Loading in Data for Homework 2 
```{r}
#| echo: TRUE
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Year 2/Fall 2024/ST 511 (TA)/Labs/Lab 2")
```

## Loading in Data for Homework 2 
```{r}
#| echo: TRUE
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Year 2/Fall 2024/ST 511 (TA)/Labs/Lab 2")
HW2Dat <- read.csv("bird_stress.csv", stringsAsFactors = TRUE)
```

## Loading in Data for Homework 2
```{r}
#| echo: TRUE
bird_stress <- read.csv("C:/Users/Ghcto/OneDrive/Desktop/School/Year 2/Fall 2024/ST 511 (TA)/Labs/Lab 2/bird_stress.csv")
```

## Loading in Data for Homework 2
```{r}
#| echo: TRUE
HW2Dat <- read.csv("C:/Users/Ghcto/OneDrive/Desktop/School/Year 2/Fall 2024/ST 511 (TA)/Labs/Lab 2/bird_stress.csv")
```

## Loading in Data for Homework 2
```{r}
#| echo: TRUE
HW2Dat
```

## Loading in Data for Homework 2
```{r}
#| echo: TRUE
str(HW2Dat)
```


