---
title: "(ST 411/511) Lab 5 <br> Ah Yeah! - ANOVA"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## Lab Objectives

- Perform an ANOVA F-test to evaluate if mean lifetimes differ across various diets (case study 5.1.1)
- Generate an ANOVA table for finch beak depth data (case study 2.1.1) and compare with previous t-test results.
- Compute the pooled standard deviation $s_{p}$.
- Write a confidence interval for the difference in population means (case study 5.1.1)

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Dataset
- `case0501` (Diet Restriction and Longevity)
- **Description**: Female mice were randomly assigned to six treatment groups to investigate whether restricting dietary intake increases life expectancy.
    
## Dataset
- 178 observations
- 2 variables:
  - `Diet`: Type of diet group
    - `NP`: Mice ate unlimited amount of nonpurified, standard diet.
    - `N/N85`: Mice fed normally before and after weaning. After weaning, ration was controlled at 85 kcal/week.
    - `N/R50`: Normal diet before weaning and reduced calorie diet (50 kcal/week) after weaning.
    - `R/R50`: Reduced calorie diet of 50 kcal/week both before and after weaning.
    - `N/R50 lopro`: Normal diet before weaning, restricted diet (50 kcal/week) after weaning, and dietary protein content decreased with age.
    - `N/R40`: Normal diet before weaning and reduced diet (40 kcal/week) after weaning.
  - `Lifetime`: Lifetime of each subject (in months)
    
## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0501
```

## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# To see how R orders the groups and some summary statistics
summary(case0501)
```

## Side-by-Side Boxplot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Side-by-side boxplots
qplot(Diet, Lifetime, data = case0501, geom = "boxplot")
```

## Side-by-Side Boxplot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
ggplot(case0501, aes(x = Diet, y = Lifetime)) +
  geom_boxplot(fill = "lightblue", color = "darkblue", outlier.color = "red", outlier.shape = 16) +
  labs(
    title = "Lifetime by Diet Group",
    x = "Diet Type",
    y = "Lifetime (months)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## ANOVA {auto-animate="true" auto-animate-easing="ease-in-out"}
- To compare means, we are unable to use $t$-tests here since there are more than 2 groups.
  - Lucky for us there exists a method to compare more than two population means
  - ANOVA can be thought about (loosely) as the generalization of the $t$-test to multiple groups $(> 2)$

## ANOVA Assumptions {auto-animate="true" auto-animate-easing="ease-in-out"}
- Independence
- Normality
- Constant (Equal) Variance

## ANOVA Hypotheses {auto-animate="true" auto-animate-easing="ease-in-out"}
- Null Hypothesis: All (6) population means are equal
- Alternative Hypothesis: At least one population mean is different than the others
  - Notice the language, *at least one*, means that only one has to be statistically different from the rest, **not** that all means must be different.
  
## ANOVA in R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Running it in R
case0501_aov <- aov(Lifetime ~ Diet, data = case0501)
```

## ANOVA in R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0501_aov
```

## ANOVA in R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Obtain the ANOVA table
anova(case0501_aov)
```

## ANOVA in R {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Equivalently
summary(case0501_aov)
```

## ANOVA Table {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
# Load necessary libraries
library(knitr)
library(dplyr)

# Get the summary of the ANOVA
anova_summary <- summary(case0501_aov)

# Extract the relevant part from the summary
anova_table <- as.data.frame(anova_summary[[1]])

# Optionally rename the columns for clarity
colnames(anova_table) <- c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)")

# Replace NAs with empty strings, round numeric values
anova_table <- anova_table %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), "", round(., 2))))

# Replace the top right cell (the last row, p-value) with 2.2e-16
anova_table[1, "Pr(>F)"] <- "2.2e-16"

# Use kable to display the table
kable(anova_table, caption = "ANOVA Summary Table", digits = 4)
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
F.Dist.Plot <- function(df1 = 5, df2 = 10, cooler = "#8a17e3") {
  # Generate a sequence of x values for plotting the F-distribution
  x_vals <- seq(0, 5, length.out = 100)

  # Calculate the density of the F-distribution at each x value
  f_density <- data.frame(
    x = x_vals,
    y = df(x_vals, df1, df2)
  )

  # Plot the F-distribution
  plop <- ggplot(f_density, aes(x = x, y = y)) +
    geom_line(color = cooler, size = 1) +
    labs(
      title = bquote("F-distribution with " ~ df[1] == .(df1) ~ "and" ~ df[2] == .(df2)),
      x = "F value",
      y = "Density"
      ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#333333"),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(size = 10, color = "#555555"),
      panel.grid.major = element_line(color = "gray85"),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "#f9f9f9", color = NA),
    )
  print(plop)
}
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Plotting our F-distribution in case0501
F.Dist.Plot(df1 = 5, df2 = 343)
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# We will typically have a higher denominator degrees of freedom
F.Dist.Plot(df1 = 5, df2 = 20, cooler = "#4287f5")
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# What happens if we increase the denominator degrees of freedom?
F.Dist.Plot(df1 = 5, df2 = 50, cooler = "#f54284")
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# What happens if we increase the numerator degrees of freedom?
F.Dist.Plot(df1 = 10, df2 = 20, cooler = "#f5a442")
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# What happens if we increase the numerator degrees of freedom?
F.Dist.Plot(df1 = 20, df2 = 20, cooler = "#42eff5")
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# What happens if we increase the both degrees of freedom?
F.Dist.Plot(df1 = 50, df2 = 50, cooler = "#42f54b")
```

## [2 Different Degrees of Freedom?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Alright, last one
F.Dist.Plot(df1 = 100, df2 = 100, cooler = "#ed6840")
```

## [Back to the t-test \*Insert Sunglasses Emoji\*]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.test(Depth ~ Year, data = case0201, var.equal = TRUE)
```

## Compare with ANOVA {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# ANOVA always makes the equal variance assumption
case0201_aov <- aov(Depth ~ Year, data = case0201)
```

## Compare with ANOVA {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(case0201_aov)
```

## Comparing test outputs {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| scroll: FALSE
t.test(Depth ~ Year, data = case0201, var.equal = TRUE)

anova(case0201_aov)
```


## Comparing test outputs {auto-animate="true" auto-animate-easing="ease-in-out"}
- We see the $p$-values are the same
- The test statistics are also the same since $(-4.5833)^2 = 21.00664$
  - A $t$-statistic (with $n-1$ degrees of freedom) squared is equal to an $F$-statistic with one numerator degrees of freedom and $n-1$ denominator degrees of freedom.


## Comparing test outputs {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\left(t_{n-1}\right)^{2} = F_{1, n-1}
$$

## Comparing test outputs {auto-animate="true" auto-animate-easing="ease-in-out"}
- Thus we can see that the ANOVA F-test is the same as a two-sided two-sample equal variance t-test when there are two groups
- When there are more than two groups, you can think of the ANOVA F-test as a generalization
of the two-sided two-sample t-test

## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\begin{align*}
s_{p}^{2} & = \frac{(n_{1}-1)s^{2}_{1} + (n_{2}-1)s^{2}_{2}}{(n_{1}+n_{2}-2)} \\[0.25cm]
s_{p} & = \sqrt{\frac{(n_{1}-1)s^{2}_{1} + (n_{2}-1)s^{2}_{2}}{(n_{1}+n_{2}-2)}}
\end{align*}
$$


## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculate pooled standard deviation
sp <- sqrt(((89-1)*1.0353309^2 + (89-1)*0.9064799^2) / (89 + 89 - 2))
```

## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
sp <- sqrt(((89-1)*1.0353309^2 + (89-1)*0.9064799^2) / (89 + 89 - 2))
sp
```

## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(case0201_aov)
```

## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(case0201_aov)
sp^2
```

## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(case0201_aov)
166.638/176
```

## Calculating $s_{p}$ {auto-animate="true" auto-animate-easing="ease-in-out"}
- $s_{p}^{2}$ is called:
  - pooled variance estimate
  - residual mean square
  - mean squared error (MSE)
- and estimates the population variance $\sigma^{2}$

## Confidence Intervals
$$
(\bar{Y}_{1} - \bar{Y}_{2}) \pm t_{(n_{1}+n_{2}-2)\left(1-\frac{\alpha}{2}\right)} \cdot s_{p}\sqrt{\left(\dfrac{1}{n_{1}} + \dfrac{1}{n_{2}}\right)}
$$

## Confidence Intervals
$$
(\bar{Y}_{i} - \bar{Y}_{j}) \pm t_{(n_{1}+n_{2}-2)\left(1-\frac{\alpha}{2}\right)} \cdot s_{p}\sqrt{\left(\dfrac{1}{n_{i}} + \dfrac{1}{n_{j}}\right)}
$$

## Confidence Intervals
- Notice the ladder equation allows for any two groups to be used
- Since we have the *constant variance* assumption it seems reasonable to use all the data to calculate the estimate for variance and use that for whichever pairwise comparison confidence interval
- Because we are using this $s_{p}^{2}$ estimate from the ANOVA table we will use its corresponding degrees of freedom (df) for the confidence interval df

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
Let's write a confidence interval for the difference between population mean lifetimes for the N/R50 and lopro diets.
```{r}
#| echo: TRUE
with(case0501, aggregate(Lifetime ~ Diet, FUN = mean))
```

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
with(case0501, aggregate(Lifetime ~ Diet, FUN = length))
```

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have also used good ol summary()
summary(case0501$Diet)
```

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
For the N/R50 and lopro diets, we are interested in groups 3 and 6, thus 
$$
\bar{Y}_{3} = 42.29718, \quad \bar{Y}_{6} = 39.68571, \quad n_{3} = 71, \quad n_{6} = 56
$$

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculating our critical value (notice we are using the residual dfs)
qt(0.975, df = 343)
```

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\bar{Y}_{3} - \bar{Y}_{6} \pm t_{343}\left(1-\frac{\alpha}{2}\right) \cdot \sqrt{s_{p}^{2}} \cdot \sqrt{\left(\dfrac{1}{n_{3}} + \dfrac{1}{n_{6}}\right)} \\[0.25cm]
42.29718 - 39.68571 \pm t_{343}\left(1-\frac{\alpha}{2}\right) * \sqrt{44.6} * \sqrt{\frac{1}{71} + \frac{1}{56}} \\[0.25cm]
$$


## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculate confidence interval for mean difference
lower_bound <- 42.29718 - 39.68571 - qt(0.975, 343) * sqrt(44.6) * sqrt(1/71 + 1/56)
```

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculate confidence interval for mean difference
lower_bound <- 42.29718 - 39.68571 - qt(0.975, 343) * sqrt(44.6) * sqrt(1/71 + 1/56)
upper_bound <- 42.29718 - 39.68571 + qt(0.975, 343) * sqrt(44.6) * sqrt(1/71 + 1/56)
```

## Confidence Intervals (`case0501`) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculate confidence interval for mean difference
lower_bound <- 42.29718 - 39.68571 - qt(0.975, 343) * sqrt(44.6) * sqrt(1/71 + 1/56)
upper_bound <- 42.29718 - 39.68571 + qt(0.975, 343) * sqrt(44.6) * sqrt(1/71 + 1/56)
c(lower_bound, upper_bound)
```
