---
title: "(ST 412/512) Lab 7 <br> Variable Selection"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
header-includes:
  - |
    <style>
      .reveal .barelysmaller {
          font-size: 0.90em;  /* Adjust as needed */
        }
      
      .reveal .smaller {
        font-size: 0.75em;  /* Adjust as needed */
      }

      .reveal .tiny-font {
        font-size: 0.5em;  /* Even smaller font size */
        white-space: nowrap;
        overflow-x: auto;  /* Allow horizontal scrolling */
      }
    </style>
---


## Lab Objectives
* Perform best subsets variable selection using $AIC$, $BIC$, $C_{p}$, and adjusted $R^{2}$
  * Using the `leap` package for sequential model selection
* Make a $C_{p}$ plot
* Use stepwise regression to select variables
* Calculate variance inflation factors

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
library(leaps) # regsubsets()
library(car) # vif()
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* `case1201` (State Average SAT Scores)
  * Data on the average SAT scores for US states in 1982 and possible associated factors.
  * 50 observations (rows)
  * 8 variables (columns)
    * `State`: US state
    * `SAT`: state averages of the total SAT (verbal + quantitative) scores
    * `Takers`: the percentage of the total eligible students (high school seniors) in the state who took the exam
    * `Income`: the median income of families of test–takers (in hundreds of dollars)
    * `Years`: the average number of years that the test–takers had formal studies in social sciences, natural sciences and humanities
    * `Public`: the percentage of the test–takers who attended public secondary schools
    * `Expend`: he total state expenditure on secondary schools (in hundreds of dollars per student)
    * `Rank`: the median percentile ranking of the test–takers within their secondary school classes

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create a scatterplot matrix without State names
pairs(case1201[, -1], )
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# The lab notes want to perform a log transformation due to the 
# nonlinear relationship
case1201$lTakers <- log(case1201$Takers)
```

## Odd Observation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Finding more about the large expend observation (State)
which.max(case1201$Expend)
```

## Odd Observation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Finding more about the large expend observation (State)
which.max(case1201$Expend)
case1201[29, ]
```

## Odd Observation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Removing State and Takers (not logged) variables at once
# also removing odd observation pointed out earlier
SATdata <- case1201[-29, c(-1, -3)]
```

## Best-Subsets Variable Selection {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
SAT_best <- regsubsets(SAT ~ ., data = SATdata, nbest = 7)
```

* What this code is doing
  * The first argument specifies the largest model you are willing to consider. Here, I specify my response is `SAT`, and the possible terms all other terms in the dataset, which is what the `~ .` notation implies.
  * The dataset is excluding Alaska which was done earlier.
  * `nbest` specifies how many models should be returned for each size; I want the best seven models.  
  * `method` states the methods for finding the best models:  
    * `"exhaustive"` considers all models (not a sequential procedure!). 
      * `"exhaustive"` is default.
    * Other options are `"forward"` and `"backward"` (these are sequential procedures).  
  * Another useful argument is `nvmax`, the largest size model you want to consider.  
    * By default, it is set at 8, so here we don’t need to change it.
    
## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Contents can be printed out as is
SAT_best <- regsubsets(SAT ~ ., data = SATdata, nbest = 7)
SAT_best
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Or using summary we can obtain more information
summary(SAT_best)
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
* What the summary output tells us is the ranking for each sized model
  * I believe by default adjusted $R^{2}$ is used to rank the models within group sizes 

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# This object can even be plotted
plot(SAT_best)
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Let's name this object
SAT_best_summary <- summary(SAT_best)
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What can we pull out of it?
SAT_best_summary <- summary(SAT_best)
names(SAT_best_summary)
```

## Best-Subsets Variable Selection {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
SAT_best_summary$which
```

* This is a logical matrix with a row for every model selected and a column for every variable, including the intercept. The entries in a row are TRUE if the corresponding variable is in the model and FALSE otherwise

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Turns the first row of that matrix into 1's and 0's
as.numeric(SAT_best_summary$which[1, ])
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Can sum to count the number of parameters in model
sum(as.numeric(SAT_best_summary$which[1, ]))
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Can do this easily for all rows with rowSums
rowSums(SAT_best_summary$which)
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Give it a name for later
p <- rowSums(SAT_best_summary$which)
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating a dateframe from our summary output of each model's statistics
models <- with(SAT_best_summary, data.frame(p, rss, adjr2, cp, bic))
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting
models <- with(SAT_best_summary, data.frame(p, rss, adjr2, cp, bic))
qplot(p, cp, data = models) 
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Adding like Cp = p
models <- with(SAT_best_summary, data.frame(p, rss, adjr2, cp, bic))
qplot(p, cp, data = models) +
  geom_abline(slope = 1, intercept = 0)
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Zooming in
models <- with(SAT_best_summary, data.frame(p, rss, adjr2, cp, bic))
qplot(p, cp, data = models, ylim = c(0, 10)) +
  geom_abline(slope = 1, intercept = 0)
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot syntax()
ggplot(data = models, aes(x = p, y = cp)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  coord_cartesian(ylim = c(0, 10))
```

## Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# If you wanted it to not look terrible
ggplot(data = models, aes(x = p, y = cp)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  coord_cartesian(ylim = c(0, 10)) +
  labs(title = bquote(paste("Mallows ", C[p], " Versus Number of Parameters (p)")),
       y = expression(C[p]),
       x = "p") + 
  theme_bw()
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What model has the smallest Cp?
which.min(models$cp)
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Finding that model
which.min(models$cp)
SAT_best_summary$which[which.min(models$cp), ]
```

## Information Criterion {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **AIC Calculation**  
  * The Akaike Information Criterion (AIC) is calculated as:  
    $$
    AIC = n \log \left(\frac{\text{RSS}}{n} \right) + 2(p + 1)
    $$
    where:  
    * $n$ is the number of observations.  
    * $\text{RSS}$ is the residual sum of squares.  
    * $p$ is the number of regression coefficients ($\beta$). 
* **BIC Calculation**  
  * The Bayesian Information Criterion (BIC) is given by:  
    $$
    BIC = n \log \left(\frac{\text{RSS}}{n} \right) + (p + 1) \log(n)
    $$
  * Like AIC, a **smaller value indicates a better model**.  
* **Adjusted $R^2$ Calculation**  
  * Adjusted $R^2$ accounts for model complexity:  
    $$
    R^2_{\text{Adj}} = 1 - \left(\frac{\text{RSS} / (n - p - 1)}{\text{TSS} / (n - 1)}\right)
    $$
    where $\text{TSS}$ is the total sum of squares.  
  * Unlike AIC and BIC, **larger values of adjusted $R^2$ indicate a better model**.

## Information Criterion {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **Key Considerations**  
  * AIC and BIC assume **normality** and are valid for **linear regression models**.  
  * The `AIC()` and `BIC()` functions in R include an additional constant $n \log(2\pi) + n$, but this does not affect **model comparison**.  
   * The `regsubsets()` function calculates BIC **differently** by subtracting a term based on the **intercept-only model**:  
    $$
    \begin{align*}
    \text{BIC}_{\text{regsubsets}} & = n \log \left(\frac{\text{RSS}}{n} \right) + (p + 1) \log(n) - n \log \left(\frac{\text{RSS}_0}{n} \right) + \log(n) \\[0.25cm]
    & = \text{BIC}_{\text{Sleuth}} - n \log \left(\frac{\text{RSS}_0}{\text{RSS}} \right) + \log(n)
    \end{align*}
    $$
    where $\text{RSS}_0$ is the residual sum of squares for the **intercept-only model**.  
  * **This means that `regsubsets()` produces values that are**  
    $$ n \log(2\pi) + n $$  
    **less than the values from the Sleuth’s formula.**  
* **Rule of Thumb for Model Selection**  
  * **BIC penalizes complexity more heavily** than AIC, leading to simpler models
  * The choice of formula does not change which model is selected, since the differences are consistent

## Information Criterion {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculating AIC as in the Sleuth
n <- nrow(SATdata)
models$aic <- with(models, n * log(rss / n) + 2 * (p + 1))
```

## Back to Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting AIC vs p
best_aic <- which.min(models$aic) # Which model has the lowest AIC?
qplot(p, aic, data = models) +
  geom_point(aes(x = p[best_aic], y = aic[best_aic]), color = 'red', size = I(2.5))
```

## Back to Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = models, aes(x = p, y = aic)) +
  geom_point() +
  geom_point(aes(x = p[best_aic], y = aic[best_aic]), color = 'red', size = I(2.5))
```

## Back to Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting BIC vs p
best_bic <- which.min(models$bic) # Which model has the lowest BIC?
qplot(p, bic, data = models) +
  geom_point(aes(x = p[best_bic], y = bic[best_bic]), color = 'red', size = I(2.5))
```

## Back to Plotting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = models, aes(x = p, y = bic)) +
  geom_point() +
  geom_point(aes(x = p[best_bic], y = bic[best_bic]), color = 'red', size = I(2.5))
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Finding model with highest Adjusted R-squared
which.max(models$adjr2)
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
SAT_best_summary$which[which.min(models$cp), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
SAT_best_summary$which[which.min(models$cp), ]
SAT_best_summary$which[which.min(models$aic), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
SAT_best_summary$which[which.min(models$cp), ]
SAT_best_summary$which[which.min(models$aic), ]
SAT_best_summary$which[which.min(models$bic), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
SAT_best_summary$which[which.min(models$cp), ]
SAT_best_summary$which[which.min(models$aic), ]
SAT_best_summary$which[which.min(models$bic), ]
SAT_best_summary$which[which.max(models$adjr2), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What is the best model according to R-squared?
SAT_best_summary$which[which.max(SAT_best_summary$rsq), ]
```

* We shouldn't use $R^{2}$ as a criteria since it will always favor the largest model, it doesn't adjust for the number of parameters like the other methods.

## Sequential Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Performing Stepwise regression
SAT_step <- regsubsets(SAT ~ ., data = SATdata,
                       nbest = 2, method = "seqrep")
```

## Sequential Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Lookin at it
SAT_step
```

## Sequential Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating a dateframe from our summary output of each model's statistics
models_step <- with(summary(SAT_step), data.frame(p = rowSums(which), rss, adjr2, cp, bic))
```

## Sequential Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculating AIC (Sleuth formula)
models_step$aic <- with(models_step, n * log(rss / n) + 2 * (p + 1))
```

## Sequential Variable Selection {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **Stepwise Process**  
  * **Start** with an empty model or a full model, depending on the implementation.
  * **Add variables** that improve the model based on a chosen criterion (e.g., $p$-value, AIC, BIC).
  * After adding a variable, check whether any existing variables should be **removed** based on the same criterion.
  * The process stops when no more variables can be added or removed without significantly affecting the model.

* **Criteria for Selection**  
  * **$p$-values**: Variables with $p$-values greater than a threshold (usually 0.05) are candidates for removal.
  * **AIC/BIC**: Both are penalized for complexity, with the goal to minimize either criterion.
  * **Adjusted $R^2$**: A higher adjusted $R^2$ indicates a better model fit, accounting for the number of predictors.

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
summary(SAT_step)$which[which.min(models_step$cp), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
summary(SAT_step)$which[which.min(models_step$cp), ]
summary(SAT_step)$which[which.min(models_step$aic), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
summary(SAT_step)$which[which.min(models_step$cp), ]
summary(SAT_step)$which[which.min(models_step$aic), ]
summary(SAT_step)$which[which.min(models_step$bic), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# "Best" model for each criteria
summary(SAT_step)$which[which.min(models_step$cp), ]
summary(SAT_step)$which[which.min(models_step$aic), ]
summary(SAT_step)$which[which.min(models_step$bic), ]
summary(SAT_step)$which[which.max(models_step$adjr2), ]
```

## "Best" Models {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating most popular model selected
popular_model <- lm(SAT ~ Years + Expend + Rank + lTakers, data = SATdata)
```

## VIFs {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **Variance Inflation Factor (VIF)**  
  * The **Variance Inflation Factor (VIF)** is used to assess **multicollinearity** between predictor variables in a regression model.
  * VIF quantifies how much the variance of a regression coefficient is inflated due to its correlation with other predictors.

* **Formula for VIF**  
  * The formula for the VIF of a predictor variable $X_j$ is:
    $$
    \text{VIF}(X_j) = \frac{1}{1 - R_j^2}
    $$
    where:
    * $R_j^2$ is the **coefficient of determination** (i.e., $R^2$) obtained by regressing predictor $X_j$ on all other predictors.

* **Steps to Calculate VIF**  
  1. **Fit a linear regression model** where $X_j$ (the predictor of interest) is the dependent variable, and all other predictors are the independent variables.
  2. **Compute $R_j^2$** from the regression model.
  3. **Calculate VIF** using the formula above.
  * Or use `R`, easy peasy

* **Rules of Thumb for Interpreting VIF**  
  * $\text{VIF} = 1$: No correlation between the predictor $X_j$ and the other predictors. No multicollinearity.
  * $1 < \text{VIF} < 5$: Moderate correlation, generally acceptable.
  * $\text{VIF} > 5$: High correlation, indicating significant multicollinearity.
  * $\text{VIF} > 10$: Very high correlation, suggesting severe multicollinearity that needs attention.

## VIFs {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculating VIFs
vif(popular_model)
```

## VIFs {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Looking at large VIF valued explanatory variables
qplot(Rank, lTakers, data = SATdata)
```

## VIFs {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# ggplot() synatx
ggplot(data = SATdata, aes(x = Rank, y = lTakers)) +
  geom_point()
```

