---
title: "(ST 412/512) Lab 8 <br> No-Way Two-Way ANOVA"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
header-includes:
  - |
    <style>
      .reveal .barelysmaller {
          font-size: 0.90em;
        }
      
      .reveal .smaller {
        font-size: 0.75em;
      }

      .reveal .tiny-font {
        font-size: 0.5em;
        white-space: nowrap;
        overflow-x: auto;  /* Allow horizontal scrolling */
      }
    </style>
---

## Lab Objectives
* Create an interaction plot
* Estimate contrasts

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
library(dplyr) # Data manipulation functions
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* `case1301` (Seaweed Grazers)
  * Something about seaweed?
  * 96 observations (rows)
  * 3 variables (columns)
    * `Cover`: percent of regenerated seaweed cover
    * `Block`: a factor with levels "B1", "B2", "B3", "B4", "B5", "B6", "B7" and "B8"
    * `Treat`: a factor indicating treatment, with levels "C", "f", "fF", "L", "Lf" and "LfF"

## Viewing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1301
```

## Viewing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Copying dataset with a different name
seaweed <- case1301

# Logit-transforming response variable
seaweed$logitCover <- with(case1301, log(Cover / (100 - Cover)))
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
with(seaweed, interaction.plot(Block, Treat, response = logitCover))
```

## Interaction Plot {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **`interaction.plot(x.factor, trace.factor, response, fun = mean, ...)`**  
  * Creates an interaction plot to visualize the interaction between two categorical variables on a numeric response.
* **Arguments:**
  * `x.factor` – The factor plotted on the x-axis.
  * `trace.factor` – The factor used to create different traces (lines) in the plot.
  * `response` – The numeric variable whose mean (or another function) is plotted.
  * `fun` – A function to summarize the response variable (default: `mean`).
  * `type` – Type of plot (default: `"l"` for lines, but can be `"p"` for points, `"b"` for both).
  * `col` – Colors for different lines.
  * `lty` – Line types.
  * `pch` – Plotting characters for points.
  * `xlab`, `ylab`, `main` – Axis labels and title.
* We could also make an interaction plot by manipulating the data with `dplyr`

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
seaweed1 <- group_by(seaweed, Block, Treat)
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
seaweed1 <- group_by(seaweed, Block, Treat)
seaweed1
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
seaweed2 <- summarize(seaweed1, Trt_mean = mean(logitCover))
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
seaweed2 <- summarize(seaweed1, Trt_mean = mean(logitCover))
seaweed2
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have done this in one pipe operation as
seaweed %>%
  group_by(Block, Treat) %>%
  summarize(Trt_mean = mean(logitCover)) -> seaweed2
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have done this in one pipe operation as
seaweed %>%
  group_by(Block, Treat) %>%
  summarize(Trt_mean = mean(logitCover)) -> seaweed2
seaweed2
```


## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
ggplot(data = seaweed2, aes(x = Block, y = Trt_mean, shape = Treat)) +
  geom_point() + 
  geom_line(aes(group = Treat))
```

## Interaction Plot {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Some additions
ggplot(data = seaweed2, aes(x = Block, y = Trt_mean, shape = Treat, color = Treat)) +
  geom_point() + 
  geom_line(aes(group = Treat, color = Treat)) +
  labs(y = "Treatment Mean") +
  theme_bw()
```

## Contrasts {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **Estimating Contrasts for Seaweed Grazer Effects**  
  * Contrasts help quantify the effects of different grazers by forming linear combinations of treatment means.

* **Defined Contrasts:**  
  * **Large Fish:**  
  $$ \frac{1}{2} \mu_{LfF} - \frac{1}{2} \mu_{Lf} + \frac{1}{2} \mu_{fF} - \frac{1}{2} \mu_{f} $$  
  * **Small Fish:**  
  $$ \frac{1}{2} \mu_{Lf} - \frac{1}{2} \mu_{f} + \frac{1}{2} \mu_{f} - \frac{1}{2} \mu_{C} $$  
  * **Limpets:**  
  $$ \frac{1}{3} \mu_{LfF} - \frac{1}{3} \mu_{fF} + \frac{1}{3} \mu_{Lf} - \frac{1}{3} \mu_{f} + \frac{1}{3} \mu_{L} - \frac{1}{3} \mu_{C} $$  
  * **Limpets × Small Fish Interaction:**  
  $$ \frac{1}{2} \mu_{LfF} - \frac{1}{2} \mu_{fF} + \frac{1}{2} \mu_{Lf} - \frac{1}{2} \mu_{f} - \mu_L + \mu_C $$  
  * **Limpets × Large Fish Interaction:**  
  $$ \mu_{LfF} - \mu_{fF} - \mu_{Lf} + \mu_{f} $$  

* **Contrast Coefficients:**  
  * Contrasts are linear combinations of means where the sum of the coefficients is 0.
  * Example for the **Large Fish** contrast:
    * Coefficients: $$ \frac{1}{2}, -\frac{1}{2}, \frac{1}{2}, -\frac{1}{2}, 0, 0 $$
    * Corresponding means: $$ \mu_{LfF}, \mu_{Lf}, \mu_{fF}, \mu_{f}, \mu_{L}, \mu_{C} $$

## Contrasts {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* **Estimating Contrasts:**  
  * One approach is to compute sample means and sizes, then apply Chapter 6 formulas:  
    * **Point Estimate:**  
      $$
      \hat{g} = C_1 \bar{Y}_1 + \dots + C_I \bar{Y}_I
      $$
    * **Standard Error:**  
      $$
      SE(\hat{g}) = \hat{\sigma} \sqrt{ \frac{C_1^2}{n_1} + \dots + \frac{C_I^2}{n_I} }
      $$

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(seaweed$Treat)
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(seaweed$Treat)
Means <- with(seaweed, unlist(lapply(split(logitCover, Treat), mean)))
```

## Contrasts {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
What this code is doing:

```{r}
#| echo: TRUE
with(seaweed, split(logitCover, Treat))
```

```{r}
#| echo: TRUE
with(seaweed, lapply(split(logitCover, Treat), mean))
```

```{r}
#| echo: TRUE
with(seaweed, unlist(lapply(split(logitCover, Treat), mean)))
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Means <- with(seaweed, unlist(lapply(split(logitCover, Treat), mean)))
names(Means) <- names(summary(seaweed$Treat))
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Means <- with(seaweed, unlist(lapply(split(logitCover, Treat), mean)))
names(Means) <- names(summary(seaweed$Treat))
Means
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(lm(logitCover ~ Treat + Block, data = seaweed))
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pulling out estimate of sigma
s <- summary(lm(logitCover ~ Treat + Block, data = seaweed))$sigma
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
ghat <- 0.5 * Means["LfF"] - 0.5 * Means["Lf"] + 0.5 * Means["fF"] - 0.5 * Means["f"]
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculations from previous formulas
ghat <- 0.5 * Means["LfF"] - 0.5 * Means["Lf"] + 0.5 * Means["fF"] - 0.5 * Means["f"]
SE_ghat <- s * sqrt((0.5)^2 / 16 + (-0.5)^2 / 16 + (0.5)^2 / 16 + (-0.5)^2 /16)
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Upper and lower confidence interval bounds
ghat - qt(0.975, 83) * SE_ghat
ghat + qt(0.975, 83) * SE_ghat
```

## Contrasts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Upper and lower confidence interval bounds
lwr <- ghat - qt(0.975, 83) * SE_ghat
upr <- ghat + qt(0.975, 83) * SE_ghat
CI <- c(lwr, upr)
CI
```

## Contrasts (Sophisticated Way) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# One can calculate all 5 contrasts at once
M <- matrix(c(0, 0, -1/2, 1/2, -1/2, 1/2,
              -1/2, -1/2, 1/2, 0, 1/2, 0,
              1/3, -1/3, -1/3, -1/3, 1/3, 1/3,
              1, -1, 1/2, 1/2, -1/2, -1/2,
              0, 0, -1, 1, 1, -1),
            nrow = 6, ncol = 5)
M
```

## Contrasts (Sophisticated Way) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
contrasts(seaweed$Treat) <- M %*% solve(t(M) %*% M)
```

## Contrasts (Sophisticated Way) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(lm(logitCover ~ Treat + Block, data = seaweed))$coefficients
```

## Contrasts (Sophisticated Way) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(lm(logitCover ~ Treat + Block, data = seaweed))$coefficients
confint(lm(logitCover ~ Treat + Block, data = seaweed))
```

## Contrasts (Sophisticated Way) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Indicators <- contr.treatment(6)         # A diagonal matrix
colnames(Indicators) <- names(Means)[-1] # Name variables as before
contrasts(seaweed$Treat) <- Indicators   # Putting it back to normal
```