---
title: "(ST 412/512) Lab 3 <br> Multiple Linear Regression (MLR) Inference"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
header-includes:
  - |
    <style>
      .reveal .smaller {
        font-size: 0.75em;  /* Adjust as needed */
      }

      .reveal .tiny-font {
        font-size: 0.5em;  /* Even smaller font size */
        white-space: nowrap;
        overflow-x: auto;  /* Allow horizontal scrolling */
      }
    </style>
---


## Lab Objectives
* Perform hypothesis tests 
* Confidence intervals on single parameters
* Answer Questions on Practice Quiz

## Dataset {.smaller}
* `case1002` (The Energy Costs of Echolocation by Bats)
  * Looking at bats and how much energy they expend in-flight
  * 20 observations (rows)
  * 3 variables (columns)
    * `Mass`: mass (in grams)
    * `Type`: a factor with 3 levels indicating the type of flying vertebrate: 
      * non-echolocating bats
      * non-echolocating birds
      * echolocating bats
    * `Energy`: in–flight energy expenditure (in W)

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# View Data
case1002
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Observing the structure of our data
str(case1002)
```

## Running a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
* Recall the question of interest from lecture:
  * Zoologists wonder whether the combined energy costs of echolocation and flight in bats is just the sum of the flight energy cost and the at-rest echolocation, or whether bats have developed a more efficient mechanism for the in-flight echolocation
* Could be answered with a test that the coefficient on the echo-locating bat indicator variable was zero

## Running a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
fit_bats <- lm(log(Energy) ~ log(Mass) + Type, data = case1002)
summary(fit_bats)
```

## Running a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# If we want the coefficients information
summary(fit_bats)$coefficients
```

## Running a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(fit_bats)$coefficients[, 3]   # t-statistics
```

## Running a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(fit_bats)$coefficients[, 3]   # t-statistics
summary(fit_bats)$coefficients[, 4]   # p-value
```

## Running a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(fit_bats)$sigma   # Estimate of sigma
```

## Changing Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Our current model is

::: {.fragment}

\begin{align*}
\mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) \\
& \quad + \beta_{2}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bat}\right\} \\
& \quad + \beta_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\}
\end{align*}

:::

* We can see that `echolocating bats` is the reference group (since it comes first alphabetically)
* If we want to change this, we can change the reference group manually as talked about in last lab

## Changing Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating a new column where the variable has a different reference group
case1002$Type2 <- relevel(case1002$Type, ref = "non-echolocating bats")
```

## Changing Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fitting model with newly created variable
fit2 <- lm(log(Energy) ~ log(Mass) + Type2, data = case1002)
summary(fit2)
```

## Changing Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Now our current model is

::: {.fragment}

\begin{align*}
\mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) \\
& \quad + \beta_{2}\mathbf{I}\left\{\text{Type} = \text{Echo Bat}\right\} \\
& \quad + \beta_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\}
\end{align*}

:::

## Exercise {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Note that the parameter estimate for the non-echolocating bats in the previous model fit and the parameter estimate for the echolocating bats in this model fit are the same. Why?

## Exercise {.tiny-font auto-animate="true" auto-animate-easing="ease-in-out"}
* Parameter estimate for the non-echolocating bats in the previous model fit
  * Model 1: Echolocating bats as reference group

::: {.fragment}


\begin{align*}
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) \\
    & \quad + \beta_{2}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bat}\right\} \\
    & \quad + \beta_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\} \\[0.25cm]
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type = Echo Bat}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass})\\[0.25cm]
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type = Non-Echo Bat}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) + \beta_{2} \\[0.25cm]
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type = Non-Echo Bird}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) + \beta_{3} 
\end{align*}

:::

* Parameter estimate for the echolocating bats in this model fit
  * Model 2: Non-Echolocating bats as reference group

::: {.fragment}

\small
\begin{align*}
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \alpha_{0} + \alpha_{1}\ln(\text{Mass}) \\
    & \quad + \alpha_{2}\mathbf{I}\left\{\text{Type} = \text{Echo Bat}\right\} \\
    & \quad + \alpha_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\} \\[0.25cm]
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type = Echo Bat}\right) & = \alpha_{0} + \alpha_{1}\ln(\text{Mass}) + \alpha_{2} \\[0.25cm]
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type = Non-Echo Bat}\right) & = \alpha_{0} + \alpha_{1}\ln(\text{Mass}) \\[0.25cm]
    \mu\left(\ln(\text{Energy}) \mid \text{Mass, Type = Non-Echo Bird}\right) & = \alpha_{0} + \alpha_{1}\ln(\text{Mass}) + \alpha_{3} 
\end{align*}

:::

## Exercise {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* We then would obtain a system of equations
  * Equation 1: Type = Echo Bat
  
  ::: {.fragment}
  
  \begin{align*}
        \beta_{0} + \beta_{1}\ln(\text{Mass}) & = \alpha_{0} + \alpha_{1}\ln(\text{Mass}) + \alpha_{2} 
  \end{align*}
  
  :::
    
  * Equation 2: Type = Non-Echo Bat
  
  ::: {.fragment}
  
  \begin{align*}
        \beta_{0} + \beta_{1}\ln(\text{Mass}) + \beta_{2} & = \alpha_{0} + \alpha_{1}\ln(\text{Mass})
    \end{align*}
    
  :::
  
  
  * Equation 3: Type = Non-Echo Bird
  
  ::: {.fragment}
    
  \begin{align*}
      \beta_{0} + \beta_{1}\ln(\text{Mass}) + \beta_{3}  & = \alpha_{0} + \alpha_{1}\ln(\text{Mass}) + \alpha_{3} 
  \end{align*}
  
  :::

## Exercise {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Just rearranging the equations by isolating the constants on one side and the terms with $\ln(\text{Mass})$ on the other side
* Equation 1: Type = Echo Bat
  
  ::: {.fragment}
  
  \begin{align*}
        \beta_0 - \alpha_0 = (\alpha_1 - \beta_1) \ln(\text{Mass}) + \alpha_2
  \end{align*}
  
  :::
    
  * Equation 2: Type = Non-Echo Bat
  
  ::: {.fragment}
  
  \begin{align*}
        \beta_0 - \alpha_0 = (\alpha_1 - \beta_1) \ln(\text{Mass}) + \beta_2
    \end{align*}
    
  :::
  
  
  * Equation 3: Type = Non-Echo Bird
  
  ::: {.fragment}
    
  \begin{align*}
      \beta_0 - \alpha_0 = (\alpha_1 - \beta_1) \ln(\text{Mass}) + \beta_3 - \alpha_3 
  \end{align*}
  
  :::
  
## Exercise {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Equate the terms for the constants and the $\ln(\text{Mass})$ coefficients From the first and second equations, we have
  * From the first and second equations, we have:
  
  ::: {.fragment}
  
  \begin{align*}
       \alpha_2 = \beta_2
  \end{align*}
  
  :::
    
  * From the second and third equations, we have:
  
  ::: {.fragment}
  
  \begin{align*}
        \beta_2 = \beta_3 - \alpha_3
    \end{align*}
    
  :::
  
  
  * Thus, we can substitute $\alpha_2$ for $\beta_2$ and find the relationship between $\beta_3$ and $\alpha_3$
  
  ::: {.fragment}
    
  \begin{align*}
      \alpha_2 = \beta_3 - \alpha_3
  \end{align*}
  
  \begin{align*}
      \beta_3 = \alpha_2 + \alpha_3
  \end{align*}
  
  :::

## Question {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Test the hypothesis that the mean of log-energy expenditure by non-echolocating bats and birds are different
* Using the following model:

::: {.fragment}

\begin{align*}
\mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) \\
& \quad + \beta_{2}\mathbf{I}\left\{\text{Type} = \text{Echo Bat}\right\} \\
& \quad + \beta_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\}
\end{align*}

:::

* Since `non-echolocating bats` are the reference group our hypotheses are

::: {.fragment}

$$
H_{0}: \beta_{3} = 0, \quad\quad\quad H_{1}: \beta_{3} \neq 0
$$
:::
 
## Question {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fitting model with manually created binary variables???
case1002$ebat <- ifelse(case1002$Type == "echolocating bats", 1, 0)
case1002$bird <- ifelse(case1002$Type == "non-echolocating birds", 1, 0)
```

## Question {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fitting model with manually created binary variables???
fit_3 <- lm(log(Energy) ~ log(Mass) + ebat + bird, data = case1002)
summary(fit_3)
```

## Exercise 2 {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Test the hypothesis that the mean of log-energy expenditure by echolocating bats and nonecholocating birds are the same.
  * Write the model and hypotheses
  
::: {.fragment}

\begin{align*}
\mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) \\
& \quad + \beta_{2}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bat}\right\} \\
& \quad + \beta_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\}
\end{align*}

:::

::: {.fragment}

$$
H_{0}: \beta_{2} = 0, \quad\quad\quad H_{1}: \beta_{2} \neq 0
$$

:::

  * $p$-value and conclusion

## Exercise 2 {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Using the first (default) model
fit_bats <- lm(log(Energy) ~ log(Mass) + Type, data = case1002)
summary(fit_bats)
```

## Exercise 3 {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Test the hypothesis that the slopes of the linear relationship between the mean of log-energy expenditure and log of Mass are different between echolocating bats and non-echolocating bats.
  * Write the model and hypotheses
  
::: {.fragment}

\begin{align*}
\mu\left(\ln(\text{Energy}) \mid \text{Mass, Type}\right) & = \beta_{0} + \beta_{1}\ln(\text{Mass}) \\
& \quad + \beta_{2}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bat}\right\} \\
& \quad + \beta_{3}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\} \\
& \quad + \beta_{4}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bat}\right\}\cdot\ln(\text{Mass}) \\
& \quad + \beta_{5}\mathbf{I}\left\{\text{Type} = \text{Non-Echo Bird}\right\}\cdot\ln(\text{Mass}) \\
\end{align*}

:::

::: {.fragment}

$$
H_{0}: \beta_{4} = 0, \quad\quad\quad H_{1}: \beta_{4} \neq 0
$$

:::

  * $p$-value and conclusion
  
## Exercise 3 {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Using the first (default) model but now with interactions
fit_bats_int <- lm(log(Energy) ~ log(Mass) + Type + Type:log(Mass), data = case1002)
summary(fit_bats_int)
```

## Confidence Intervals {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Using R
confint(fit2)

```

## Confidence Intervals {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Using R
confint(fit2)

# By hand
-1.57636 + c(-1, 1) * qt(0.975, 16) * 0.28724
```