---
title: "(ST 412/512) Lab 8 <br> Two-Way ANOVA"
format:
  revealjs:
    incremental: true
    scrollable: true
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: true
    chalkboard: true
    code-line-numbers: true
    css: (ST 512) StyleSheets.css
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

## Welcome Back
* Any questions before we get started?
* Office Hours: Thursday 10:00am - 11:00am

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
case1302 <- read.csv("case1302.csv",header=TRUE) 
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3) # Contains case1302 dataset
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1302
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(case1302)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
levels(case1302$Company)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1302$Company <- factor(case1302$Company, levels = c("C1", "C2", "C3", "C4", "C5",
                                                        "C6", "C7", "C8", "C9", "C10"))
levels(case1302$Company)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(as.numeric(case1302$Company), case1302$Score,
  xlab = "Company",
  ylab = "Score",
  pch = ifelse(case1302$Treat == "Pygmalion", 17, 16),
  col = ifelse(case1302$Treat == "Pygmalion", "red", "black")
)
legend(
  "topright",
  pch = c(17, 16),
  col = c("red", "black"),
  legend = c("Pygmalion", "Control")
)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
boxplot(case1302$Score ~ case1302$Company,
  main = "Scores vs Company",
  xlab = "Company",
  ylab = "Score",
  col = 1:10
)
# Add the overall mean of scores and standard deviations to facilitate comparisons
abline(h = mean(case1302$Score))
abline(h = mean(case1302$Score) + sd(case1302$Score),
       lty = 2)
abline(h = mean(case1302$Score) - sd(case1302$Score),
       lty = 2)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
boxplot(case1302$Score ~ case1302$Treat,
  main = "Scores vs Treatment",
  xlab = "Company",
  ylab = "Score",
  col = 3:4
)
# Add the overall mean of scores and standard deviations to facilitate comparisons
abline(h = mean(case1302$Score))
abline(h = mean(case1302$Score) + sd(case1302$Score),
       lty = 2)
abline(h = mean(case1302$Score) - sd(case1302$Score),
       lty = 2)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
interaction.plot(
  case1302$Company,
  case1302$Treat,
  case1302$Score,
  col = ifelse(case1302$Treat == "Pygmalion", "red", "blue"),
  xlab = "Company",
  ylab = "Scores",
  trace.label = "Treatment"
)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Fit the non-additive model
non_add <- lm(Score ~ Treat + Company + Treat:Company, data = case1302)
summary(non_add)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Fit model without intercept term
summary(lm(Score ~ Treat + Company + Treat:Company - 1, data = case1302))
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(non_add)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(non_add$fitted, non_add$resid,
  xlab = "Fitted Values",
  ylab = "Residuals",
  pch = 16
)
abline(h = 0)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
add <- lm(Score ~ Treat + Company, data = case1302) 
summary(add)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(add)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(add$fitted, add$resid,
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch = 16)
abline(h = 0)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(add, non_add)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
one <- lm(Score ~ Treat, data = case1302)
summary(one)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(one)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(one$fitted, one$resid,
     xlab = "Fitted Values",
     ylab = "Residuals",
     pch = 16)
abline(h = 0)
```

## Fittin' {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(one, add)
```