---
title: "(ST 412/512) Lab 4 <br> Polynomial Regression"
format:
  revealjs:
    incremental: true
    scrollable: true
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: true
    chalkboard: true
    code-line-numbers: true
    css: styles.css
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

## Welcome Back
* Any questions before we get started?
* Office Hours: Thursday 10:00am - 11:00am
* Pretty short lab today (hurrah?)

## [R Studio Customization]{.r-fit-text}

For those interested in customizing RStudio, the following are some of my favorite enhancements, ranked by the level of enthusiasm they usually spark.

## [R Studio Customization - Dark Mode]{.r-fit-text}
* To change the editor's theme in RStudio:
* **Tools → Global Options → Appearance → Editor theme**
* If you want more options, you can install the **`rsthemes`** package, which provides additional RStudio IDE themes for syntax highlighting and the user interface.

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: FALSE
install.packages("rsthemes")
```
:::

## [R Studio Customization - Rainbow Parentheses]{.r-fit-text}
* For multicolored matching parentheses:
* **Tools → Global Options → Code → Display → Syntax**

## [R Studio Customization - Window Pane Adjustments]{.r-fit-text}
* I (and many others) find that moving the console to be side-by-side with the source files is convenient. To adjust your windows:
* **Tools → Global Options → Pane Layout**
* or you can access the panes by clicking on their icon (below **Profile**). If you want your console side-by-side you would swap its position with the environment.
* If you want to have multiple scripts set up to edit you can click **Add Column**

## Polynomial Regression
Polynomial regression is just a specific version of multiple linear regression (MLR) where some of our explanatory variables have higher-order powers, for example:

::: {.fragment}
$$
\operatorname{E}\left(Y \mid \mathbf{X}\right) = \beta_{0} + \beta_{1}X + \beta_{2}X^{2} + \beta_{3}X^{3}
$$
:::

<div class="footnote fragment" style="position: absolute; bottom: 1em; font-size: 0.8em;">
   There could be more or less polynomial terms included, this was just an example
</div>

## [Dataset Summary: Galileo's Data on Falling Bodies]{.r-fit-text}
* This dataset comes from one of Galileo’s experiments (circa 1609) investigating the motion of falling bodies with a horizontal component of motion.
* The data were used to support his mathematical proof that such trajectories follow a parabolic path.

## Dataset Details {.smaller}

- **Dataset name:** `case1001` (Galileo's Data on the Motion of Falling Bodies)  
- **Number of observations:** 7  
- **Number of variables:** 2  

::: {.fragment}

| Variable   | Description |
|-----------|-------------|
| `Distance` | Horizontal distance traveled (in *punti*) |
| `Height`   | Initial height of the falling body (in *punti*) |

:::

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
case1001 <- read.csv("case1001.csv", header = TRUE)
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Loading packages (optional)
library(Sleuth3) # Actually contains case1001 dataset and much more
library(ggplot2)
```

## Data's Structure {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Observing the structure of our dataframe
str(case1001)
```

## Plotting Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Base plot code
plot(case1001$Height, case1001$Distance, 
     pch = 20, 
     col = "blue")
```

## Plotting Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Base plot code
plot(case1001$Height, case1001$Distance, 
     pch = 20, 
     col = "blue")

# Plot regression line
abline(lm(Distance ~ Height, data = case1001), col = "blue")
```

## Plotting Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Correlation between X and Y
cor(case1001$Height, case1001$Distance)
```

## Plotting Data {auto-animate="true" auto-animate-easing="ease-in-out"}
* Some people might have looked at the given plot and thought there was some curvature apparent
  * Wanting to "capture" such curvature, an option would be to fit a polynomial regression (add in higher order terms to the model)
  
## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Polynomial of degree 1 (or order 1), i.e., a straight line
fit1 <- lm(Distance ~ Height, data = case1001)
summary(fit1)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Corresponding to the following model:
\begin{align*}
\text{Model Equation: } \quad  \mu(\text{Distance} \mid \text{Height}) & = \beta_0 + \beta_1 \text{Height} \\[0.25cm]
\text{Fitted Equation: }\quad  \hat{\mu}(\text{Distance} \mid \text{Height}) & = \hat{\beta}_{0} + \hat{\beta}_{1}\,\text{Height} \\[0.20cm]
    & = `r summary(fit1)[["coefficients"]][1, 1]` + `r summary(fit1)[["coefficients"]][2, 1]`\,\text{Height}
\end{align*}

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting the residuals
plot(case1001$Height, fit1$res,
     pch = 20,
     col = "blue")
abline(h = 0) 
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* We see some pattern in the residual plot (i.e. it does not look to be randomly scattered)
* If we are to obtain our $R^{2}$ and $R^{2}_{\text{Adj.}}$ values

::: {.fragment}
$$
R^{2} = `r summary(fit1)$r.squared`, \quad\quad\quad R^{2}_{\text{Adj.}} = `r summary(fit1)$adj.r.squared`
$$
:::

* This is indicating that a large part of the variance is $Y$ is being explained by our explanatory variable(s)
* One could use these as a metric when fitting higher order terms (to assess fit)

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1001$Height2 <- case1001$Height^2 # Need to define the variable to include in the model

# Polynomial of degree 2 (or order 2)
fit2 <- lm(Distance ~ Height + Height2, data = case1001)
summary(fit2)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting the residuals
plot(case1001$Height, fit2$res, 
     pch = 20, 
     col = "blue")
abline(h = 0)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Many ways of coding the higher order terms
fit2.I <- lm(Distance ~ Height + I(Height^2), data = case1001)
fit2.poly <- lm(Distance ~ poly(Height, 2, raw = TRUE), data = case1001)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* `I()` stands for “as-is” or “inhibit interpretation”.
  * In a formula, R normally treats operators like `^`, `*`, and `+` in a special modeling sense. `I()` tells R: “don’t treat this as a model formula — just calculate it literally.”
  * Equivalent to when we previously created a separate squared variable
* `poly(Height, 2, raw = TRUE)` also creates polynomial terms — but does it automatically
  * `poly(Height, 2)` would be  `raw = FALSE`, meaning `poly()` would create orthogonal (uncorrelated) polynomial terms instead of raw powers.
  
## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# We can see they're all the same results
summary(fit2)$coef
summary(fit2.I)$coef
summary(fit2.poly)$coef
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Polynomial of degree 3 (or order 3)
fit3 <- lm(Distance ~ poly(Height, 3, raw = TRUE), data = case1001)
summary(fit3)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting the residuals
plot(case1001$Height, fit3$res, 
     pch = 20, 
     col = "blue")
abline(h = 0)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Polynomial of degree 4 (or order 4)
fit4 <- lm(Distance ~ poly(Height, 4, raw = TRUE), data = case1001)
summary(fit4)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting the residuals
plot(case1001$Height, fit4$res, 
     pch = 20,
     col = "blue")
abline(h = 0)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Polynomial of degree 5 (or order 5)
fit5 <- lm(Distance ~ poly(Height, 5, raw = TRUE), data = case1001)
summary(fit5)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting the residuals
plot(case1001$Height, fit5$res, 
     pch = 20,
     col = "blue")
abline(h = 0)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Polynomial of degree 6 (or order 6)
fit6 <- lm(Distance ~ poly(Height, 6, raw = TRUE), data = case1001)
summary(fit6)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting the residuals
plot(case1001$Height, fit6$res, 
     pch = 20,
     col = "blue")
abline(h = 0)
```

## Fitting Polynomial Regressions {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* One could look at all of their $R^{2}$ and $R^{2}_{\text{Adj.}}$ values and compare
* One could look at the residuals plots
* One could compare via ESS $F$-tests (two models at a time)

## Model Comparisons {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Comparing the linear and quadratic models
anova(fit1, fit2) # Order matters (reduced, full)
```

## Model Comparisons {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
\begin{align*}
\text{Model Equation (\text{fit1}): } \quad  \mu(\text{Distance} \mid \text{Height}) & = \beta_0 + \beta_1 \text{Height} \\[0.25cm]
\text{Model Equation (\text{fit2}): }\quad  \mu(\text{Distance} \mid \text{Height}) & = \alpha_0 + \alpha_1 \text{Height} + \alpha_2 \text{Height}^{2}
\end{align*}

$$
H_{0}: \alpha_{2} = 0, \qquad H_{1}: \alpha_{2} \neq 0
$$


## Model Comparisons {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Comparing the quadratic and cubic models
anova(fit2, fit3)
```

## Model Comparisons {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(fit3, fit4)
```

## Model Comparisons {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(fit4, fit5)
```

## Model Comparisons {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(fit5, fit6)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* To plot the different curves, we basically need to create a "grid"" of of points in the x-axis and evaluate our polynomials for each one of the points in the grid

::: {.fragment}
```{r}
#| echo: TRUE
#| output-location: slide
grid <- data.frame(Height = seq(from = 0, to = 1100, by = 1)) # Range based on observed data
grid$pred1 <- predict(fit1, grid) # Evaluate the grid using fitted model
plot(grid$Height, grid$pred1, col = "red") # Plotting the points we just created
```
:::

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
grid$pred2 <- predict(fit2.poly, grid) # Using the version where we used the function poly()
grid$pred3 <- predict(fit3, grid)
grid$pred4 <- predict(fit4, grid)
grid$pred5 <- predict(fit5, grid)
grid$pred6 <- predict(fit6, grid)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
par(mfrow = c(2, 3))
# Linear
plot(case1001$Height, case1001$Distance,
     pch = 20,
     col = "blue",
     main = "linear")
lines(grid$Height, grid$pred1, col = "red")

# Quadratic
plot(case1001$Height, case1001$Distance,
     pch = 20,
     col = "blue",
     main = "quadratic")
lines(grid$Height, grid$pred2, col = "red")

# Cubic
plot(case1001$Height, case1001$Distance,
     pch = 20,
     col = "blue",
     main = "cubic")
lines(grid$Height, grid$pred3, col = "red")

# Quartic
plot(case1001$Height, case1001$Distance,
     pch = 20,
     col = "blue",
     main = "quartic")
lines(grid$Height, grid$pred4, col = "red")

# Quintic
plot(case1001$Height, case1001$Distance,
     pch = 20,
     col = "blue",
     main = "quintic")
lines(grid$Height, grid$pred5, col = "red")

# Sextic
plot(case1001$Height, case1001$Distance,
     pch = 20,
     col = "blue",
     main = "sextic")
lines(grid$Height, grid$pred6, col = "red")
```

## But Really... {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* Though we teach this exploratory process here, in reality it shouldn't be done preceding inference
* There are several problems that arise from the previous work shown today:
  * Multiple hypothesis testing
  * Data snooping
  
  