---
title: "(ST 412/512) <br> Case Influence Statistics"
format:
  revealjs:
    incremental: true
    scrollable: true
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: true
    chalkboard: true
    code-line-numbers: true
    css: (ST 512) StyleSheets.css
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

## Welcome Back
* Any questions before we get started?
* Office Hours: Thursday 10:00am - 11:00am

## What Are Case Influence Statistics? {.smaller}
- Case influence statistics help us identify **observations that disproportionately affect a regression model’s results**.
- In regression analysis, some data points can:
  - Strongly affect coefficient estimates  
  - Change statistical significance
  - Distort predictions
  - Alter other model fit aspects
- These points may be:
  - **Outliers** (large residual values)  
  - **High leverage points** (unusual predictor values)  
  - **Influential observations** (large impact on model estimates)

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
case1101 <- read.csv("case1101.csv", header = TRUE)
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Loading packages (optional)
library(Sleuth3) # Actually contains case1101 dataset and much more
library(ggplot2)
```

## Dataset Details {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
These data were collected on **18 women** and **14 men** to investigate a theory explaining why:

- Women exhibit **lower tolerance for alcohol**
- Women develop **alcohol-related liver disease more readily** than men
- Variables
  - **Subject** : Subject number in the study
  - **Metabol** : First-pass metabolism of alcohol *(mmol/liter-hour)* 
  - **Gastric** : Gastric alcohol dehydrogenase (AD) activity *(μmol/min/g of tissue)*
  - **Sex** : Sex of the subject
  - **Alcohol** : Alcoholic or non-alcoholic status
  
  
## Checkin' Out Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# This specifies the colors for the symbols
myBg <- ifelse(case1101$Sex == "Female", "black", "white")

# This specifies the symbols (circle or triangle)
myPch <- ifelse(case1101$Alcohol == "Alcoholic", 21, 24)
```

## Checkin' Out Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# This specifies the colors for the symbols
myBg <- ifelse(case1101$Sex == "Female", "black", "white")

# This specifies the symbols (circle or triangle)
myPch <- ifelse(case1101$Alcohol == "Alcoholic", 21, 24)

# Plot Action
plot(case1101$Gastric, case1101$Metabol,
  pch = myPch,
  bg = myBg,
  xlab = "Gastric AD Activity",
  ylab = "First-Pass Metabolism"
)
```

## Checkin' Out Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: slide
# This specifies the colors for the symbols
myBg <- ifelse(case1101$Sex == "Female", "black", "white")

# This specifies the symbols (circle or triangle)
myPch <- ifelse(case1101$Alcohol == "Alcoholic", 21, 24)

# Plot Action
plot(case1101$Gastric, case1101$Metabol,
  pch = myPch,
  bg = myBg,
  xlab = "Gastric AD Activity",
  ylab = "First-Pass Metabolism"
)
legend(1, 12,
  pch = c(24, 24, 21, 21),
  pt.bg = c("white", "black", "white", "black"),
  c("Non-alcoholic Males",
    "Non-alcoholic Females",
    "Alcoholic Males",
    "Alcoholic Females")
)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Suppose we want to know which observations are the one's far right
which(case1101$Gastric > 4) # Which observations have values of Gastric greater than 4 
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
which(case1101$Gastric > 5) # Which observations have values of Gastric greater than 5 
```

## Checkin' Out Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: slide
# This specifies the colors for the symbols
myBg <- ifelse(case1101$Sex == "Female", "black", "white")

# This specifies the symbols (circle or triangle)
myPch <- ifelse(case1101$Alcohol == "Alcoholic", 21, 24)

# Plot Action
plot(case1101$Gastric, case1101$Metabol,
  pch = myPch,
  bg = myBg,
  xlab = "Gastric AD Activity",
  ylab = "First-Pass Metabolism"
)
legend(1, 12,
  pch = c(24, 24, 21, 21),
  pt.bg = c("white", "black", "white", "black"),
  c("Non-alcoholic Males",
    "Non-alcoholic Females",
    "Alcoholic Males",
    "Alcoholic Females")
)
labelsubs <- subset(case1101, Gastric > 4) # Identify observations with Gastric values > 4
text(
  labelsubs$Gastric,
  labelsubs$Metabol,
  row.names(labelsubs),
  pos = 1,
  col = "red"
) # Add the observations numbers to the plot
```

## Fitting Model {.text-xxs auto-animate="true" auto-animate-easing="ease-in-out"}
- Variables
- **meta** : First-pass metabolism (μmol/liter-hour)
- **gast** : Gastric alcohol dehydrogenase activity  (μmol/min/g tissue)
- **fem** : Indicator variable (1 = Female, 0 = Male)
- **alc** : Indicator variable (1 = Alcoholic, 0 = Non-alcoholic)

::: {.fragment}
$$
\begin{align*}
\mu\left(\text{meta} \mid gast,fem,alc\right) & =
\beta_0 + \beta_1 gast + \beta_2 fem + \beta_3 alc
+ \beta_4 (gast \times fem) \\[0.20cm]
& + \beta_5 (fem \times alc)
+ \beta_6 (gast \times alc)
+ \beta_7 (gast \times fem \times alc)
\end{align*}
$$
:::

## Fitting Model {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Defining some variables
meta <- case1101$Metabol
gast <- case1101$Gastric
alc <- as.numeric(case1101$Alcohol == "Alcoholic") # Indicator variables
fem <- as.numeric(case1101$Sex == "Female")        # Indicator variables
```

## Fitting Model {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Defining some variables
meta <- case1101$Metabol
gast <- case1101$Gastric
alc <- as.numeric(case1101$Alcohol == "Alcoholic") # Indicator variables
fem <- as.numeric(case1101$Sex == "Female")        # Indicator variables

# A data frame containing the original data
data_orig <- data.frame(meta, gast, alc, fem)
```

## Fitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Defining some variables
meta <- case1101$Metabol
gast <- case1101$Gastric
alc <- as.numeric(case1101$Alcohol == "Alcoholic") # Indicator variables
fem <- as.numeric(case1101$Sex == "Female")        # Indicator variables

# A data frame containing the original data
data_orig <- data.frame(meta, gast, alc, fem)

# Note that "*" will produce the model above with all possible interactions
mod1 <- lm(meta ~ gast * alc * fem, data = data_orig) 
summary(mod1)
```

## Fitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: slide
# We observe that, other than gast, the only term that is significant is the interaction gast × fem
# Looking at the residuals from this model we obtain
plot(mod1$fitted, mod1$resid,
  xlab = "Fitted Values",
  ylab = "Residual FPM",
  pch = 16,
  ylim = c(-4, 4)
)
abline(h = 0)
```

## Fitting Model {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We observe a couple of points away from the rest again
which(mod1$fitted > 8) # Which observations fitted values greater than 8
```

## Fitting Model {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
which(as.vector(mod1$fitted) > 8) # Observe the difference with this statement
```

## Fitting Model {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
mod1$fitted # We can also inspect the entire list of fitted values
```

## Refitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Remove observations 31 and 32 from the original data
data_rem <- data_orig[-c(31, 32), ] 

# Re-fit the model removing the observations
mod1_rem <- lm(meta ~ gast * alc * fem, data = data_rem) 

summary(mod1_rem) # Print summary for comparisons 
```

## Case-Influence Statistics {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
What are case-influence statistics?
- Numerical measures used to detect **influential observations**

## [Case-Influence Statistics - Leverage]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Measures distance of predictor values from their overall mean  
- High leverage → observation is far in predictor space  
- May strongly influence the fitted model

## [Case-Influence Statistics - Standardized Residuals]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Residual divided by its estimated standard deviation  
- Places residuals on a common scale  
- Helps detect unusual observations  

::: {.aside}
Note: Some sources use *studentized residuals* interchangeably. R distinguishes between standardized and studentized residuals.
:::

## [Case-Influence Statistics - Cook’s Distance]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Measures overall influence of removing an observation  
- Assesses impact on estimated regression coefficients  
- Large values → potentially influential case

## [Case-Influence Statistics - Leverage]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can extract the leverage of each observation in the model using the function hatvalues(.)
hatvalues(mod1) # Looking at the leverage for each observation in the original model
```

## [Case-Influence Statistics - Leverage]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: slide
# Plot leverage against the observations
plot(1:nrow(case1101),
     hatvalues(mod1),
     ylab = "Leverage",
     xlab = "Observations")
# Rule of thumb 2 * p / n = 2 * 8 / 32 = 0.50
abline(h = 0.5, lty = "dashed", col = "red") # We include this value for reference
```

## [Case-Influence Statistics - Leverage]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
which(hatvalues(mod1) > 0.5) # Identify observations greater than 0.5
```

## [Case-Influence Statistics - Standardized Residuals]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: slide
plot(1:nrow(case1101), rstandard(mod1), 
     ylab = "Standardized residuals", 
     xlab = "Observations") # Plot leverage against the observations
abline(h = 0, lty = "dashed") # We include line at 0 for reference
abline(h = 2, lty = "dashed", col = "red") # We include this value for reference
abline(h = -2, lty = "dashed", col = "red") # We include this value for reference
```

## [Case-Influence Statistics - Standardized Residuals]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
which(abs((rstandard(mod1))) > 2) # Identify observations greater than 2
```


## [Case-Influence Statistics - Cook’s Distance]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
cooks.distance(mod1) # Looking at the Cook distances for each observation in the original model
```

## [Case-Influence Statistics - Cook’s Distance]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
plot(1:nrow(case1101), cooks.distance(mod1), 
     ylab = "Cook distance", 
     xlab = "Observations") # Plot leverage against the observations
abline(h = 1, lty = "dashed", col = "red") # We include this value for reference
```

## [Case-Influence Statistics - Cook’s Distance]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
which(cooks.distance(mod1) > 1) # Identify observations greater than 1
```

## [Case-Influence Statistics - Important Principles]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Reference lines are **guidelines**, not strict rules.
- Do **not** rely on a single diagnostic measure.
- Combine:
  - Leverage
  - Standardized residuals
  - Cook’s distance
- Influential observations are **not automatically errors**.

## [Case-Influence Statistics - This Example]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- **Obs 31 & 32**
  - Large standardized residuals
  - Large Cook’s distance
  - High leverage (31 especially)
  → Strong influence on model results

- **Obs 1, 2, 23**
  - High leverage
  - Small residuals
  - Smaller overall influence

- **After Removing 31 & 32**
  - Interaction terms lose significance
  - $R^{2}$ decreases substantially
  - Model conclusions change

## [Case-Influence Statistics - Cook’s Distance]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Recall the model's output
summary(mod1)
```

## [Case-Influence Statistics - Cook’s Distance]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Removing the "unusual" observations, how does our output change?
data_rem <- data_orig[-c(31, 32), ]
mod1_new <- lm(meta ~ gast * alc * fem, data = data_rem)
summary(mod1_new)
```

