---
title: "(ST 412/512) <br> Model selection"
format:
  revealjs:
    incremental: true
    scrollable: true
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: true
    chalkboard: true
    code-line-numbers: true
    css: (ST 512) StyleSheets.css
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

## Welcome Back
* Any questions before we get started?
* Office Hours: Thursday 10:00am - 11:00am
* Today we speak of my conceptual nemesis

## Lab Objectives
* Perform best subsets variable selection using $AIC$, $BIC$, $C_{p}$, and adjusted $R^{2}$
  * Using the `leap` package for sequential model selection
* Make a $C_{p}$ plot
* Use stepwise regression to select variables
* Calculate variance inflation factors

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
case1201 <- read.csv("case1201.csv", header = TRUE)
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
library(leaps) # regsubsets()
library(car) # vif()
```

## Dataset {.xxs auto-animate="true" auto-animate-easing="ease-in-out"}
* `case1201` (State Average SAT Scores)
  * Data on the average SAT scores for US states in 1982 and possible associated factors.
  * 50 observations (rows)
  * 8 variables (columns)
    * `State`: US state
    * `SAT`: state averages of the total SAT (verbal + quantitative) scores
    * `Takers`: the percentage of the total eligible students (high school seniors) in the state who took the exam
    * `Income`: the median income of families of test–takers (in hundreds of dollars)
    * `Years`: the average number of years that the test–takers had formal studies in social sciences, natural sciences and humanities
    * `Public`: the percentage of the test–takers who attended public secondary schools
    * `Expend`: he total state expenditure on secondary schools (in hundreds of dollars per student)
    * `Rank`: the median percentile ranking of the test–takers within their secondary school classes

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Create a scatterplot matrix without State names
pairs(case1201[, -1], col = "blue")
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Taking a closer look at the scatterplot for SAT vs Takers
# Check possible curvature in data
plot(case1201$Takers, case1201$SAT, 
     pch = 20, 
     col = "blue")
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# The lab notes want to perform a log transformation due to the 
# nonlinear relationship
case1201$lTakers <- log(case1201$Takers)
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Seeing now
plot(case1201$lTakers, case1201$SAT, 
     pch = 20, 
     col = "blue")
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Create a scatterplot matrix without State names
pairs(case1201[, -1], col = "blue")
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# One might claim there are potential influential
# points when looking at Public and SAT
plot(case1201$Public, case1201$SAT, 
     pch = 20, 
     col = "blue")
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[case1201$Public < 50] # Identify state with low Public 
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# One might claim there are potential influential
# points when looking at Expend and SAT
plot(case1201$Expend, case1201$SAT, 
     pch = 20, 
     col = "blue")
```

## Visualizing Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[case1201$Expend > 40] # Identify state with high Expend 
```

## Model Formulation {.text-xxs auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\mu\left(\text{SAT} \mid \text{predictors}\right)
= \beta_0 
+ \beta_1 \log(\text{Takers}) 
+ \beta_2 \text{Income} 
+ \beta_3 \text{Years} 
+ \beta_4 \text{Public} 
+ \beta_5 \text{Expend} 
+ \beta_6 \text{Rank}
$$

## Model Formulation {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Start with tentative model (all variables)
mod1 <- lm(SAT ~ lTakers + Income + Years + Public + Expend + Rank, data = case1201)
summary(mod1)
```

## Model Formulation {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Check residuals
plot(mod1$fit, mod1$res,
  type = "p",
  pch = 20,
  col = "blue",
  main = "Residual vs fitted values",
  xlab = "Fitted values",
  ylab = "Residuals"
)
abline(h = 0, lty = "dashed") 
```

## Case Influence Stats {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Standardized residuals
Obs <- seq(1:50)
plot(Obs, rstandard(mod1),
  ylab = "Standardized residuals",
  xlab = "Observations",
  ylim = c(-4, 4)
)
abline(h = 0, lty = "dashed") # We include line at 0 for reference
abline(h = 2, lty = "dashed", col = "red") # We include this value for reference
abline(h = -2, lty = "dashed", col = "red") # We include this value for reference
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(abs((rstandard(mod1))) > 2)] # Identify states that  greater than 2
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(abs((rstandard(mod1))) > 3)] # Identify states that  greater than 3
```

## Case Influence Stats {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Leverage 2p'/n = 2(7)/50 = 0.28
plot(Obs, hatvalues(mod1), 
     ylab = "Leverage", 
     xlab = "Observations") # Plot leverage against the observations
abline(h = 0.28, 
       lty = "dashed", 
       col = "red") # We include this value for reference
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(hatvalues(mod1)>0.28)] # Identify observations greater than 0.28
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(hatvalues(mod1)>0.50)] # Identify observations greater than 0.50
```

## Case Influence Stats {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Cook’s Distance
plot(Obs, cooks.distance(mod1), 
     ylab = "Cook distance", 
     xlab = "Observations")
abline(h = 1, 
       lty = "dashed", 
       col = "red") # We include this value for reference
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(cooks.distance(mod1) > 1)] # Identify states with distance greater than 1
```

## Refitting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Re-fit the model, removing "Alaska"
mod2 <- update(mod1, . ~ . , data = case1201, subset = (State != "Alaska"))
summary(mod2)
```

## Refitting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Re-fit the model, removing "Alaska"
plot(mod2$fit, mod2$res,
  type = "p",
  pch = 20,
  col = "blue",
  main = "Residual vs fitted values",
  xlab = "Fitted values",
  ylab = "Residuals"
)
abline(h = 0, lty = "dashed") 
```

## Case Influence Stats {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Standardized residuals
Obs <- seq(1:49)
plot(Obs, rstandard(mod2),
  ylab = "Standardized residuals",
  xlab = "Observations",
  ylim = c(-4, 4)
)
abline(h = 0, lty = "dashed") # We include line at 0 for reference
abline(h = 2, lty = "dashed", col = "red") # We include this value for reference
abline(h = -2, lty = "dashed", col = "red") # We include this value for reference
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(abs((rstandard(mod2))) > 2)] # Identify states with distance greater than 2
```

## Case Influence Stats {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Leverage 2p'/n = 2(7)/50 = 0.29
plot(Obs, hatvalues(mod2), 
     ylab = "Leverage",
     xlab = "Observations") # Plot leverage against the observations
abline(h = 0.29, 
       lty = "dashed",
       col = "red") # We include this value for reference
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(hatvalues(mod2) > 0.29)] # Identify observations greater than 0.28
```

## Case Influence Stats {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Cook’s Distance
plot(Obs, cooks.distance(mod2), 
     ylab = "Cook distance", 
     xlab = "Observations")
abline(h = 1, 
       lty = "dashed", 
       col = "red") # We include this value for reference
```

## Case Influence Stats {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1201$State[which(cooks.distance(mod2) > 1)] # Identify states with distance greater than 1
```

## Model Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adjusting for variables 
mod3 <- lm(SAT ~ lTakers + Rank, data = subset(case1201, State != "Alaska"))
```

## Model Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adjusting for variables 
mod3 <- lm(SAT ~ lTakers + Rank, data = subset(case1201, State != "Alaska"))
best_SAT <- order(mod3$res, decreasing = TRUE)
```

## Model Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adjusting for variables 
mod3 <- lm(SAT ~ lTakers + Rank, data = subset(case1201, State != "Alaska"))
best_SAT <- order(mod3$res, decreasing = TRUE)
case1201$State[best_SAT]
```

## Best-Subsets Variable Selection {.text-xs auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
SAT_best <- regsubsets(SAT ~ Expend + lTakers + Income + Years + Public + Rank, data = case1201, nbest = 7)
```

* What this code is doing
  * The first argument specifies the largest model you are willing to consider. Here, I specify my response is `SAT`, and the possible terms all other terms in the dataset, which is what the `~ .` notation implies.
  * The dataset is excluding Alaska which was done earlier.
  * `nbest` specifies how many models should be returned for each size; I want the best seven models.  
  * `method` states the methods for finding the best models:  
    * `"exhaustive"` considers all models (not a sequential procedure!). 
      * `"exhaustive"` is default.
    * Other options are `"forward"` and `"backward"` (these are sequential procedures).  
  * Another useful argument is `nvmax`, the largest size model you want to consider.  
    * By default, it is set at 8, so here we don’t need to change it.
    
## Best-Subsets Variable Selection {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Contents can be printed out as is
all <- regsubsets(SAT ~ Expend + lTakers + Income + Years + Public + Rank, 
                  data = subset(case1201, State != "Alaska"), 
                  nbest = 7, 
                  method = "exhaustive")
```

## Best-Subsets Variable Selection {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Or using summary we can obtain more information
summary(all)
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
* What the summary output tells us is the ranking for each sized model
  * I believe by default adjusted $R^{2}$ is used to rank the models within group sizes 

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# This object can even be plotted
plot(all)
```

## Best-Subsets Variable Selection {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(ggplot2) 
library(plyr)

# Code bellow will allow us to pull out some other useful information from these output
fortify.regsubsets <- function(model, data, ...) {
  require(plyr)
  stopifnot(model$intercept)
  models <- summary(model)$which
  rownames(models) <- NULL
  model_stats <- as.data.frame(summary(model)[c("bic", "cp", "rss", "rsq", "adjr2")])
  dfs <- lapply(coef(model, 1:nrow(models)), function(x)
    as.data.frame(t(x)))
  model_coefs <- plyr::rbind.fill(dfs)
  model_coefs[is.na(model_coefs)] <- 0
  model_stats <- cbind(model_stats, model_coefs)
  # terms_short <- abbreviate(colnames(models))
  terms_short <- colnames(models)
  model_stats$model_words <- aaply(models, 1, function(row)
    paste(terms_short[row], collapse = "+"))
  model_stats$size <- rowSums(summary(model)$which)
  model_stats
}

get_model_coefs <- function(model) {
  models <- summary(model)$which
  dfs <- lapply(coef(model, 1:nrow(models)), function(x)
    as.data.frame(t(x)))
  model_coefs <- plyr::rbind.fill(dfs)
  model_coefs[is.na(model_coefs)] <- 0
  model_coefs
}
```

## Best-Subsets Variable Selection {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Extract the information from the regsubset(.) above using fortify(.)
models <- fortify(all)
head(models)
```

## Some Plots {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(models$size, models$cp)
abline(0, 1) # Notice the scale
```

## Some Plots {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(models$size, models$cp, ylim = c(1, 10))
abline(0, 1) # Change the scale
```

## Some Plots {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# BIC plot
qplot(size, bic, data = models)
```

## Some Plots {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Identify the model with the lowest Cp
models[models$cp == min(models$cp), ]
```

## Some Plots {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fit model
mod_lowestCp <- lm(SAT ~ Expend + lTakers + Years + Rank, data = subset(case1201, State != "Alaska"))
summary(mod_lowestCp)
```