---
title: "(ST 412/512) Lab 2 <br> Multiple Linear Regression (MLR) in R!"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## Lab Objectives
* Understand how factors work in `R`, and how to specify indicator variables
* Learn how to specify multiple regression models in `R`
* Practice matching shorthand, longhand and code for models

## Dataset {.smaller}
* `case0901` (Effects of Light on Meadowfoam Flowering)
  * Shining lights on some plants to see if they would grow?
  * 24 observations (rows)
  * 3 variables (columns)
    * `Flowers`: average number of flowers per meadowfoam plant
    * `Time`: time light intensity regiments started; 1 = Late, 2 = Early
    * `Intensity`: light intensity (in some wacky unit, idk)

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Loading packages
library(Sleuth3)
library(ggplot2)
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# View Data
case0901
```

## Factors?! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Observing the structure of our dataframe
str(case0901)
```

## Brief Overview of MLR {.smaller}
* If we are to treat `Intensity` as a continuous random variable, our model equation looks like

::: {.fragment}

$$
  \mu\left(Y \mid \mathbf{X}\right) = \beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2}
$$

:::

::: {.fragment}

$$
  \mu\left(\text{Flower} \mid \text{Intensity, Time}\right) = \beta_{0} + \beta_{1}\text{Intensity} + \beta_{2}\text{EarlyStart}
$$

:::

::: {.fragment}

$$
  \mu\left(\text{Flower} \mid \text{Intensity, Time}\right) = \beta_{0} + \beta_{1}\text{Intensity} + \beta_{2}\mathbf{I}\left\{\text{EarlyStart} == 1\right\}
$$

:::

## Brief Overview of MLR {auto-animate="true" auto-animate-easing="ease-in-out"}

:::: {.columns}

::: {.column width="40%"}

```{r}
#| echo: FALSE
Data <- case0901
Data
```

:::

::: {.column width="60%"}

```{r}
#| echo: FALSE
HideData <- Data
HideData$Time <- factor(Data$Time)
model.matrix(Flowers ~ Time + Intensity, data = HideData)
```

:::

::::


## Brief Overview of MLR {auto-animate="true" auto-animate-easing="ease-in-out"} {.smaller}
::: {.fragment}

\begin{align*}
  \mu\left(\text{Flower} \mid \text{Intensity, Time}\right) & = \beta_{0} + \beta_{1}\text{EarlyStart} + \beta_{2}\text{Intensity300}  \\
  & \quad + \beta_{3}\text{Intensity450} + \beta_{4}\text{Intensity600} \\
  & \quad + \beta_{5}\text{Intensity750} + \beta_{6}\text{Intensity900}
\end{align*}

:::

:::{.column-page}

```{r}
#| echo: FALSE
options(width = 100)
HideData$Intensity <- factor(Data$Intensity)
model.matrix(Flowers ~ Time + Intensity, data = HideData)
```

:::

## Brief Overview of MLR {.smaller}
* Since we are working with an intercept, there will be a reference group assigned to one of the levels in a factor variable
  * If the levels are character variables (strings) then the reference group is chosen by whichever comes first alphabetically
  * If the levels are numbers then the lowest number is the reference group (as in our example)
  * Level order can also be specified in the `factor()` or `relevel()` functions
* If we fail to change numeric levels to factors then R will treat them as continuous variables and different results will ensue

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
str(case0901)
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# We can manually change the values to binary
case0901$Day24 <- as.numeric(case0901$Time == 2)
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Breaking down what this is doing
case0901$Time == 2 # Which observations have an EarlyStart?
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Breaking down what this is doing
as.numeric(case0901$Time == 2) # Turn T/F into 1/0
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$Day24 <- as.numeric(case0901$Time == 2)
case0901$Day24
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$Day24 <- as.numeric(case0901$Time == 2)
case0901$Day24
```

## Linear Model Action {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(lm(Flowers ~ Intensity + Day24, data = case0901))
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$Day24 <- ifelse(case0901$Time == 2, 1, 0)
case0901$Day24
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$intensity_f <- factor(case0901$Intensity)
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$intensity_f <- factor(case0901$Intensity)
str(case0901$intensity_f)
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$intensity_f2 <- factor(case0901$Intensity, 
                                levels = c("750", "300", "450", "600", "150", "900"))
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$intensity_f2 <- factor(case0901$Intensity, 
                                levels = c("750", "300", "450", "600", "150", "900"))
levels(case0901$intensity_f2)
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$intensity_f3 <- relevel(case0901$intensity_f, ref = "150")
```

## Making a Factor {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$intensity_f3 <- relevel(case0901$intensity_f, ref = "150")
levels(case0901$intensity_f3)
```

## Fitting a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
summary(lm(Flowers ~ Intensity + Time, data = case0901))
```

## Fitting a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# More lines of code, but more explicit
fit_1 <- lm(Flowers ~ factor(Intensity) + Time, data = case0901)
summary(fit_1)
```

## Fitting a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
fit_2 <- lm(Flowers ~ relevel(factor(Intensity), ref = '600') + Time, data = case0901)
summary(fit_2)
```

## Fitting a MLR {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0901$Intensity_new <- relevel(factor(case0901$Intensity), ref = "600")
fit_3 <- lm(Flowers ~ Intensity_new + Time, data = case0901)
summary(fit_3)
```

## Interactions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding interactions
fit_int <- lm(Flowers ~ Intensity + Time + Time:Intensity, data = case0901)
summary(fit_int)
```

## Interactions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Similar syntax which will create interaction and lower order terms
fit_int <- lm(Flowers ~ Time*Intensity, data = case0901)
summary(fit_int)
```

## Math Operators in lm() Function {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
ex0915  # the corn yield data
```

## Math Operators in lm() Function {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Add a square term
summary(lm(Yield ~ Rainfall + I(Rainfall^2), data = ex0915))
```

## Math Operators in lm() Function {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could create another variable outside of lm() then include
ex0915$RainSquared <- (ex0915$Rainfall)^2
summary(lm(Yield ~ Rainfall + RainSquared, data = ex0915))
```

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
ex0923
```

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
str(ex0923)
```

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Example 1
lm(log(Income2005) ~ Educ + Gender, data = ex0923)
```

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Example 1
summary(lm(log(Income2005) ~ Educ + Gender, data = ex0923))
```

## Examples - From R to Math {.smaller}
* `Gender` is coded as a factor and thus translated to indicator variables
  * F comes before M alphabetically hence Female is the reference group
* `Educ` is an integer and thus will be treated as a continuous variables

::: {.fragment}

\begin{align*}
  \mu\left(\ln(\text{Income}) \mid \text{Gender, Education}\right) & = \beta_{0} + \beta_{1}\text{Educ} + \beta_{2}\text{Male}
\end{align*}

:::

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Example 2
summary(lm(log(Income2005) ~ Educ + relevel(Gender, ref = "male"), data = ex0923))
```

## Examples - From R to Math {.smaller}
* The reference group for `Gender` is now change to be male rather than the previous default set to female

::: {.fragment}

\begin{align*}
  \mu\left(\ln(\text{Income}) \mid \text{Gender, Education}\right) & = \beta_{0} + \beta_{1}\text{Educ} + \beta_{2}\text{Female}
\end{align*}

:::

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Example 3
summary(lm(log(Income2005) ~ Gender + factor(Educ), data = ex0923))
```

## Examples - From R to Math {.smaller}
* The continuous variables `Educ` was changed to a factor thus we will have many indicator variables now

::: {.fragment}

\begin{align*}
  \mu\left(\ln(\text{Income}) \mid \text{Gender, Education}\right) & = \beta_{0} + \beta_{1}\text{Male} + \beta_{2}\text{Educ7} \\
  & \quad + \beta_{3}\text{Educ8} + \beta_{4}\text{Educ9} \\
  & \quad + \beta_{5}\text{Educ10} + \cdots + \beta_{15}\text{Educ20}
\end{align*}

:::

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Example 4
summary(lm(log(Income2005) ~ Educ + Gender + Educ:Gender, data = ex0923))
```

## Examples - From R to Math {.smaller}
* Treating `Educ` as a continuous variables again and adding the interaction between `Educ` and `Gender`

::: {.fragment}

\begin{align*}
  \mu\left(\ln(\text{Income}) \mid \text{Gender, Education}\right) & = \beta_{0} + \beta_{1}\text{Educ} + \beta_{2}\text{Male} \\
  & \quad + \beta_{3}\text{Educ}*\text{Male}
\end{align*}

:::

## Examples - From R to Math {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Example 5
summary(lm(log(Income2005) ~ Gender + factor(Educ) + factor(Educ):Gender, data = ex0923))
```

## Examples - From R to Math {.smaller}
* Both variables treated as factors with interactions

::: {.fragment}

\begin{align*}
  \mu\left(\ln(\text{Income}) \mid \text{Gender, Education}\right) & = \beta_{0} + \beta_{1}\text{Male} + \beta_{2}\text{Educ7} \\
  & \quad + \beta_{3}\text{Educ8} + \beta_{4}\text{Educ9} \\
  & \quad + \beta_{5}\text{Educ10} + \cdots + \beta_{15}\text{Educ20} \\
  & \quad + \beta_{16}\text{Educ7}\cdot\text{Male} + \beta_{17}\text{Educ8}\cdot\text{Male} \\
  & \quad + \cdots \beta_{29}\text{Educ20}\cdot\text{Male}
\end{align*}

:::