---
title: "(ST 412/512) Lab 5 <br> Case Influence Statistics"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: concave
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
header-includes:
  - |
    <style>
      .reveal .barelysmaller {
          font-size: 0.90em;  /* Adjust as needed */
        }
      
      .reveal .smaller {
        font-size: 0.75em;  /* Adjust as needed */
      }

      .reveal .tiny-font {
        font-size: 0.5em;  /* Even smaller font size */
        white-space: nowrap;
        overflow-x: auto;  /* Allow horizontal scrolling */
      }
    </style>
---


## Lab Objectives
* Case Influence Statistics
* Partial Residual Plots

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* `case1102` (The Blood–Brain Barrier)
  * The human brain is protected from bacteria and toxins, which course through the blood–stream, by a single layer of cells called the blood–brain barrier. These data come from an experiment (on rats, which possess a similar barrier) to study a method of disrupting the barrier by infusing a solution of concentrated sugars.
  * 34 observations (rows)
  * 9 variables (columns)
    * `Brain`: Brain tumor count (per gm)
    * `Liver`: Liver count (per gm)
    * `Time`: Sacrifice time (in hours)
    * Etc.

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1102
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Rename dataset (technically copying under a different name)
Data <- case1102
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
### Adjusting some variables
# Setting up response variable
Data$Antibody <- log(Data$Brain/Data$Liver)
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
### Adjusting some variables
# Setting up response variable
Data$Antibody <- log(Data$Brain/Data$Liver)

# Turning sacrifice time and days to categorical variables
Data$Sac <- as.factor(Data$Time)
Data$Days <- as.factor(Data$Days)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create linear model
lin_mod <- lm(Antibody ~ Sac*Treatment + Days + Sex + Weight + Loss + Tumor, data = Data)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pull residuals out of model
res <- residuals(lin_mod)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pull residuals out of model
res <- residuals(lin_mod)

# Pull fitted values out of model
fit <- fitted(lin_mod)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot the fitted versus the residuals
plot(fit, res)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot the fitted versus the residuals
plot(fit, res)
abline(0, 0, lty = 2) # Add horizontal dashed line at 0
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot the fitted versus the residuals
plot(fit, res)
abline(0, 0, lty = 2) # Add horizontal dashed line at 0
text(fit, res)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot the fitted versus the residuals
plot(fit, res, cex = 0) # Get rid of point plotting characters
abline(0, 0, lty = 2) # Add horizontal dashed line at 0
text(fit, res)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Base R linear model plots
plot(lin_mod)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Using ggplot
Data <- fortify(lin_mod, Data)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Data
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding a column for observation numbers
Data$obs_num <- 1:nrow(Data)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Data
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Saving numbers to be used in rule of thumb
n <- nrow(Data)
p <- summary(lin_mod)$df[1]
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting Leverage Values
ggplot(Data, aes(obs_num, .hat)) + 
  geom_point() + 
  geom_hline(yintercept = 2*p/n) + 
  ggtitle("Leverage")
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting Cooks (Influential) Values
ggplot(Data, aes(obs_num, .cooksd)) + 
  geom_point() + 
  ggtitle("Cook's distance")
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting Residual Values
ggplot(Data, aes(obs_num, .stdresid)) + 
  geom_point() +
  geom_hline(yintercept = 2) + 
  geom_hline(yintercept = -2) +
  ggtitle("Studentized residual")
```

## Subsetting Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Finding all the rows that have standardized residuals > 2 or < -2
subset(Data, .stdresid > 2)
subset(Data, .stdresid < -2)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Plotting all on a single plot
library(reshape)
ggplot(melt(Data[, c("obs_num",".hat",".cooksd",".stdresid")], id.vars = "obs_num"), aes(obs_num, value)) + 
  geom_point() + 
  facet_grid(variable ~ ., scale = "free_y")
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(reshape)
# What the melt() function is doing
# melt(Data[, c("obs_num",".hat",".cooksd",".stdresid")], id.vars = "obs_num")
Data[, c("obs_num",".hat",".cooksd",".stdresid")]
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What the melt() function is doing
melt(Data[, c("obs_num",".hat",".cooksd",".stdresid")], id.vars = "obs_num")
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting all on a single plot
ggplot(melt(Data[, c("obs_num",".hat",".cooksd",".stdresid")], id.vars = "obs_num"), aes(obs_num, value)) + 
  geom_point() + 
  facet_grid(variable ~ ., scale = "free_y")
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Removing rows 31 and 34 then refitting
Data2 <- Data[-c(31,34),]
lin_mod2 <- lm(Antibody ~ Sac*Treatment + Days + Sex + Weight + Loss + Tumor, data = Data2)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Removing rows 31 and 34 then refitting
lin_mod2 <- lm(Antibody ~ Sac*Treatment + Days + Sex + Weight + Loss + Tumor, data = Data2)
summary(lin_mod2)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pulling our residuals and fitted values
res2 <- residuals(lin_mod2)
fit2 <- fitted(lin_mod2)
```

## Plotting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pulling our residuals and fitted values
res2 <- residuals(lin_mod2)
fit2 <- fitted(lin_mod2)

# Plotting them
plot(fit2, res2)
abline(0, 0, lty = 2)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fitting another model
lin_mod3 <- lm(Antibody ~ Sac + Treatment + Loss, data = Data2)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fitting another model
lin_mod3 <- lm(Antibody ~ Sac + Treatment + Loss, data = Data2)
summary(lin_mod3)
```

## Fitting Linear Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Compare the two models with an F-test
anova(lin_mod3, lin_mod2)
```

## Next dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Loading in new dataset
load(url("http://stat512.cwick.co.nz/data/case_inf.rda"))
```

## Next dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Loading in new dataset
df
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting
ggplot(data = df, aes(X1, Y, colour = obs_num == 101)) + 
  geom_point()
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting
ggplot(data = df, aes(X2, Y, colour = obs_num == 101)) + 
  geom_point()
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plotting
ggplot(data = df, aes(X1, X2, colour = obs_num == 101)) + 
  geom_point()
```

## Plotting  {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
library(gridExtra)
plot1 <- ggplot(data = df, aes(X1, Y, colour = obs_num == 101)) + 
  geom_point() +
  theme_minimal() +
  theme(legend.position = "none")
plot2 <- ggplot(data = df, aes(X2, Y, colour = obs_num == 101)) + 
  geom_point() +
  theme_minimal() +
  theme(legend.position = "none")
plot3 <- ggplot(data = df, aes(X1, X2, colour = obs_num == 101)) + 
  geom_point() +
  theme_minimal() +
  theme(legend.position = "none")
grid.arrange(plot1, plot2, plot3, ncol = 3)
```

## Fitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
fit <- lm(Y ~ X1 + X2, data = df)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
fit_diag <- fortify(fit, df)
ggplot(melt(fit_diag[, c("obs_num",".hat",".cooksd",".stdresid")], id.vars = "obs_num"), 
       aes(obs_num, value, colour = obs_num == 101)) +
  geom_point() + facet_grid(variable ~ ., scale = "free_y")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Recall linear model 3 created earlier
lin_mod3 <- lm(Antibody ~ Sac + Treatment + Loss, data = Data2)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Recall linear model 3 created earlier
lin_mod3 <- lm(Antibody ~ Sac + Treatment + Loss, data = Data2)
residuals(lin_mod3, type = "partial")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Name partial residual object
p_res <- residuals(lin_mod3, type = "partial")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Name partial residual object
p_res <- residuals(lin_mod3, type = "partial")

# Change column names to be more distinct
colnames(p_res) <- paste("res_", c("Sac", "Treatment", "Loss"), sep = "")
Data_res <- cbind(Data2, p_res)
head(Data_res)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot
ggplot(data = Data_res, aes(Sac, res_Sac)) + 
  geom_point()
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot
ggplot(data = Data_res, aes(Sac, Antibody)) + 
  geom_point()
```