---
title: "(ST 412/512) Lab 6 <br> Modeling Means"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: concave
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
header-includes:
  - |
    <style>
      .reveal .barelysmaller {
          font-size: 0.90em;  /* Adjust as needed */
        }
      
      .reveal .smaller {
        font-size: 0.75em;  /* Adjust as needed */
      }

      .reveal .tiny-font {
        font-size: 0.5em;  /* Even smaller font size */
        white-space: nowrap;
        overflow-x: auto;  /* Allow horizontal scrolling */
      }
    </style>
---

## Lab Objectives
* Displaying modeled means

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* `case1301` (Seaweed Grazers)
  * The grazers were limpets (L), small fishes (f) and large fishes (F). Each plot received one of six treatments named by which grazers were allowed access. In addition, the researcher applied the treatments in eight blocks of 12 plots each. Within each block she randomly assigned treatments to plots. The blocks covered a wide range of tidal conditions.
  * 96 observations (rows)
  * 3 variables (columns)
    * `Cover`: percent of regenerated seaweed cover
    * `Block`: a factor with levels "B1", "B2", "B3", "B4", "B5", "B6", "B7" and "B8"
    * `Treat`: a factor indicating treatment, with levels "C", "f", "fF", "L", "Lf" and "LfF"

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case1301
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create log-odds ratio
case1301$log_r_ratio <- with(case1301, log(Cover/(100-Cover)))
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Check what it looks like
case1301
```

## Cell Average {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Find all cell averages
cell_avgs <- aggregate(log_r_ratio ~ Block + Treat, data = case1301, FUN = mean)
```

## Cell Average {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Find all cell averages
cell_avgs <- aggregate(log_r_ratio ~ Block + Treat, data = case1301, FUN = mean)
cell_avgs
```

## Cell Average {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# What this code is doing
round(addmargins(xtabs(log_r_ratio ~ Block + Treat, cell_avgs), FUN = mean), 2)
```

## Cell Average {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What this code is doing
xtabs(log_r_ratio ~ Block + Treat, data = cell_avgs)
```

## Cell Average {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What this code is doing
addmargins(xtabs(log_r_ratio ~ Block + Treat, data = cell_avgs), FUN = mean)
```

## Cell Average {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# What this code is doing
round(addmargins(xtabs(log_r_ratio ~ Block + Treat, data = cell_avgs), FUN = mean), 2)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(Block, log_r_ratio, data = cell_avgs, colour = Treat) +
  ylab("Average log regeneratio ratio")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = cell_avgs, aes(x = Block, y = log_r_ratio, color = Treat)) +
  geom_point() +
  labs(y = "Average Log Regeneratio Ratio")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Reordering Block by the ordering of the log ratio
qplot(reorder(Block, log_r_ratio), log_r_ratio, data = cell_avgs,
  colour = Treat, geom = "line", group = Treat) + 
  ylab("Average log regeneratio ratio")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = cell_avgs, aes(x = reorder(Block, log_r_ratio), y = log_r_ratio, 
                             color = Treat, group = Treat)) +
  geom_line() +
  labs(y = "Average Log Regeneratio Ratio")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Adding points to plot
qplot(reorder(Block, log_r_ratio), log_r_ratio, data = cell_avgs,
  colour = Treat, geom = "line", group = Treat) + 
  ylab("Average log regeneratio ratio") +
  geom_point(data = cell_avgs)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = cell_avgs, aes(x = reorder(Block, log_r_ratio), y = log_r_ratio, 
                             color = Treat, group = Treat)) +
  geom_line() +
  geom_point() +
  labs(y = "Average Log Regeneratio Ratio")
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating additive model
fit_add <- lm(log_r_ratio ~ Block + Treat, data = case1301)
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating additive model
fit_add <- lm(log_r_ratio ~ Block + Treat, data = case1301)

# Creating saturated model
fit_sat <- lm(log_r_ratio ~ Block + Treat + Block:Treat, data = case1301)
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating all explanatory combinations (in a new object)
new_data <- expand.grid(
  Block = unique(case1301$Block),
  Treat = unique(case1301$Treat))
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating all explanatory combinations (in a new object)
new_data <- expand.grid(
  Block = unique(case1301$Block),
  Treat = unique(case1301$Treat))

# Predicting all explanatory value combinations on additive model
new_data$pred <- predict(fit_add, newdata = new_data)
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Looking at new_data object we're creating
new_data
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Present as a table
round(xtabs(pred ~ Block + Treat, new_data), 2)
```

## Plotting Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting reordered Block and predictions for additive model
qplot(reorder(Block, pred), pred, data = new_data, 
      colour = Treat, geom = "line", group = Treat)
```

## Plotting Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = new_data, aes(x = reorder(Block, pred), y = pred, 
                             color = Treat, group = Treat)) +
  geom_line()
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Predicting all explanatory value combinations on saturated model
new_data$pred_sat <- predict(fit_sat, newdata = new_data)
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Looking at new_data object we're creating
new_data
```

## Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Present as a table
round(xtabs(pred_sat ~ Block + Treat, new_data), 2)
```

## Plotting Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting reordered Block and predictions for saturated model
qplot(reorder(Block, pred_sat), pred_sat, data = new_data,
      colour = Treat, geom = "line", group = Treat)
```

## Plotting Predicting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = new_data, aes(x = reorder(Block, pred_sat), y = pred_sat, 
                             color = Treat, group = Treat)) +
  geom_line()
```

## Next Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Load another package
library(MASS)
```

## Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
* `ex1416` (Blood Brain Barrier)
  * Researchers designed an experiment to investigate how delivery of brain cancer antibody is influenced by tumor size, antibody molecular weight, blood-brain barrier disruption, and delivery route.
  * 36 observations (rows)
  * 6 variables (columns)
    * `Agent`: a factor with levels "AIB", "DEX7" and "MTX"
    * `Treatment`: a factor with levels "BD" and "NS"
    * `Route`: a factor with levels "IA" and "IV"
    * `DaysPost`: days after inoculation
    * `BAT`: concentration of antibody in the part of the brain around the tumor
    * `LH`: concentration of antibody in the unaffected part of the brain
    
## Next Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Next data set
ex1416
```

## Next Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Renaming data as "Data1" (optional)
Data1 <- ex1416
```

## Next Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating repsonse variables
Data1$Response <- Data1$BAT/Data1$LH
```

## Next Dataset {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating repsonse variables
Data1$Response <- Data1$BAT/Data1$LH

# Changing DaysPost to a factor variable
Data1$fac.DaysPost <- factor(Data1$DaysPost)
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Not sure why the points are made to be so small
# Grouping by `Treatment'
ggplot(data = Data1, aes(x = fac.DaysPost, y = Response, group = Treatment, colour = Treatment)) +
  geom_point(size = 1) + 
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"))
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Fixing point size
# Grouping by `Treatment'
ggplot(data = Data1, aes(x = fac.DaysPost, y = Response, group = Treatment, colour = Treatment)) +
  geom_point() + 
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"))
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Removed "size = 1" from lab notes
# Grouping by `Route'
ggplot(data = Data1, aes(x = fac.DaysPost, y = Response, group = Route, colour = Route)) +
  geom_point() + 
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"))
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Removed "size = 1" from lab notes
# Grouping by `Agent'
ggplot(data = Data1, aes(x = fac.DaysPost, y = Response, group = Agent, colour = Agent)) +
  geom_point() + 
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"))
```

## Fitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Create linear model on new dataset
lmod1 <- lm(Response ~ fac.DaysPost + Treatment + Agent + Route, data = Data1)
```

## Fitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Fortify data
lmod1 <- lm(Response ~ fac.DaysPost + Treatment + Agent + Route, data = Data1)
Data2 <- fortify(lmod1, Data1)
```

## Fitting Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Looking at Data2
Data2
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting Fitted vs. Residuals
ggplot(data = Data2, aes(.fitted, .resid)) + 
  geom_point() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"))
```

## Box-Cox Transformation {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Checking for transformation
boxcox(lmod1)
```

## Box-Cox Transformation {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
$$
y^{(\lambda)} = \begin{cases}
  \frac{y^{\lambda} - 1}{\lambda}, & \text{if } \lambda \neq 0 \\[0.25cm]
  \ln(y), & \text{if } \lambda = 0
\end{cases}
$$

* Where $y$ is our positive response variable
* and $\lambda$ is the transformation parameter that determines the shape of the transformation, notice if
  * $\lambda = 1$ it is the identity function and thus no transformation
  * $\lambda = 0$ implies a natural log transformation
  * $\lambda = -1$ implies a reciprocal transformation $\frac{1}{y}$
  * $\lambda = 0.5$ implies a square root transformation
  * etc.
  
## Box-Cox Transformation {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Performing a log transformation
Data1$log.Response <- log(Data1$Response)
```

## Fitting Log Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
lmod2 <- lm(log.Response ~ fac.DaysPost + Treatment + Agent + Route, data = Data1)
Data3 <- fortify(lmod2, Data1)
```

## Fitting Log Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fortify data
lmod2 <- lm(log.Response ~ fac.DaysPost + Treatment + Agent + Route, data = Data1)
Data3 <- fortify(lmod2, Data1)
```

## Fitting Log Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Looking at Data3
Data3
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting Fitted vs. Residuals
ggplot(data = Data3, aes(.fitted, .resid)) + 
  geom_point() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"))
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting reordered DaysPost with fitted values - grouped by Treatment
ggplot(data = Data3, aes(x = reorder(DaysPost, Response), y = .fitted, 
                         group = Treatment, color = Treatment)) + 
  geom_point() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold")) +
  labs(y = "Mean log(Reponse)")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting reordered DaysPost with fitted values - grouped by Route
ggplot(data = Data3, aes(x = reorder(DaysPost, Response), y = .fitted, 
                         group = Route, color = Route)) + 
  geom_point() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold")) +
  labs(y = "Mean log(Reponse)")
```

## Plotting {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting reordered DaysPost with fitted values - grouped by Agent
ggplot(data = Data3, aes(x = reorder(DaysPost, Response), y = .fitted, 
                         group = Agent, color = Agent)) + 
  geom_point() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold")) +
  labs(y = "Mean log(Reponse)")
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}

* With 4 factors:
  * `Agent`: Factor w/ 3 levels "AIB", "DEX7" and "MTX"
  * `Treatment`: Factor w/ 2 with levels "BD" and "NS"
  * `Route`: Factor w/ 2 levels "IA" and "IV"
  * `DaysPost`: Factor w/ 3 levels "8","12","16"
* This means there are $3 \cdot 2 \cdot 2 \cdot 3 = 36$ cells
* Recall we only have 36 observations, thus this model cannot be fit, if we try...

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Trying to fit saturated model
lmod3 <- lm(log.Response ~ fac.DaysPost + Treatment + Agent + Route + 
              fac.DaysPost * Treatment * Agent * Route,
            data = Data1)
summary(lmod3)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Uh-Oh
summary(lmod3)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Uh-Oh
anova(lmod3)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Fitting model with continuous version of DaysPost
lmod4 <- lm(log.Response ~ DaysPost + Treatment + Agent + Route + 
              DaysPost * Treatment * Agent * Route, 
            data = Data1)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Nice
summary(lmod4)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
anova(lmod4)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Removing `Agent` from model
lmod5 <- lm(log.Response ~ DaysPost + Treatment + Route + 
              DaysPost * Treatment * Route, 
            data = Data1)
```

## Fitting Saturated Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# It works!
summary(lmod5)
```