---
title: "(ST 412/512) Lab 1 <br> Review of Regression Diagnostic Plots"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## NOTICE THERE ARE TWO DIFFERENT DATASETS THE LAB PDF DOES NOT MATCH THE R CODE


```{r}
#| echo: FALSE
library(gridExtra)
```

## Hello! {.smaller}
* Who am I?
  * A deep philosophical question
  * So instead I will answer any questions anyone may have about my identity
  * If no questions I am content to be anonymous for the term and we can continue
* Office Hours
  * Monday 1:00pm - 2:00pm
  * Monday 3:00pm - 4:00pm
* Lab Structure
  * I'll yap for a while
  * After I have put everyone to sleep, I'll go around answering any homework (or whatever) questions

## R we Ready to go?
* Hopefully everyone has R and RStudio installed on their system

## Lab Objectives
* Long objective list today...
  * Review of regression diagnostic plots

## Why Plots? {.smaller}
* To assess the common assumptions underpinning our models, we often look at what we refer to as diagnostic plots
  * Some of plots that might be familiar from ST 411/511 include:
    * Residual vs fitted values
    * Residuals vs explanatory variable(s)
    * Normal probability plot (QQ plot)
* Though we can never be certain about the assumptions from mere plot inspections they do offer some insight into the model's fit

## Loading Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3) # Package that contains the data
library(ggplot2) # For plotting
library(Rmisc) # For plotting
```

## Loading Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# If you are curious about a dataset's details
help(ex0824)
```

## Loading Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# If you are curious about a dataset's details
help(ex0824)

# Or equivalently
?ex0824
```

## Dataset {.smaller}
* `ex0824` (Respiratory Rates for Children)
* **Description**: A high respiratory rate is a potential diagnostic indicator of respiratory infection in children. To judge whether a respiratory rate is “high” however, a physician must have a clear picture of the distribution of normal rates. To this end, Italian researchers measured the respiratory rates of 618 children between the ages of 15 days and 3 years.
* 618 observations
* 2 variables:
  * `Age`: age in months of child 
  * `Rate`: respiratory rate (breaths per minute)
    
## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
ex0824
```

## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
ex0824
```

## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# First 6 observations (rows)
head(ex0824)
```

## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# First 6 observations (rows)
head(ex0824)

# Last 6 observations (rows)
tail(ex0824)
```
    
## Fit Linear Regression {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Assigning the dataset a different name
Data <- ex0824
```

## Fit Linear Regression {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Loading Data
Data <- ex0824

# Running Simple Linear Regression on the Data
lin_mod <- lm(Mortality ~ Wine, data = Data)
```

## Fit Linear Regression {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Running Simple Linear Regression on the Data
lin_mod <- lm(Mortality ~ Wine, data = Data)
summary(lin_mod)
```

## Fit Linear Regression {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Updating our dataset to include the information from the model
Data <- fortify(lin_mod, Data)
```

## Fit Linear Regression {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Viewing the updated dataset
Data
```

## Plot Linear Regression Fit {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plot the scatterplot with fitted line
ggplot(aes(Wine, Mortality), data = Data) + 
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE)
```

## Plot Linear Regression Fit {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Adjust size of text in axis labels        
ggplot(aes(Wine, Mortality), data = Data) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) +
  theme(text = element_text(size = 30),
        axis.text.x = element_text(angle = 90, vjust = 1))
```

## Plot Linear Regression Fit {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Using prebuilt themes   
ggplot(aes(Wine, Mortality), data = Data) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) +
  theme_minimal()
```

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Fitted vs Residual Plot
ggplot(aes(.fitted, .resid), data = Data) + 
  geom_point() + 
  geom_hline(aes(yintercept = 0)) +
  theme_minimal()
```

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Explanatory variable vs Residual Plot
ggplot(aes(Wine, .resid), data = Data) + 
  geom_point() +
  geom_hline(aes(yintercept = 0)) +
  theme_minimal()
```

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
* What are we looking for?
  * We want a boring plot, no apparent patterns
  * More specifically, we are looking for a constant variation in points with no clear pattern
  * An a typical pattern might present in a number of different ways:
    * Curvature
    * Clustering
    * Funnel shaped
    * Etc.

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# QQ-Plot for checking normality
ggplot(aes(sample = .resid), data = Data) + 
  geom_qq() +
  stat_qq_line() +
  theme_minimal()
```

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
* What are we looking for?
  * Ideally we see the points oscillate closely around the reference line

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
#| code-line-numbers: "4"
# QQ-Plot for checking normality
ggplot(aes(sample = .resid), data = Data) + 
  geom_qq() +
  stat_qq_line() + # Notice the function is corrected from the notes
  theme_minimal()
```

## Plotting Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Fitted vs Residual Plot
plot1 <- ggplot(aes(.fitted, .resid), data = Data) + 
  geom_point() + 
  geom_hline(aes(yintercept = 0)) +
  labs(title = "Fitted vs Residual Plot") +
  theme_minimal()

# Explanatory variable vs Residual Plot
plot2 <- ggplot(aes(Wine, .resid), data = Data) + 
  geom_point() +
  geom_hline(aes(yintercept = 0)) +
  labs(title = "Explanatory variable vs Residual Plot") +
  theme_minimal()

# QQ-Plot for checking normality
plot3 <- ggplot(aes(sample = .resid), data = Data) + 
  geom_qq() +
  stat_qq_line() +
  labs(title = "QQ-Plot of Residuals") +
  theme_minimal()

multiplot(plot1, plot2, plot3, cols = 3)
```

## Log-Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plot the log-Transformed scatterplot with fitted line
ggplot(aes(log(Wine), log(Mortality)), data = Data) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE) +
  theme_minimal()
```

## Log-Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding the log-transformed data as new variables
Data$log.Wine <- log(Data$Wine)
```

## Log-Transformation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding the log-transformed data as new variables
Data$log.Wine <- log(Data$Wine)
Data$log.Mortality <- log(Data$Mortality)
```

## [Log-Transformed Response Linear Regression]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Running Simple Linear Regression on the log-Transformed Data
lin_mod1 <- lm(log.Mortality ~ log.Wine, data = Data)
```

## [Log-Transformed Response Linear Regression]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Running Simple Linear Regression on the log-Transformed Data
lin_mod1 <- lm(log.Mortality ~ log.Wine, data = Data)
summary(lin_mod1)
```

## [Log-Transformed Response Linear Regression]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Updating our dataset to include the information from the new model
Data <- fortify(lin_mod1, Data) # This will overwrite the other model's information
```

## [Log-Transformed Response Linear Regression]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Viewing the updated dataset
Data
```

## Reassessing Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Fitted vs Residual Plot
ggplot(aes(.fitted, .resid), data = Data) + 
  geom_point() +
  geom_hline(aes(yintercept = 0)) +
  theme_minimal()
```

## Reassessing Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Explanatory variable vs Residual Plot
ggplot(aes(Wine, .resid), data = Data) + 
  geom_point() +
  geom_hline(aes(yintercept = 0)) +
  theme_minimal()
```

## Reassessing Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# QQ-Plot for checking normality
ggplot(aes(sample = .resid), data = Data) + 
  stat_qq() + 
  stat_qq_line() + # Notice the function is corrected from the notes
  theme_minimal()
```

## Reassessing Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Fitted vs Residual Plot
plot1 <- ggplot(aes(.fitted, .resid), data = Data) + 
  geom_point() + 
  geom_hline(aes(yintercept = 0)) +
  labs(title = "Fitted vs Residual Plot") +
  theme_minimal()

# Explanatory variable vs Residual Plot
plot2 <- ggplot(aes(Wine, .resid), data = Data) + 
  geom_point() +
  geom_hline(aes(yintercept = 0)) +
  labs(title = "Explanatory variable vs Residual Plot") +
  theme_minimal()

# QQ-Plot for checking normality
plot3 <- ggplot(aes(sample = .resid), data = Data) + 
  geom_qq() +
  stat_qq_line() +
  labs(title = "QQ-Plot of Residuals") +
  theme_minimal()

multiplot(plot1, plot2, plot3, cols = 3)
```

## Reassessing Residual Plots {auto-animate="true" auto-animate-easing="ease-in-out"}
* Do the residual plots look more like we hope?

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text}
- Data transformations are often suggested for skewed data, data that "doesn't look Normal"
- A common transformation is the natural log
  - But, when we are interested in the population mean (as is the case in this class)
- Without a big assumption, we are talking about something called a geometric mean on the original scale

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
- What is taught in ST 411/511 is the interpretation of a median on the original scale
- To make this interpretation we've got to make some big leaps in assumption land
  - If we assume the log-transformed data have a symmetric distribution, we know

::: {.fragment}
$$
\text{Mean}\left[\log(Y)\right] = \text{Median}\left[\log(Y)\right]
$$
:::
- The log preserves ordering meaning that the median of the logs is the log of the medians

::: {.fragment}
$$
\text{Median}\left[\log(Y)\right] = \log\left[\text{Median}(Y)\right]
$$
:::

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
- So in a two sample case (instead of linear regression) the logic goes as such:

::: {.fragment}
$$
\frac{1}{n}\sum\limits_{i=1}^{n}\log(X_{i}) - \frac{1}{m}\sum\limits_{i=1}^{m}\log(Y_{i}) = \bar{U} - \bar{W}
$$
:::

::: {.fragment}
Estimates
$$
\text{Mean}\left[\log(X)\right] - \text{Mean}\left[\log(Y)\right]
$$
:::

::: {.fragment}
If we assume the log-transformed data have a symmetric distribution
$$
\text{Median}\left[\log(X)\right] - \text{Median}\left[\log(Y)\right]
$$
:::

::: {.fragment}
Log preserves ordering 
$$
\log\left[\text{Median}(X)\right] - \log\left[\text{Median}(Y)\right] = \log\left(\frac{\text{Median}(X)}{\text{Median}(Y)}\right)
$$
:::

::: {.fragment}
Exponentiating we have
$$
e^{\bar{U} - \bar{W}} \quad \text{ estimates } \quad \frac{\text{Median}(X)}{\text{Median}(Y)}
$$
:::

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text}
- Notice that symmetry is not technically required, we just need a setting where $\text{Mean}\left[\log(Y)\right] = \text{Median}\left[\log(Y)\right]$, symmetry just guarantees this
- To demonstrate this assumption, I will show the best case and a seemingly indistinguishable case to that of the best that does not work
- Assumption: If we have right skewed positive data that once transformed has a **symmetric distribution**

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating right-skewed positive data (log-normal data)
da.data <- data.frame(vals = rlnorm(10000, mean = 0, sd = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Log-Normal Data)") +
  xlim(0, 20) +
  theme_bw()
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Natural log of that data
ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
p1 <- ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Log-Normal Data)") +
  xlim(0, 20) +
  theme_bw()

p2 <- ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- The right skewed data shown was log-normal so unless you assume that your data is log-normal you won't have completely Normal data (symmetric) once transformed
- Aka, just because it is positive and right-skewed, doesn't mean everything is golden
  - The couple following of slides will illustrate this
  
## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating right-skewed positive data (exponential data)
da.data <- data.frame(vals = rexp(10000, rate = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Exponential)") +
  xlim(0, 20) +
  theme_bw()
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Natural log of that data
ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Plotting side-by-side
p3 <- ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Exponential)") +
  xlim(0, 20) +
  theme_bw()
p4 <- ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
grid.arrange(p3, p4, ncol = 2)
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- We can see that since our original data was not log-normal we don't end up with a symmetric distribution

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text}
- Notice, I mentioned positively skewed data, we cannot have negative values before we log-transform (or the world will explode)
- Some people try to circumvent this by shifting the data beforehand but... 
  - then interpretation becomes challenging / hopeless and dependent on a perhaps arbitrary shift parameter
  
## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
* Thus a one unit increase in $X$ corresponds to a multiplicative increase of $e^{\beta_{1}}$ in the geometric mean of $Y$
* The *big assumption* of our data being symmetric after log transforming is sometimes made (although it probably shouldn't be) which leads to a slightly different interpretation
  * A one unit increase in $X$ corresponds to a multiplicative increase of $e^{\beta_{1}}$ in the median of $Y$

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Generating Data (X is randomly generated)
# For reproducibility, can set a seed
set.seed(23)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Generating Data (X is randomly generated)
# For reproducibility, can set a seed
set.seed(23)

# n represents the sample size
n <- 200
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# n represents the sample size
n <- 200

# Continuous Explanatory
x <- rnorm(n, 0, 1)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Continuous Explanatory
x <- rnorm(n, 0, 1)

# Discrete Explanatory
x <- sample(0:4, n, replace = T)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Plotting Function for a linear case
plot_diagnostics <- function(n = 200, Continuous = TRUE) {
  if(Continuous == TRUE){
    x <- rnorm(n, 0, 1)
  } else {
    x <- sample(0:4, n, replace = T)
  }
  # Initialize our parameters
  e <- rnorm(n, 0, 1)
  alpha <- 4
  beta <- 7
  y <- alpha + beta * x + e
  
  # Create our dataframe
  data <- as.data.frame(cbind(x, y))
  colnames(data) <- c("x", "y")
  
  # Create linear model
  lmod <- lm(y ~ x)
  
  # Combine data and model output
  data <- fortify(lmod, data)
  
  # Plots
  plot1 <- qplot(.fitted, .resid, data = data) +
    xlab("fitted values") + 
    ylab("residuals") + 
    geom_line(y = 0, col = 2)
  
  plot2 <- qplot(x, .resid, data = data) +
    xlab("Explanatory variable") + 
    ylab("residuals") + 
    geom_line(y = 0, col = 2)
  
  plot3 <- qplot(sample = data$.resid) + 
    stat_qq_line()
  
  multiplot(plot1, plot2, plot3, cols = 2)
}
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics(n = 20, Continuous = TRUE)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics(n = 20, Continuous = FALSE)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics(n = 50, Continuous = TRUE)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics(n = 50, Continuous = FALSE)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics(n = 200, Continuous = TRUE)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics(n = 200, Continuous = FALSE)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Plotting Function for a non-linear case
plot_diagnostics2 <- function(n = 200) {
  # Only continuous option now
  x <- rnorm(n, 0, 1)
  
  # Initialize our parameters
  e <- rnorm(n, 0, 1)
  alpha <- 4
  beta <- 7
  y <- alpha + beta * (x^(4)) + e
  
  # Create our dataframe
  data <- as.data.frame(cbind(x, y))
  colnames(data) <- c("x", "y")
  
  # Create linear model
  lmod <- lm(y ~ x)
  
  # Combine data and model output
  data <- fortify(lmod, data)
  
  # Plots
  plot1 <- qplot(.fitted, .resid, data = data) +
    xlab("fitted values") + 
    ylab("residuals") + 
    geom_line(y = 0, col = 2)
  
  plot2 <- qplot(x, .resid, data = data) +
    xlab("Explanatory variable") + 
    ylab("residuals") + 
    geom_line(y = 0, col = 2)
  
  plot3 <- qplot(sample = data$.resid) + 
    stat_qq_line()
  
  multiplot(plot1, plot2, plot3, cols = 2)
}
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics2(n = 20)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics2(n = 50)
```

## More plots! {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot_diagnostics2(n = 200)
```