---
title: "(ST 412/512) Lab 4 <br> Inference on Multiple Linear Regression in R"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: concave
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
header-includes:
  - |
    <style>
      .reveal .barelysmaller {
          font-size: 0.90em;  /* Adjust as needed */
        }
      
      .reveal .smaller {
        font-size: 0.75em;  /* Adjust as needed */
      }

      .reveal .tiny-font {
        font-size: 0.5em;  /* Even smaller font size */
        white-space: nowrap;
        overflow-x: auto;  /* Allow horizontal scrolling */
      }
    </style>
---


## Lab Objectives
* F-test
* Estimated means, confidence intervals and prediction intervals

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3)
library(ggplot2)
```

## Dataset {.smaller}
* `case0901` (Effects of Light on Meadowfoam Flowering)
  * Shining lights on some plants to see if they would grow?
  * 24 observations (rows)
  * 3 variables (columns)
    * `Flowers`: average number of flowers per meadowfoam plant
    * `Time`: time light intensity regiments started; 1 = Late, 2 = Early
    * `Intensity`: light intensity (in some wacky unit, idk)

## ANOVA F Test {.smaller}
* When two models are "nested" they can be compared using the ANOVA (ESS) F Test
  * What is a nested model?
    * A model is said to be nested in another if it is a special case of it for example we could look the model:
    
::: {.fragment}

$$
\begin{align*}
\mu\left(\text{Flowers} \mid \text{Intensity}, \text{Time}\right) & = \beta_{0} + \beta_{1}\text{Intensity300} + \beta_{2}\text{Intensity450} \\
& \quad + \beta_{3}\text{Intensity600} + \beta_{4}\text{Intensity750} \\
& \quad + \beta_{5}\text{Intensity900} + \beta_{6}\text{Time}
\end{align*}
$$

:::

* So instead of treating `intensity` as a factor we can make special case by treating it as a continuous variable which leads to a nested model

::: {.fragment}

$$
\mu\left(\text{Flowers} \mid \text{Intensity}, \text{Time}\right) = \beta_{0} + \beta_{1}\text{Intensity} + \beta_{2}\text{Time}
$$

:::

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Categorical Intensity Variable
fit_indic <- lm(Flowers ~ factor(Intensity) + Time, data = case0901)
```

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Categorical Intensity Variable
fit_indic <- lm(Flowers ~ factor(Intensity) + Time, data = case0901)

## Continuous Intensity Variable
fit_contin <- lm(Flowers ~ Intensity + Time, data = case0901)
```

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Comparing Models
anova(fit_contin, fit_indic)
```

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* Perhaps an obvious model that can be found nested in another is the null model, that is

::: {.fragment}

$$
\mu\left(\text{Flowers}\right) = \beta_{0} 
$$

:::

* We can again use the `anova()` function to get an ANOVA table to do an $F$-test
  * We don't have to formally create the null model since R does it for us but I will show they are the same

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating the null model, though it is not required
fit_null <- lm(Flowers ~ 1, data = case0901)
```

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating the null model, though it is not required
fit_null <- lm(Flowers ~ 1, data = case0901)

# Same model as before with Intensity as continuous
fit_contin <- lm(Flowers ~ Intensity + Time, data = case0901)
```

## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Comparing Models
anova(fit_null, fit_contin)
```


## ANOVA F Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
## Equivalent to just having the single model in the function
## Though it looks slightly different since this is the generic ANOVA table
anova(fit_contin)
```

## ANOVA Table Exercise {.barelysmaller}

* The following table summarizes the results of an ANOVA F-test:

::: {.fragment}

```{=latex}
\begin{array}{|c|c|c|c|}
\hline
\textbf{Source} & \textbf{DF} & \textbf{Sum of Squares} & \textbf{Mean Square} \\ \hline
\text{Regression} & 2 & 3466.7 & 1733.35 \\ \hline
\text{Residual} & 21 & 871.24 & 41.49 \\ \hline
\text{Total} & 23 & 4337.94 & \\ \hline
\end{array}
```

:::

* The task is to find the $p$-value, aka complete the table

## ANOVA Table Exercise {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Manual way
F.stat <- 1733.35/41.49
```

## ANOVA Table Exercise {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Manual way
F.stat <- 1733.35/41.49
p.value <- 1 - pf(F.stat, 2, 21)
p.value
```

## ANOVA Table Exercise {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pulling from summary() output
summary(lm(Flowers ~ Intensity + Time, data = case0901))$fstatistic
```

## ANOVA Table Exercise {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Pulling from summary() output
F.value <- summary(lm(Flowers ~ Intensity + Time, data = case0901))$fstatistic

# p-value of the F-test
pf(F.value[1], F.value[2], F.value[3], lower.tail=FALSE)
```

## Dataset {.smaller}
* `ex0923` (Comparing Male and Female Incomes)
  * 2,584 observations (rows)
  * 5 variables (columns)
    * `Subject`: the subject identification number
    * `Gender`: a factor with levels "female" and "male"
    * `AFQT`: percentile score on the AFQT intelligence test
    * `Educ`: years of education achieved by 2005
    * `Income2005`: annual income in 2005

## Dataset {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
ex0923
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Model we will use
fit <- lm(log(Income2005) ~ AFQT + Educ + Gender,
          data = ex0923, na.action = na.exclude)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Model we will use
summary(fit)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating a grid of all possible combinations of explanatory values
newdf <- expand.grid(
  Educ = unique(ex0923$Educ),
  AFQT = seq(0, 100, 10),
  Gender = unique(ex0923$Gender)
)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Creating a grid of all possible combinations of explanatory values
newdf <- expand.grid(
  Educ = unique(ex0923$Educ),
  AFQT = seq(0, 100, 10),
  Gender = unique(ex0923$Gender)
)
newdf
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# We can then use predict to estimate the mean response for each combination of explanatory variables
pred_values <- predict(fit, newdata = newdf)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# We can then use predict to estimate the mean response for each combination of explanatory variables
pred_values <- predict(fit, newdata = newdf)
pred_values
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Just renaming to perhaps something more clear
# and attaching it to our grip of explanatory variables
newdf$est_mean <- predict(fit, newdata = newdf)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
newdf
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(AFQT, est_mean, data = subset(newdf, Educ == 12), 
      geom = "line", colour = Gender) + 
  ylab("Log Income")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ == 12), aes(x = AFQT, y = est_mean)) +
  geom_line(aes(color = Gender)) +
  labs(y = "Log Income")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ == 12), aes(x = AFQT, y = est_mean)) +
  geom_line(aes(color = Gender)) +
  labs(y = "Log Income") +
  scale_color_manual(
    values = c("female" = "orange", "male" = "green"),
    labels = c("Female", "Male")) +
  theme_bw()
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qplot(AFQT, exp(est_mean), data = subset(newdf, Educ == 12), 
      geom = "line", colour = Gender) + 
  ylab("Estimated median income")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ == 12), aes(x = AFQT, y = exp(est_mean))) +
  geom_line(aes(color = Gender)) +
  labs(y = "Estimated median income")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Confidence interval time!
pred_interval <- predict(fit, newdata = newdf, interval = "confidence")
head(pred_interval)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}

:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
# Confidence interval time!
pred_interval <- predict(fit, newdata = newdf, interval = "confidence")
pred_interval
```

:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
# Predictions
pred_values <- predict(fit, newdata = newdf)
pred_values
```

::: 

::::

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Confidence interval time!
cis <- predict(fit, newdata = newdf, interval = "confidence")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Confidence interval time!
cis <- predict(fit, newdata = newdf, interval = "confidence")
newdf <- cbind(cis, newdf)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Updated dataframe
newdf
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Back to plotting
# Line for mean response
qplot(AFQT, fit, data = subset(newdf, Educ==12 & Gender == "male"), 
      geom = "line")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ==12 & Gender == "male"), aes(x = AFQT, y = fit)) +
  geom_line()
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Add another line for upper limit of CI
qplot(AFQT, fit, data = subset(newdf, Educ==12 & Gender == "male"), 
      geom = "line") + 
  geom_line(aes(y = upr), colour = "red")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ==12 & Gender == "male"), aes(x = AFQT, y = fit)) +
  geom_line() +
  geom_line(aes(y = upr), color = "red")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Add another for the lower limits
qplot(AFQT, fit, data = subset(newdf, Educ == 12 & Gender == "male"), 
      geom = "line") + 
  geom_line(aes(y = upr), colour = "red") + 
  geom_line(aes(y = lwr), colour = "red", linetype = "dashed")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ==12 & Gender == "male"), aes(x = AFQT, y = fit)) +
  geom_line() +
  geom_line(aes(y = upr), color = "red") +
  geom_line(aes(y = lwr), color = "red", linetype = "dashed")
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Add a ribbon for the interval instead
qplot(AFQT, fit, data = subset(newdf, Educ == 12 & Gender == "male"), 
      geom = "line") + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ==12 & Gender == "male"), aes(x = AFQT, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Create a line for each gender
qplot(AFQT, fit, data = subset(newdf, Educ == 12), 
      colour = Gender, geom = "line") + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ == 12), aes(x = AFQT, y = fit, color = Gender)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Estimated median response on the original scale just back-transform 
# the mean and interval endpoints - Under symmetry assumption
qplot(AFQT, exp(fit), data = subset(newdf, Educ == 12 & Gender == "male"), 
      geom = "line") +
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)), linetype = "dotted", alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ == 12 & Gender == "male"), aes(x = AFQT, y = exp(fit))) +
  geom_line() +
  geom_ribbon(aes(ymin = exp(lwr), ymax = exp(upr)), linetype = "dotted", alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Prediction intervals!
pis <- predict(fit, newdata = newdf, interval = "prediction")
head(pis)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Prediction intervals!
pis <- predict(fit, newdata = newdf, interval = "prediction")
pis
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Prediction intervals!
colnames(pis) <- c("pi_fit", "pi_lwr", "pi_upr")
newdf <- cbind(pis, newdf)
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Prediction intervals!
newdf
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plotting prediction intervals
qplot(AFQT, fit, data = subset(newdf, Educ == 12 & Gender == "male"), 
      geom = "line") + 
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", alpha = I(0.2)) +
  geom_ribbon(aes(ymin = pi_lwr, ymax = pi_upr), alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# ggplot() syntax
ggplot(data = subset(newdf, Educ == 12 & Gender == "male"), aes(x = AFQT, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", alpha = I(0.2)) +
  geom_ribbon(aes(ymin = pi_lwr, ymax = pi_upr), alpha = I(0.2))
```

## Confidence and Prediction Intervals {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# A nicer plot...
ggplot(data = subset(newdf, Educ == 12 & Gender == "male"), aes(x = AFQT, y = fit)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), linetype = "dotted", color = "red", fill = "pink", alpha = I(0.8)) +
  geom_ribbon(aes(ymin = pi_lwr, ymax = pi_upr), color = "blue", fill = "lightblue", alpha = I(0.333)) +
  labs(
    y = "Fitted Mean Response",
    title = "Regression Line with Confidence and Prediction Intervals") +
  theme_bw()
```