---
title: "(ST 412/512) Lab 9 <br> Serial Correlation"
format:
  revealjs:
    incremental: true
    scrollable: true
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: true
    chalkboard: true
    code-line-numbers: true
    css: (ST 512) StyleSheets.css
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

## Welcome Back
* Any questions before we get started?
* Office Hours: Thursday 10:00am - 11:00am

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
case1502 <- read.csv("case1502.csv", header = TRUE) 
```

## Loading Data and Packages {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
library(Sleuth3) # Contains case1502 dataset
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
head(case1502, 10) # Print first 10 observations
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(case1502$Year, case1502$Temperature,
  main = "Annual temperature anomalies from 1850 to 2010",
  xlab = "Year",
  ylab = "Temperature in C",
  col = "blue",
  pch = 20,
  type = "b"
)
abline(h = 0, lty = "dashed")
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\mu\left\{Y \mid \text{year}\right\} = \beta_{0} + \beta_{1}\text{year} + \beta_{2}\text{year}^{2}
$$

```{r}
#| echo: TRUE
# One might argue a curved trend in the scatter plot
mod.raw <- lm(Temperature ~ Year + I(Year^2), data = case1502)
summary(mod.raw)
```

## Checkin' Out Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Checking out the resulting residuals
plot(case1502$Year, mod.raw$res,
  col = "black",
  pch = 20,
  type = "b",
  ylab = "Residuals",
  xlab = "Years"
)
abline(h = 0, lty = "dashed")
```

## Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
- From the residual plot, one might argue that we see the points move above or below the center line in **chunks**. In other words:
  - When the residuals are **positive**, they tend to remain positive for a while.
  - When the residuals are **negative**, they tend to remain negative for a while.
- This behavior breaks the **random pattern** that we expect from residuals and suggests the **presence of serial correlation** in the data.

## Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
**Serial correlation** occurs when observations in a time series are influenced by their **recent past**.

- For example:
  - **First-order serial correlation** occurs when the observation at time $t$ depends on the observation at time $t-1$. If this occurs, it violates the **assumption of independence**.

## Checking Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
How can we check whether this is happening? One approach is to examine the **correlation between consecutive residuals**.

- We can do this by constructing a **scatterplot of lagged residuals**:
  - Residuals for years **1851–2010**
  - vs residuals for years **1850–2009**

## Checking Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(mod.raw$res[1:160], mod.raw$res[2:161],
     main = "Scatterplot of residuals at lag 1",
     xlab = "Residuals for years 1851–2010",
     ylab = "Residuals for years 1850–2009")

abline(lm(mod.raw$res[2:161] ~ mod.raw$res[1:160]))  # Add regression line

cor(mod.raw$res[1:160], mod.raw$res[2:161]) # This correlation is known as r1
```

## [Why Is Serial Correlation a Problem?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
Failing to properly characterize the correlation structure can affect our estimate of variability in the process.

- Consequences include:
  - Incorrect standard errors
  - Unreliable P-values
  - Potentially misleading statistical inference
  
## Positive Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
- When observations are positively correlated, consecutive observations tend to attract each other.

- As a result, we may observe less variability in the response, simply due to the correlation structure.

- When serial correlation is positive the usual standard error is too small, thus it needs to be inflated

## Positive Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Generate 100 observations with first-order serial correlation 0.9
ts.sim1 <- arima.sim(list(order = c(1,0,0), ar = 0.9), n = 100)

plot(c(1:100), ts.sim1,
     col = "blue", pch = 20, type = "b",
     main = "Positive serial correlation",
     xlab = "Time", ylab = "Y")
```

## Negative Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
- When observations are negatively correlated, consecutive observations tend to repel each other.

- This leads to greater variability in the response, again due to the correlation structure.

- When serial correlation is negative the usual standard error is too large, thus it needs to be reduced

## Negative Serial Correlation {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
# Generate 100 observations with first-order serial correlation -0.9
ts.sim2 <- arima.sim(list(order = c(1,0,0), ar = -0.9), n = 100)

plot(c(1:100), ts.sim2,
     col = "blue", pch = 20, type = "b",
     main = "Negative serial correlation",
     xlab = "Time", ylab = "Y")
```

## The Autoregressive Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Because our observations are measured at **equally spaced points in time** (one observation per year in this example), we denote the series as:

$$
\{Y_t\}
$$

Using this notation:

- $Y_1$ corresponds to the observation at time $t=1$ (year **1850**).
- $Y_2$ corresponds to the observation at time $t=2$ (year **1851**).
- And so on.

## [First-Order Autoregressive Model (AR(1))]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}

Suppose that:

- We have a time series $\{Y_t\}$ observed at equally spaced times.
- The **long-run mean** of the series is $\nu$.

Then the **first-order autoregressive model**, AR(1), is defined as

$$
\mu\{(Y_t-\nu)\mid Y_1,Y_2,\dots,Y_{t-1}\}=\alpha(Y_{t-1}-\nu)
$$

In words:

> The mean of the deviation $Y_t-\nu$ depends on the previous observation $Y_{t-1}$.

---

## [First-Order Autoregressive Model (AR(1))]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
If we are to assume that the data follow an AR(1) model:

:::fragment
$$
\mathbb{E}[Y_t - \nu \mid Y_1, \dots, Y_{t-1}] = \alpha (Y_{t-1} - \nu)
$$
:::

* The first serial correlation coefficient, $\alpha$, is also called the **lag-1 autocorrelation**.
  * $\alpha$ is also known as the population first serial correlation coefficient thus it may be estimated by the sample first serial correlation coefficient denoted $\hat{\alpha} = r_{1}$.  
* It measures the correlation between $Y_t$ and itself at a separation of one time step.
* The pooled estimate of $\alpha$ is then given by:

  $$
  r_{1} = \hat{\alpha} = \frac{\sum_{t=2}^{T} (Y_t - \bar{Y}) (Y_{t-1} - \bar{Y})}{\sum_{t=1}^{T} (Y_t - \bar{Y})^2}.
  $$
  * We don't need to do any of these calculations by hand though since we have already done `cor(mod.raw$res[1:160], mod.raw$res[2:161])` = `r cor(mod.raw$res[1:160], mod.raw$res[2:161])` giving us $r_{1}$

## Why Is This Model Useful? {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}

- If the **AR(1)** model adequately describes the correlation structure, we can **filter the series** to obtain a new regression model where the **errors are independent**.
- This allows us to:
  - Preserve the **same regression coefficients**
  - Remove **serial correlation from the residuals**, "fixing" our standard errors

---

## Filtering the Series {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}

Consider the transformations:

$$
V_t = Y_t - \alpha Y_{t-1}
$$

$$
U_t = X_t - \alpha X_{t-1}
$$

After this transformation we obtain

$$
\mu\{V_t \mid U_t\} = \gamma_0 + \beta_1 U_t
$$

which is a **linear regression model with the same slope** as the original model, but with **independent errors**.

---

## Extension to Multiple Regression {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}

This idea extends naturally to **multiple linear regression**.

We estimate the serial correlation coefficient $\alpha$ using the correlation between consecutive residuals.

From the previous analysis we obtained:

$$
r_1 = 0.5943
$$

The filtered variables become:

$$
V_t = Y_t - r_1 Y_{t-1}
$$

$$
U_{1t} = Year_t - r_1 Year_{t-1}
$$

$$
U_{2t} = Year_t^2 - r_1 Year_{t-1}^2
$$

---

## Filtering the Variables in R {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
r1 <- cor(mod.raw$res[1:160], mod.raw$res[2:161]) # Same as before
```

## Filtering the Variables in R {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
r1 <- cor(mod.raw$res[1:160], mod.raw$res[2:161]) # Same as before

# Create new squared variable
case1502$Year2 <- case1502$Year^2

# Sample size
n <- nrow(case1502) # 161
```

## Filtering the Variables in R {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Calculating "new" / filtered response
v  <- case1502$Temperature[2:n] - r1 * case1502$Temperature[1:(n - 1)]

# Calculating "new" / filtered explanatory variables
u1 <- case1502$Year[2:n] - r1 * case1502$Year[1:(n - 1)]    # Filtered Year
u2 <- case1502$Year2[2:n] - r1 * case1502$Year2[1:(n - 1)]  # Filtered Year^2
```

## Fit the Filtered Model {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
We now fit a regression model using the filtered variables.
```{r}
#| echo: TRUE
mod.filt <- lm(v ~ u1 + u2)
summary(mod.filt)
```

## Residuals After Filtering {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}

We again examine the correlation between lagged residuals.

```{r}
#| echo: TRUE
#| output-location: slide
plot(mod.filt$res[1:159], mod.filt$res[2:160],
     main = "Scatterplot of residuals at lag 1",
     xlab = "Residuals for years 1851–2010",
     ylab = "Residuals for years 1850–2009")

abline(lm(mod.filt$res[2:160] ~ mod.filt$res[1:159]))
```

## Residuals After Filtering {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
cor(mod.filt$res[1:159], mod.filt$res[2:160]) # Essentially no correlation now!
```

## Residuals After Filtering{auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: slide
plot(case1502$Year[2:161], mod.filt$res,
  col = "black",
  pch = 20,
  type = "b",
  ylab = "Residuals",
  xlab = "Years"
)

abline(h = 0, lty = "dashed")
```


