---
title: "(ST 551) Lab 5 <br> Test Galore!"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
    code-line-numbers: TRUE
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

<style>
.slightlysmaller {
  font-size: 0.85em;   /* Slightly smaller than normal */
}

.evensmaller {
  font-size: 0.666em;   /* Slightly smaller than normal */
}

.bigger {
  font-size: 1.15em;   /* Slightly bigger than normal */
}

.biggest {
  font-size: 1.3em;    /* Larger, if needed */
}

.footnote {
  position: absolute;
  bottom: 1em;
  font-size: 0.8em;
  color: #555;
}
</style>

## Yo!
* No lab next week (midterm)
* Questions, Comments, Concerns?

## [Wilcoxon Signed-Rank Test Procedure (Quick Review)]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
To test that $c_{0}$ is the population *center*:

- **Compute distances from $c_{0}$**

::: {.fragment}
$$
  N_i = |Y_i - c_{0}|
$$
:::

- **Rank the distances (absolute deviations)**

::: {.fragment}
$$
   R_i = \text{Rank}(N_i) = \sum_{j=1}^{n} \mathbb{I}\{N_{i} \ge N_{j}\}
$$
:::

- **Test statistic**: sum of ranks corresponding to observations greater than $c_{0}$ 

::: {.fragment}
$$
   S = \sum_{i=1}^{n} R_i \mathbb{I}\{Y_i > c_{0}\}
$$
:::

## [Reference Distribution - Exact p-value Method]{.r-fit-text} {.slightlysmaller}
- Assume each rank is equally likely to be assigned to a value above or below $c_{0}$.  
- If the population is symmetric about $c_{0}$:
  - Each rank $1,2,\ldots,n$ independently has probability $0.5$ of being assigned to an observation above $c_{0}$.
- Consider all possible assignments of ranks above/below $c_{0}$ to derive the exact distribution of $S$.

---

## [Reference Distribution - Normal Approximation]{.r-fit-text} {.slightlysmaller}
If the population distribution is symmetric about $c_{0}$, then:

- The test statistic $S$ is approximately normally distributed:

::: {.fragment}
$$
  S \;\;\dot\sim\;\; \text{Normal}\left(\frac{n(n+1)}{4}, \frac{n(n+1)(2n+1)}{24}\right)
$$
:::

- Construct a $Z$-statistic and compare to $N(0,1)$:

::: {.fragment}
$$
  Z = \frac{S - \frac{n(n+1)}{4}}{\sqrt{\frac{n(n+1)(2n+1)}{24}}} \;\;\dot\sim\;\; \text{Normal}(0,1)
$$
:::
  
## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* Setting the scene for the Wilcoxon Signed-Rank test

::: {.fragment}
$$
n = 20, \qquad Y_{1}, \ldots, Y_{n} \overset{\text{iid}}{\sim} \chi^{2}_{(3)}
$$
:::

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 20
sampA <- rchisq(n, 3)
```
:::

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
rslt1 <- wilcox.test(sampA, mu = 3)
rslt1
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
rslt2 <- wilcox.test(sampA, mu = 3, exact = FALSE, correct = FALSE)
rslt2
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* What is the difference between the two code chunks?
  * (1) By default (if `exact` is not specified), an exact p-value is computed if the samples contain less than 50 finite values and there are no ties. Otherwise, a normal approximation is used
  * (2) Normal approximation without a continuity correction

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# What are the different components (elements) of this object (list)
names(rslt2)
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can confirm the 2nd result `by hand'
mu0 <- 3                                  # Null hypothesis value
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can confirm the 2nd result `by hand'
mu0 <- 3                                  # Null hypothesis value
sampA.dists <- abs(sampA - mu0)           # Compute distances
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can confirm the 2nd result `by hand'
mu0 <- 3                                  # Null hypothesis value
sampA.dists <- abs(sampA - mu0)           # Compute distances
sampA.ranks <- rank(sampA.dists)          # Rank observation distances
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can confirm the 2nd result `by hand'
mu0 <- 3                                  # Null hypothesis value
sampA.dists <- abs(sampA - mu0)           # Compute distances
sampA.ranks <- rank(sampA.dists)          # Rank observation distances

# Slap them together in a matrix to view it all
cbind(sampA, sampA.dists, sampA.ranks)    
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can confirm the 2nd result `by hand'
mu0 <- 3                                  # Null hypothesis value
sampA.dists <- abs(sampA - mu0)           # Compute distances
sampA.ranks <- rank(sampA.dists)          # Rank observation distances

# Calculate our S test statistic
# Summing up all the ranks that corresponds to observations larger than null
S.stat <- sum(sampA.ranks[sampA > mu0])
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can confirm the 2nd result `by hand'
mu0 <- 3                                  # Null hypothesis value
sampA.dists <- abs(sampA - mu0)           # Compute distances
sampA.ranks <- rank(sampA.dists)          # Rank observation distances

# Calculate our S test statistic
# Summing up all the ranks that corresponds to observations larger than null
S.stat <- sum(sampA.ranks[sampA > mu0])
S.stat
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Calculate our S test statistic
# Summing up all the ranks that corresponds to observations larger than null
S.stat <- sum(sampA.ranks[sampA > mu0])   
S.stat

mean.S <- n * (n + 1) / 4                 # Mean of S
var.S <- n * (n + 1) * (2 * n + 1) / 24   # Variance of S
```

## Wilcoxon Signed-Rank Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Calculate our S test statistic
# Summing up all the ranks that corresponds to observations larger than null
S.stat <- sum(sampA.ranks[sampA > mu0])   
S.stat

mean.S <- n * (n + 1) / 4                 # Mean of S
var.S <- n * (n + 1) * (2 * n + 1) / 24   # Variance of S
Z.stat <- (S.stat - mean.S) / sqrt(var.S) # Z-stat
pval <- 2 * (1 - pnorm(abs(Z.stat)))      # p-value
pval
```


## Comparing With R {auto-animate="true" auto-animate-easing="ease-in-out"}

::: columns
::: column
**Using R**
```{r}
#| echo: TRUE
#| eval: TRUE
rslt2[["p.value"]]
```
:::

::: column
**"By Hand"**
```{r}
#| echo: TRUE
#| eval: TRUE
pval
```
:::
:::

## [t-Test on Squared Deviations (Testing Variance)]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Suppose, we are interesting in testing the population variance, that is:

$$
 H_{0}: \sigma^{2} = \sigma^{2}_{0}
$$

We have two main options:
  
1. Perform the Chi-squared test (assumes population is Normally distributed)
  
::: {.fragment}
$$
X(\sigma_0) = \dfrac{(n-1)s^2}{\sigma_{0}^2} \sim \chi^{2}_{(n-1)}
$$
:::

2. $t$-test on squared deviations from mean: $W_{i} = \left(X_{i} - \bar{X}\right)^{2}$

::: {.fragment}
$$
\frac{s^2-\sigma^2}{\sqrt{\operatorname{Var}\left(s^2\right)}} = \frac{\frac{n-1}{n}s^2 - \frac{n-1}{n}\sigma^2}{\sqrt{\operatorname{Var}(\frac{n-1}{n}s^2)}} = \frac{\bar{W} - \frac{n-1}{n}\sigma^2}{\sqrt{\operatorname{Var}(\bar{W})}}\overset{d}{\to} N(0, 1)
$$
:::

  - Therefore, we can perform a test of $H_{0}: \sigma^2 = \sigma^2_0$ using a t-test to test that the population mean of the $W_{i}$ is $\frac{n-1}{n}\sigma_0$

::: {.fragment}
$$
\quad\Rightarrow\quad t(\sigma_{0}^{2}) = \frac{\bar{W} - \frac{n-1}{n}\sigma^2_{0}}{\sqrt{\frac{s^2_W}{n}}}\overset{d}{\to} N(0, 1)
$$
:::

## [t-Test on Squared Deviations (Testing Variance)]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Suppose:

::: {.fragment}
$$
n = 10, \qquad X_{i} \overset{\text{iid}}{\sim} \text{Uniform}(0, 1)
$$
:::

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 10
newSamp <- runif(n)
```
:::

## [t-Test on Squared Deviations (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
newSamp
```

## [t-Test on Squared Deviations (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
wVals <- (newSamp - mean(newSamp))^2
wVals
```

## [t-Test on Squared Deviations (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
wVals <- (newSamp - mean(newSamp))^2
tStat <- (mean(wVals) - (n - 1) / n * 1 / 12) / sqrt(var(wVals) / n)
tStat
```

## [t-Test on Squared Deviations (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
wVals <- (newSamp - mean(newSamp))^2
tStat <- (mean(wVals) - (n - 1) / n * 1 / 12) / sqrt(var(wVals) / n)
pVal <- 2 * (1 - pt(abs(tStat), n - 1))
pVal
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {.slightlysmaller auto-animate="true" auto-animate-easing="ease-in-out"}
* As we have done many times before, we can compare the performance of these two tests

::: {.fragment}
$$
n = 50, \qquad X_{i} \overset{\text{iid}}{\sim} \text{Uniform}(0, 1), \qquad H_{0}: \sigma^{2} = \frac{1}{12}
$$
:::

* Since this is the true variance, so we are hoping that 5% of the resulting p-values will be less than 0.05
  * That is, we want our tests to be exact, $p$-values having a Uniform(0, 1) distribution
    * Note that it's only a coincidence that our $X_{i}$ and $p$-values share this distribution, our $X_{i}$ will often not be Uniform but our $p$-values will always be Uniform(0, 1) (if the test is exact)

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nsim <- 10000
n <- 50
sig0sq <- 1 / 12
pVals.t <- pVals.chi <- rep(0, nsim)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nsim <- 10000
n <- 50
sig0sq <- 1 / 12
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- runif(n)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nsim <- 10000
n <- 50
sig0sq <- 1 / 12
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- runif(n)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
n <- 50
sig0sq <- 1 / 12
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- runif(n)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2"
nsim <- 10000
n <- 10
sig0sq <- 1 / 12
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- runif(n)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
n <- 10
sig0sq <- 1 / 12
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- runif(n)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
Suppose now...

::: {.fragment}
$$
n = 50, \qquad X_{i} \overset{\text{iid}}{\sim} \chi^{2}_{(10)}, \qquad H_{0}: \sigma^{2} = 20
$$
:::


## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nsim <- 10000
n <- 10
sig0sq <- 20   # Updated to reflect null
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- rchisq(n, df = 10)   # Generating different data
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
n <- 10
sig0sq <- 20 # Updated to reflect null
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- rchisq(n, df = 10) # Generating different data
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2"
nsim <- 10000
n <- 50
sig0sq <- 20
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- rchisq(n, df = 10)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## [t-test (GOAT) vs Chi-Squared Test (Testing Variance)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
n <- 50
sig0sq <- 20
pVals.t <- pVals.chi <- rep(0, nsim)
for (i in 1:nsim) {
  newSamp <- rchisq(n, df = 10)
  chisq.stat <- (n - 1) * var(newSamp) / sig0sq
  wVals <- (newSamp - mean(newSamp))^2
  t.stat <- (mean(wVals) - (n - 1) / n * sig0sq) / sqrt(var(wVals) / n)
  pVals.chi[i] <- 1 - pchisq(chisq.stat, n - 1)
  pVals.t[i] <- 1 - pt(t.stat, n - 1)
}
mean(pVals.chi < 0.05)
mean(pVals.t < 0.05)
```

## Kolmogorov–Smirnov Test {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Testing some some population distribution $F$

- Hypothesis:

::: {.fragment}
$$
H_{0}: F = F_{0}
$$
:::

- Test statistic

::: {.fragment}
$$
D(G_{0}) = \sup_{x}\left|\hat{F}(x) - F_{0}(x)\right|
$$
:::

  - Where $\hat{F}(x)$ is the empirical cdf

::: {.fragment}
$$
\hat{F}(x) = \frac{1}{n}\sum_{i=1}^{n}\mathbb{I}\{X_{i} \le x\}
$$
:::

## Kolmogorov–Smirnov Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* Suppose

::: {.fragment}
$$
n = 20, \qquad X_{i} \overset{\text{iid}}{\sim} N(0, 1)
$$
:::

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 20
sampB <- rnorm(n, 0, 1)
```
:::


## Kolmogorov–Smirnov Test {.slightlysmaller auto-animate="true" auto-animate-easing="ease-in-out"}
* The function `ks.test()` performs the Kolmogorov–Smirnov test. 
* You give the function a data set and a hypothesized distribution function (e.g., `pnorm`, `punif`, or `pchisq`) and also the hypothesized parameters of the distribution function.
* Here we draw a sample from the $\text{Normal}(0,1)$ population distribution, and then test that the data come from a $\text{Uniform}(0,1)$ distribution.

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
ks.test(sampB, punif, 0, 1)
```
:::

## Kolmogorov–Smirnov Test {auto-animate="true" auto-animate-easing="ease-in-out"}
We could instead test the true null hypothesis that the data come from a Normal$(0, 1)$ distribution:

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
ks.test(sampB, pnorm, 0, 1)
```
:::

## Kolmogorov–Smirnov Test {auto-animate="true" auto-animate-easing="ease-in-out"}
Or we could instead test that the data come from a $t_{(5)}$ distribution:

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
ks.test(sampB, pt, 5)
```
:::