---
title: "(ST 551) Lab 4 <br> Binomial and Sign Tests"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: concave
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## Welcome Back Kiddos!
* Back Again... Woohoo...
* Questions, Comments, Concerns?

## First Some Grilling!
* Are you guys asking questions in class?
* Are you guys answering questions in class?

## Warning!
* You guys need to start answering questions in class more
  * *Dramatic pause*
* If not, Sarah might snap
  * She's already showing symptoms
    * She's concerned that she her perfection is intimidating -- her exact words were: "I don't want to challenge the Gods"

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* The following ramblings are related to your Homework 3 - Question 1 (different numbers)
* Suppose we are in a situation such that

::: {.fragment}
$$
H_{0}: p_{0} = 0.30, \quad\text{and}\quad x = 1, \quad n = 8, \quad\Rightarrow\quad \hat{p} = \frac{x}{n} = \frac{1}{8}
$$
:::

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Two-sided by default
binom.test(1, 8, 0.5)
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Lower alternative
binom.test(1, 8, 0.3, alt = "less")
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Upper alternative
binom.test(1, 8, 0.3, alt = "greater")
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Let's say we run a two-sided test and we want to extract some results
names(binom.test(1, 8, 0.3))
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We could pull out the p-value
binom.test(1, 8, 0.3)$p.val
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We could pull out the p-value
binom.test(1, 8, 0.3)$p.val

# The confidence interval
binom.test(1, 8, 0.3)$conf.int
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We could pull out the p-value
binom.test(1, 8, 0.3)$p.val

# The confidence interval
binom.test(1, 8, 0.3)$conf.int

# The test statistic
binom.test(1, 8, 0.3)$statistic
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Notice that R allows for partial tagging with the $ operator
binom.test(1, 8, 0.3)$stat
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Notice that R allows for partial tagging with the $ operator
binom.test(1, 8, 0.3)$stat
binom.test(1, 8, 0.3)$s
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Notice that R allows for partial tagging with the $ operator
binom.test(1, 8, 0.3)$stat
binom.test(1, 8, 0.3)$s
binom.test(1, 8, 0.3)$p
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# So I could just type this to get our p-values?
binom.test(1, 8, 0.3)$p
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Well, partial such that it's not ambiguous
binom.test(1, 8, 0.3)$pa # parameter
binom.test(1, 8, 0.3)$p. # p.value
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Well, partial such that it's not ambiguous
binom.test(1, 8, 0.3)$pa # parameter
binom.test(1, 8, 0.3)$p. # p.value
# We've got two components that start with p so R doesn't know which one to get
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
In class you guys talked about:

- Under $H_0: p = p_0$, $X \sim \text{Binomial}(n, p_0)$ for a two-sided alternative $H_A: p \neq p_0$
- We reject $H_0$ for values of $X$ far from the expected value under $H_0$?
- **Rejection Region (distance-based):**  
$$
|X - n p_0| > c
$$

  - Values of $X$ least probable under $H_0$?  

- **Rejection Region (probability-based):**  
$$
p_0(X) \le c
$$ 
  - where $p_0(X)$ is the pmf of a $\text{Binomial}(n, p_0)$ distribution, evaluated at $X$.

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### We can check R is doing what was discussed in class to compute p-values
# Our probability mass function (under the null) at each x
dbinom(0:8, 8, 0.3)
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### We can check R is doing what was discussed in class to compute p-values
# Our probability mass function (under the null) at each x
dbinom(0:8, 8, 0.3)

# Checking which probabilities under the null are less than or equal to
# the probabilities when we observe 1 success (like our example)
which(dbinom(0:8, 8, 0.3) <= dbinom(1, 8, 0.3))
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### We can check R is doing what was discussed in class to compute p-values
# Our probability mass function (under the null) at each x
dbinom(0:8, 8, 0.3)

# Checking which probabilities under the null are less than or equal to
# the probabilities when we observe 1 success (like our example)
which(dbinom(0:8, 8, 0.3) <= dbinom(1, 8, 0.3))

# Using this vector to index (mask) & Summing up the null probabilities 
p.two <- sum(dbinom(0:8, 8, 0.3)[dbinom(0:8, 8, 0.3) <= dbinom(1, 8, 0.3)])
p.two
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### One-sided p-values are a more straightforward
# Summing up the null probabilities less than / equal to x (= 1)
p.lo <- sum(dbinom(0:1, 8, 0.3))
p.lo
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### One-sided p-values are a more straightforward
# Summing up the null probabilities greater than / equal to x (= 1)
p.up <- sum(dbinom(1:8, 8, 0.3))
p.up
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Notice pbinom() is doing this for us, for example
# Lower alternative
pbinom(4, 12, 0.7)         # Lower tail by default
sum(dbinom(0:4, 12, 0.7))
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Notice pbinom() is doing this for us, for example
# Upper alternative
1 - pbinom(7, 12, 0.7)
sum(dbinom(8:12, 12, 0.7))
```

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* Lastly, something that might spark some ideas for Homework 3 - Question 1 (f)
* Notice that the confidence interval remains the same regardless of which null hypothesis we test, because it depends only on the observed data---not on the hypothesized value of $p$.

## Exact Binomial Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
binom.test(1, 8, 0.25)
binom.test(1, 8, 0.50)
binom.test(1, 8, 0.75)
```


## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| warning: TRUE
### If we want some Normal approximation action
prop.test(1, 8, 0.3, correct = FALSE)
```

## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
The test gives back a $\chi^{2}_{(1)}$ test statistic, which is the same thing as a $z$-stat squared, that is:
$$
\chi^{2} = \left(\frac{\hat{p} - p_{0}}{\sqrt{\frac{p_{0}(1 - p_{0})}{n}}}\right)^{2} = Z^{2}
$$

* The warning we get is just telling us our sample size might not be large enough to justify a Normal approximation
* We can verify R is doing the above:

## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Our setting
n <- 8
x <- 1
p0 <- 0.3
pHat <- x / n
pHat
```

## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Calculate z-stat
n <- 8
x <- 1
p0 <- 0.3
pHat <- x / n
zStat <- (pHat - p0) / sqrt(p0 * (1 - p0) / n)
zStat
pVal <- 2 * (1 - pnorm(abs(zStat)))
```

## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Calculate p-value
zStat <- (pHat - p0) / sqrt(p0 * (1 - p0) / n)
pVal <- 2 * (1 - pnorm(abs(zStat)))
pVal
```

## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Checking the z-statistic squared is the same as the chi-squared stat
zStat^2
prop.test(x, n, p0, correct = FALSE)[["statistic"]]
```

## Approximate z-test of Proportions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We see the warning is thrown when our expected counts are < 5
prop.test
```


## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* For the Sign test... we just use some other tests...
  * Specifically we can just use `binom.test()` or `prop.test()`
* We can try testings $H_{0}: M = 0$ on $\mathbf{X} \sim t_{(3)}$


## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Generate Data
n <- 15
newSamp <- rt(n, 3)
newSamp
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
* In practice (aka comps) we just use a z-test of proportions...
* The test statistic for the Sign Test, equal to the Binomial test with $H_{0}: p_{0} = 0.50$

::: {.fragment}
$$
  Z = \frac{\hat{p}_{M_{0}} - p_{0}}{\sqrt{\frac{p_{0}(1-p_{0})}{n}}}
$$
:::

* where $\hat{p}_{M_{0}} = \frac{x}{n}$, and $x$ = number of observations $< M_{0}$ (median under null).

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Find number of observations less than null hypothesis median
x <- sum(newSamp < 0)
x
```

## Sign Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Find number of observations less than null hypothesis median
x <- sum(newSamp < 0)

# Run our tests
binom.test(x, n, 0.5)
prop.test(x, n, 0.5, correct = FALSE)
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Classic simulation time
nSim <- 10000
n <- 15
pvals.t <- pvals.sign <- rep(0, nSim)
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nSim <- 10000
n <- 15
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3)
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
}
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nSim <- 10000
n <- 15
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3)
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
  pvals.t[i] <- 2 * (1 - pt(abs(tStat), n - 1))
  pvals.sign[i] <- binom.test(x, n, 0.5)[["p.value"]]
}
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nSim <- 10000
n <- 15
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3)
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
  pvals.t[i] <- 2 * (1 - pt(abs(tStat), n - 1))
  pvals.sign[i] <- binom.test(x, n, 0.5)[["p.value"]]
}

mean(pvals.t < 0.05)
mean(pvals.sign < 0.05)
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: fragment
# Our p-values should be Unif(0, 1) if exact
hist(pvals.t, 
     breaks = seq(0, 1, 0.05),
     col = "lightgreen",
     border = "black",
     main = "Histogram of p-values from t-test",
     xlab = "p-values")
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: fragment
# Our p-values should be Unif(0, 1) if exact
hist(pvals.sign, 
     breaks = seq(0, 1, 0.05),
     col = "lightpink",
     border = "black",
     main = "Histogram of p-values from Sign test",
     xlab = "p-values")
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
* We could instead generate:

::: {.fragment}
$$
\mathbf{X} \sim t_{(3)} + 1
$$
:::

* Meaning we just shift our data up by one (so our population mean and median are now 1)
* But we are still testing $H_{0}: M = 0$
* Since the null hypothesis is false, we can evaluate the power of these two tests now

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "6"
# New simulation set-up (basically the same)
nSim <- 10000
n <- 15
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3) + 1
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
  pvals.t[i] <- 2 * (1 - pt(abs(tStat), n - 1))
  pvals.sign[i] <- binom.test(x, n, 0.5)[["p.value"]]
}
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# New simulation set-up (basically the same)
nSim <- 10000
n <- 15
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3) + 1
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
  pvals.t[i] <- 2 * (1 - pt(abs(tStat), n - 1))
  pvals.sign[i] <- binom.test(x, n, 0.5)[["p.value"]]
}

mean(pvals.t < 0.05)
mean(pvals.sign < 0.05)
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "3"
# New simulation set-up (basically the same)
nSim <- 10000
n <- 50
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3) + 1
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
  pvals.t[i] <- 2 * (1 - pt(abs(tStat), n - 1))
  pvals.sign[i] <- binom.test(x, n, 0.5)[["p.value"]]
}
```

## t-test (GOAT) vs Sign test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# New simulation set-up (basically the same)
nSim <- 10000
n <- 50
pvals.t <- pvals.sign <- rep(0, nSim)
for (i in 1:nSim) {
  newSamp <- rt(n, 3) + 1
  tStat <- mean(newSamp) / sqrt(var(newSamp) / n)
  x <- sum(newSamp < 0)
  pvals.t[i] <- 2 * (1 - pt(abs(tStat), n - 1))
  pvals.sign[i] <- binom.test(x, n, 0.5)[["p.value"]]
}

mean(pvals.t < 0.05)
mean(pvals.sign < 0.05)
```

## [Homework 3 - Question 4 - Tech Support]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
**Problem:** 

* What is the power of the sign test for testing that the population median is $0$ ($H_0: m = 0$) if the true population distribution is $\text{Normal}(1,1)$,  for a sample of size $n = 30$? 
  * Answer this using theoretical arguments and the normal cdf, and then perform simulations to confirm your answer.

**Tips:**

* For the simulation portion, I'd just follow the hints and you'll get pretty much all the way there:
  * Just a few lines in the previous `for` loop templates shown today need to be changed
* For the theoretical portion...
  * Essentially you want to get your rejection region in terms of $\hat{p}$ then to calculate those probabilities you can use the CLT, we know:

::: {.fragment}
$$
\left|\frac{\hat{p}-p}{\sqrt{\frac{{p(1-p)}}{n}}}\right| \overset{\cdot}{\sim} N(0,1)
$$
:::

* and you can find $p = P(X > 0)$ since we know the distribution of $X$ is Normal(1, 1)