---
title: "(ST 551) Lab 10 <br> Tons of Totally Tubular Tests"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
    code-line-numbers: TRUE
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

<style>
.slightlysmaller {
  font-size: 0.85em;   /* Slightly smaller than normal */
}

.evensmaller {
  font-size: 0.666em;   /* Slightly smaller than normal */
}

.bigger {
  font-size: 1.15em;   /* Slightly bigger than normal */
}

.biggest {
  font-size: 1.3em;    /* Larger, if needed */
}

.footnote {
  position: absolute;
  bottom: 1em;
  font-size: 0.8em;
  color: #555;
}
</style>

## Last Lab of Term!
* Questions, Comments, Concerns?
* Today's Tests
  * Mood's Test
  * Levene's Test
  * Two-Sample Kolmogorov–Smirnov Test
  * Permutation Test
* Let's get cookin'

## Mood’s Test: Setting

Setting:
$$
\begin{align*}
X_{1}, X_{2}, \ldots, X_{n_X} &\overset{iid}{\sim} F_X \\[0.25cm]
Y_{1}, Y_{2}, \ldots, Y_{n_Y} &\overset{iid}{\sim} F_Y
\end{align*}
$$

::: {.fragment}
Where $F_X$ and $F_Y$ are two (possibly different) arbitrary population distributions.
:::


## Mood’s Test: Parameters of Interest {.smaller}

- Population medians $m_X$ and $m_Y$

::: {.fragment}
$$
m_X = F_X^{-1}\!\left(\frac{1}{2}\right), 
\qquad 
m_Y = F_Y^{-1}\!\left(\frac{1}{2}\right)
$$
:::

::: {.fragment}
These satisfy:  
$$
P(X \le m_X) = P(X \ge m_X), \qquad 
P(Y \le m_Y) = P(Y \ge m_Y)
$$
:::


## Mood’s Test: Null Hypothesis

::: {.fragment}
$$
H_0: m_X = m_Y
$$
:::


## Mood’s Test: Estimator {.smaller}

If $m_X = m_Y = m$ (as is the case under $H_{0}$), our best estimate of the common median is the **combined sample median**:

::: {.fragment}
$$
\hat{m} = \text{median}\{X_1, \ldots, X_{n_X}, Y_1, \ldots, Y_{n_Y}\}
$$
:::


## [Sampling Distribution of the Sample Median]{.r-fit-text} {.smaller}
For iid data $X_1, \ldots, X_n$ with density $f(x)$, the sample median, $\hat{m}$, is asymptotically normal:

::: {.fragment}
$$
\hat{m} \;\dot\sim\; N\!\left(m,\, \frac{1}{4n \cdot \left[f(m)\right]^2}\right)
$$
:::


## Mood’s Test Procedure {.smaller}

1) Find combined sample median $\hat{m}$
2) Test that the true population proportion of $X$ observations greater than $\hat{m}$ is equal to the true population proportion of $Y$ observations greater than $\hat{m}$

- It's often helpful (when you're performing this test yourself) to create a table similar to the following:

::: {.fragment}
$$
\begin{array}{|c|c|c|c|}
\hline
 & \# > \hat{m} & \# \le \hat{m} & \text{Total} \\[0.25cm]
\hline
X \text{ Sample} & s_X & n_X - s_X & n_X \\[0.25cm]
\hline
Y \text{ Sample} & s_Y & n_Y - s_Y & n_Y \\[0.25cm]
\hline
\end{array}
$$
:::

::: {.fragment}
This table summarizes the data so we can perform any test of association, independence, or homogeneity:
:::

## Mood’s Test Procedure {.smaller}
Since we have that table we could perform the usual contingency table tests:

- Fisher’s exact test
- Log-Odds Ratio
- Pearson’s Chi-Squared

::: {.fragment}
Though often it's easiest to just use the two-sample binomial proportions **z-test** (equivalent to Pearson chi-squared) as alluded to before
:::

## Mood’s Test Procedure {.evensmaller}
From our table:
$$
\begin{array}{|c|c|c|c|}
\hline
 & \# > \hat{m} & \# \le \hat{m} & \text{Total} \\[0.25cm]
\hline
X \text{ Sample} & s_X & n_X - s_X & n_X \\[0.25cm]
\hline
Y \text{ Sample} & s_Y & n_Y - s_Y & n_Y \\[0.25cm]
\hline
\end{array}
$$

::: {.fragment}
The $z$-test statistic is:
$$
z = 
\frac{\hat{p}_X - \hat{p}_Y}
{\sqrt{\hat{p}_c(1-\hat{p}_c)\left(\frac{1}{n_X} + \frac{1}{n_Y}\right)}}
$$
where
:::

- $\hat{p}_{X} = \frac{s_{X}}{n_{X}}, \qquad \hat{p}_{Y} = \frac{s_{Y}}{n_{Y}}$
- $\hat{p}_{C} = \frac{s_{X} + s_{Y}}{n_{X} + n_{Y}} = \frac{s_{X} + s_{Y}}{N}$

## Mood’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Load library needed for dataset
library(Sleuth3)
```

## Mood’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Separate Group A and Group B:
groupA <- ex0125$Zinc[ex0125$Group == 'A']
groupB <- ex0125$Zinc[ex0125$Group == 'B']
```

## Mood’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Compute the combined median
medComb <- median(ex0125[["Zinc"]])
```

## Mood’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Construct the 2x2 table counting observations above/below the median:
tabZinc <- rbind(c(sum(groupA > medComb), sum(groupA <= medComb)), 
                 c(sum(groupB > medComb), sum(groupB <= medComb)))
tabZinc
```

## Mood’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# With our table we can then just use a Chi-squared test:
chisq.test(tabZinc)
```

## Levene’s Test Setting
Setting:
$$
\begin{align*}
\text{Population 1: } & X_{1}, X_{2}, \ldots, X_{m} \\[0.25cm]
\text{Population 2: } & Y_{1}, Y_{2}, \ldots, Y_{n}
\end{align*}
$$

::: {.fragment}
We're interested in testing:
$$
\begin{align*}
H_{0}: \sigma^{2}_{X} = \sigma^{2}_{Y}
\end{align*}
$$
:::

## Levene’s Test Implementations {.evensmaller}
The variations of this test arise from the different ways we construct new variables that measure how far each observation is from a chosen measure of central tendency.

- **Step 1:** Choose how we are going to measure distance:
  - Option A: Absolute difference from sample median

::: {.fragment}
$$
U_i = |X_i - \text{med}(X)| \qquad 
V_i = |Y_i - \text{med}(Y)|
$$
:::

  - Option B: Squared difference from sample median

::: {.fragment}
$$
U_i = (X_i - \text{med}(X))^2 \qquad 
V_i = (Y_i - \text{med}(Y))^2
$$
:::


  - Option C: Absolute difference from sample mean

::: {.fragment}
$$
U_i = |X_i - \bar{X}| \qquad 
V_i = |Y_i - \bar{Y}|
$$
:::

  - Option D: Squared difference from sample mean

::: {.fragment}
$$
U_i = (X_i - \bar{X})^2 \qquad 
V_i = (Y_i - \bar{Y})^2
$$
:::

- **Step 2:** Perform a two-sample $t$-test to test the hypothesis that the population mean of the $U_{i}$ is the same as the population mean of $V_{i}$, that is, we're testing:

::: {.fragment}
$$
H_{0}: \mu_{U} = \mu_{V}
$$
:::

## Levene’s Test Interpretation
- Interpretation depends on how differences are measured.
- **Only Option D** can be interpreted as testing a difference in **population variances**.

## Levene’s Test Assumptions
- Independence within and between samples  
- Reasonably large sample sizes so that t-test assumptions hold for $U_i$ and $V_i$  
- If equal-variance t-test is used, we assume $U_i$ and $V_i$ have equal variances

## Levene’s Test Uses
- When do we use this?
  - To answer direct questions about differences in variance/spread  
  - To decide which two-sample t-test to use (**should not do this**)  
    - Not recommended: low power for detecting variance differences  
    - Prefer **Welch’s t-test** unless certain variances are equal

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Compute the squared deviations of each observation from their respective sample means
uVals <- (groupA - mean(groupA))^2
vVals <- (groupB - mean(groupB))^2
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Perform t-test
t.test(uVals, vVals)
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# R of course has built in functions
library(car)
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Syntax Version 1
leveneTest(y = ex0125$Zinc, group = ex0125$Group)
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Syntax Version 2
leveneTest(Zinc ~ Group, data = ex0125)
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Does absolute median distance by default
uVals <- abs(groupA - median(groupA))
vVals <- abs(groupB - median(groupB))
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We see the same results `by hand'
t.test(uVals, vVals, var.equal = TRUE)
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Absolute mean distance version
leveneTest(Zinc ~ Group, data = ex0125, center = mean)
```

## Levene’s Test Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Absolute mean distance version
uVals <- abs(groupA - mean(groupA))
vVals <- abs(groupB - mean(groupB))
t.test(uVals, vVals, var.equal = TRUE)
```



## [Two-Sample Kolmogorov–Smirnov Test: Setting]{.r-fit-text}

$$
\begin{align*}
X_{1}, X_{2}, \ldots, X_{n_X} &\overset{iid}{\sim} F_X \\[0.25cm]
Y_{1}, Y_{2}, \ldots, Y_{n_Y} &\overset{iid}{\sim} F_Y
\end{align*}
$$

::: {.fragment}
Where $F_X$ and $F_Y$ are two (possibly different) arbitrary population distributions.
:::

## [Two-Sample Kolmogorov–Smirnov Test: Hypotheses]{.r-fit-text}

::: {.fragment}
$$
H_0: F_X = F_Y, \qquad\qquad H_A: F_X \ne F_Y
$$
:::


## KS Test Statistic {.smaller}

::: {.fragment}
The test statistic is the maximum vertical distance between the empirical CDFs:
$$
D = \sup_x \left| \hat{F}_X(x) - \hat{F}_Y(x) \right|
$$
:::

::: {.fragment}
Where the empirical CDFs are:
$$
\hat{F}_X(x) = \frac{1}{n_X}\sum_{i=1}^{n_X} \mathbf{1}\{X_i \le x\},
\qquad
\hat{F}_Y(x) = \frac{1}{n_Y}\sum_{i=1}^{n_Y} \mathbf{1}\{Y_i \le x\}
$$
:::

## [Two-Sample Kolmogorov–Smirnov Test: Example]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# So much code we have to write...
ks.test(groupA, groupB)
```

## Permutation Tests: Setting

$$
\begin{align*}
X_{1}, X_{2}, \ldots, X_{n_X} &\overset{iid}{\sim} F_X \\[0.25cm]
Y_{1}, Y_{2}, \ldots, Y_{n_Y} &\overset{iid}{\sim} F_Y
\end{align*}
$$

::: {.fragment}
Where $F_X$ and $F_Y$ are two (possibly different) arbitrary population distributions.
:::


## Permutation Tests: Basic Idea {.smaller}

(1) **Choose a test statistic** $W$ that measures the type of difference of interest (e.g., difference in sample means, $t$-statistic, ratio of sample variances, difference in medians, etc.)

(2) **Permute group labels** across all observations and recompute the test statistic.

(3) **Repeat many times** to obtain the permutation distribution of $W$.

(4) **Compute the p-value** by comparing the observed statistic to its permutation distribution.

## Permutation Tests: Performance {.smaller}
How does this test perform? Well...

- It depends (my favorite answer!)
  - The effectiveness of a permutation test depends on the test statistic chosen, but the test can be sensitive to differences beyond what the statistic measures. 
  - Permutation tests rely on exchangeability—the assumption that both samples come from identical distributions, not merely distributions with equal test statistics of interest.
  
## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Generate some data
n1 <- 5
n2 <- 10
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Generate some data
n1 <- 5
n2 <- 10

sampA <- rnorm(n1, 0, 1)
sampB <- rnorm(n2, 0, 10)
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Generate some data
n1 <- 5
n2 <- 10

sampA <- rnorm(n1, 0, 1)
sampB <- rnorm(n2, 0, 10)
sampA
sampB
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n1 <- 5
n2 <- 10
sampA <- rnorm(n1, 0, 1)
sampB <- rnorm(n2, 0, 10)

# Calculate observed test statistic
testStat <- mean(sampA) - mean(sampB)
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n1 <- 5
n2 <- 10
sampA <- rnorm(n1, 0, 1)
sampB <- rnorm(n2, 0, 10)

# Calculate observed test statistic
testStat <- mean(sampA) - mean(sampB)
testStat
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Combining the two samples into one vector:
sampComb <- c(sampA, sampB)
sampComb
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Select a random set of n1 indices (positions) to be the new sample A
permAind <- sample(1:(n1 + n2), n1, replace = FALSE)
permAind
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Index to obtain new sample
sampA.perm <- sampComb[permAind]
sampA.perm
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Complement of indices A are sample B
sampB.perm <- sampComb[-permAind]
sampB.perm
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Compute permuted test statistic
permStat <- mean(sampA.perm) - mean(sampB.perm)
permStat
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Do this many times
nPerm <- 1000
permStats <- rep(0, nPerm)
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Do this many times
nPerm <- 1000
permStats <- rep(0, nPerm)
for (j in 1:nPerm) {
  permAind <- sample(1:(n1 + n2), n1, replace = FALSE)
  sampA.perm <- sampComb[permAind]
  sampB.perm <- sampComb[-permAind]
}
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Do this many times
nPerm <- 1000
permStats <- rep(0, nPerm)
for (j in 1:nPerm) {
  permAind <- sample(1:(n1 + n2), n1, replace = FALSE)
  sampA.perm <- sampComb[permAind]
  sampB.perm <- sampComb[-permAind]
  permStats[j] <- mean(sampA.perm) - mean(sampB.perm)
}
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Do this many times
nPerm <- 1000
permStats <- rep(0, nPerm)
for (j in 1:nPerm) {
  permAind <- sample(1:(n1 + n2), n1, replace = FALSE)
  sampA.perm <- sampComb[permAind]
  sampB.perm <- sampComb[-permAind]
  permStats[j] <- mean(sampA.perm) - mean(sampB.perm)
}
perm.pval <- 2 * min(mean(testStat >= permStats), mean(testStat <= permStats))
perm.pval
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: slide
# Easiest to visualize this p-value
hist(permStats, 
     breaks = 30, 
     col = "#ce72fc", 
     main = "Histogram of Permuted Test Statistics",
     xlab = "Permuted Statistics")
abline(v = testStat, col = "#72fca9", lwd = 3)
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Can assess the performance by performing the previous code with many datasets
nSim <- 1000
nPerm <- 500
n1 <- 40
n2 <- 20
pvals.perm <- rep(0, nSim)
```

## Permutation Tests: Simulations {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Can assess the performance by performing the previous code with many datasets
nSim <- 1000
nPerm <- 500
n1 <- 40
n2 <- 20
pvals.perm <- rep(0, nSim)

for (i in 1:nSim) {
  sampA <- rnorm(n1, 0, 1)
  sampB <- rnorm(n2, 0, 10)
  sampComb <- c(sampA, sampB)
  testStat <- mean(sampA) - mean(sampB)
}
```

## Permutation Tests: Simulations {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Can assess the performance by performing the previous code with many datasets
nSim <- 1000
nPerm <- 500
n1 <- 40
n2 <- 20
pvals.perm <- rep(0, nSim)

for (i in 1:nSim) {
  sampA <- rnorm(n1, 0, 1)
  sampB <- rnorm(n2, 0, 10)
  sampComb <- c(sampA, sampB)
  testStat <- mean(sampA) - mean(sampB)
  
  permStats <- rep(0, nPerm)
  for (j in 1:nPerm) {
    permAind <- sample(1:(n1 + n2), n1, replace = FALSE)
    sampA.perm <- sampComb[permAind]
    sampB.perm <- sampComb[-permAind]
    permStats[j] <- mean(sampA.perm) - mean(sampB.perm)
  }
}
```

## Permutation Tests: Simulations {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Can assess the performance by performing the previous code with many datasets
nSim <- 1000
nPerm <- 500
n1 <- 40
n2 <- 20
pvals.perm <- rep(0, nSim)

for (i in 1:nSim) {
  sampA <- rnorm(n1, 0, 1)
  sampB <- rnorm(n2, 0, 10)
  sampComb <- c(sampA, sampB)
  testStat <- mean(sampA) - mean(sampB)
  
  permStats <- rep(0, nPerm)
  for (j in 1:nPerm) {
    permAind <- sample(1:(n1 + n2), n1, replace = FALSE)
    sampA.perm <- sampComb[permAind]
    sampB.perm <- sampComb[-permAind]
    permStats[j] <- mean(sampA.perm) - mean(sampB.perm)
  }
  
  pvals.perm[i] <- 2 * min(mean(testStat >= permStats), mean(testStat <= permStats))
}

mean(pvals.perm < 0.05)
```
## Homework 6 - Question 2 - Part (d) {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Construct three distributions F, G, and H such that for $X \sim F$, $Y \sim G$, and $Z \sim H$, we have

$$
P(X > Y) > 0.5, \qquad
P(Y > Z) > 0.5, \qquad
P(Z > X) > 0.5
$$

Hints: One way to accomplish this is to have $G$ skewed one way and $H$ skewed the other way, and then shift them up and down the line (by adding or subtracting a constant). So for instance, maybe $F = N(0, 1)$, $G = \chi^2(3) + c_1$, $H = - \chi^2(3) + c_2$. Other skewed distributions would work as well or possibly better.

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
cGrid <- expand.grid(c1Sequence, c2Sequence)
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
cGrid <- expand.grid(c1Sequence, c2Sequence)
cGrid
```

## Homework 6 - Question 2 - Part (d) { auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: fragment
### See how it got it's name, eh?
plot(cGrid, col = "purple")
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
cGrid <- expand.grid(c1Sequence, c2Sequence)

probMat <- array(NA, dim = c(nrow(cGrid), 3))
n <- 1000
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
cGrid <- expand.grid(c1Sequence, c2Sequence)

probMat <- array(NA, dim = c(nrow(cGrid), 3))
n <- 1000
for (i in 1:nrow(cGrid)) {
  c1 <- cGrid[i, 1]
  c2 <- cGrid[i, 2]
}
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
cGrid <- expand.grid(c1Sequence, c2Sequence)

probMat <- array(NA, dim = c(nrow(cGrid), 3))
n <- 1000
for (i in 1:nrow(cGrid)) {
  c1 <- cGrid[i, 1]
  c2 <- cGrid[i, 2]
  
  X <- rnorm(n, 0, 1)
  Y <- rchisq(n, df = 3) + c1
  Z <- -rchisq(n, df = 3) + c2
}
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
### Grid search
c1Sequence <- c2Sequence <- seq(-10, 10, length = 50)
cGrid <- expand.grid(c1Sequence, c2Sequence)

probMat <- matrix(NA, nrow = nrow(cGrid), ncol = 3)
n <- 1000
for (i in 1:nrow(cGrid)) {
  c1 <- cGrid[i, 1]
  c2 <- cGrid[i, 2]
  
  X <- rnorm(n, 0, 1)
  Y <- rchisq(n, df = 3) + c1
  Z <- -rchisq(n, df = 3) + c2
  
  probMat[i, 1] <- mean(X > Y)
  probMat[i, 2] <- mean(Y > Z)
  probMat[i, 3] <- mean(Z > X)
}
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
probMat
```

## Homework 6 - Question 2 - Part (d) {auto-animate="true" auto-animate-easing="ease-in-out"}
- Find which rows have all 3 probabilities above 0.50
- What $c$ value combos do those correspond to?
- Might need to play around with your grid range and coarseness
- Once you find some $c$ value that work you can use essentially the same code just no need to change $c_{1}, c_{2}$ each iteration
- There are other methods aside from where this hint is leading you.