---
title: "(ST 551) Lab 7 <br> Efficiency & Tests"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
    code-line-numbers: TRUE
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

<style>
.slightlysmaller {
  font-size: 0.85em;   /* Slightly smaller than normal */
}

.evensmaller {
  font-size: 0.666em;   /* Slightly smaller than normal */
}

.bigger {
  font-size: 1.15em;   /* Slightly bigger than normal */
}

.biggest {
  font-size: 1.3em;    /* Larger, if needed */
}

.footnote {
  position: absolute;
  bottom: 1em;
  font-size: 0.8em;
  color: #555;
}
</style>

## Welcome Back
* Questions, Comments, Concerns? 

## [Homework Comments -- Some Dos and Donts]{.r-fit-text} {.smaller}
- When using Rmarkdown, you don't need to use `cat()` statements to print everything... you can just write in plain text (outside the code chunks)
- Please show your R code
- Don't hide answers in R comments
- Try to make different answers to different questions distinct

<div class="footnote fragment" style="position: absolute; bottom: 1em; font-size: 0.8em;">
   If you have any formatting questions, feel free to come ask me and I can try and help
</div>

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
Consider:

::: {.fragment}
$$
n = 20, \qquad \mathbf{X}_{1} \equiv X_{11}, \ldots, X_{n1} \overset{\text{iid}}{\sim} N(0, 1)
$$
:::

::: {.fragment}
Let's say we have `nSim` of these samples all stored in a matrix
$$
\mathbf{X}_{(n \times \texttt{nSim})} = \begin{pmatrix} \mathbf{X}_{1} & \mathbf{X}_{2} & \cdots & \mathbf{X}_{\texttt{nSim}} \end{pmatrix}
$$
:::

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 20
nSim <- 1000000
dataMat <- matrix(rnorm(n * nSim), nrow = n, ncol = nSim)
```
:::

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
col.means.1 <- rep(0, nSim)
for (i in 1:nSim) {
  col.means.1[i] <- mean(dataMat[, i])
}
col.means.1[1:20] # First 20 column means
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
startTime.1 <- proc.time()
col.means.1 <- rep(0, nSim)
for (i in 1:nSim) {
  col.means.1[i] <- mean(dataMat[, i])
}
compTime.1 <- proc.time() - startTime.1
compTime.1
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
startTime.2 <- proc.time()
col.means.2 <- apply(dataMat, 2, mean)
compTime.2 <- proc.time() - startTime.2
compTime.2
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
startTime.3 <- proc.time()
col.means.3 <- rep(1 / n, n) %*% dataMat
compTime.3 <- proc.time() - startTime.3
compTime.3
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
colMeans
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
startTime.4 <- proc.time()
col.means.4 <- colMeans(dataMat)
compTime.4 <- proc.time() - startTime.4
compTime.4
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
library(microbenchmark)
microbenchmark(
  colMeans.builtin = colMeans(dataMat),
  colMeans.vector  = rep(1 / n, n) %*% dataMat,
  times = 100L
)
```

## [Effciency Comparisons in R]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
library(Rcpp)
cppFunction("
NumericVector colMeansCpp(NumericMatrix x) {
  int n = x.nrow();
  int p = x.ncol();
  NumericVector means(p);
  for (int j = 0; j < p; j++) {
    double s = 0.0;
    for (int i = 0; i < n; i++) {
      s += x(i, j);
    }
    means[j] = s / n;
  }
  return means;
}
")
# Example usage:
colMeansCpp(dataMat)[1:20]
```

## [Effciency Comparisons in R]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
library(microbenchmark)
microbenchmark(
  colMeans.builtin = colMeans(dataMat),
  colMeans.vector  = rep(1 / n, n) %*% dataMat,
  colMeans.cpp     = colMeansCpp(dataMat),
  times = 100L
)
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
Consider testing:

::: {.fragment}
$$
H_{0}: \delta = \delta_{0}
$$
:::

::: {.fragment}
with data:
$$
n = 20, \qquad X_{1}, \ldots, X_{n} \overset{\text{iid}}{\sim} \chi^{2}_{(3)}
$$
:::

::: {.fragment}
$$
Y_{i} \overset{}{\sim} N(X_{i}, 1)
$$
:::

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 20
newSamp.A <- rchisq(n, 3)
newSamp.B <- rnorm(n, newSamp.A, 1)
```
:::

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
To compute paired differences and perform a one-sample $t$--test we use the test statistic:
$$
tz\left(\delta_{0}\right) = \frac{\bar{D} - \delta_{0}}{\sqrt{\frac{s^{2}_{D}}{n}}}
$$


## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
newSamp.diffs <- newSamp.A - newSamp.B
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
newSamp.diffs <- newSamp.A - newSamp.B
sampMean.diffs <- mean(newSamp.diffs)  # Computing the mean of differences
sampVar.diffs <- var(newSamp.diffs)    # Computing the variance of differences
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Setting up our test statistic
delta0 <- 0
tstat.diffs <- (sampMean.diffs - delta0) / sqrt(sampVar.diffs / n)
tstat.diffs
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Computing the p-value
pval.diffs <- 2 * (1 - pt(abs(tstat.diffs), n - 1))
pval.diffs
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
newSamp.diffs <- newSamp.A - newSamp.B
sampMean.diffs <- mean(newSamp.diffs)
sampVar.diffs <- var(newSamp.diffs)
delta0 <- 0
tstat.diffs <- (sampMean.diffs - delta0) / sqrt(sampVar.diffs / n)
pval.diffs <- 2 * (1 - pt(abs(tstat.diffs), n - 1))
tstat.diffs
pval.diffs
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
To compute the two sample means, sample variances, and sample covariance, and construct the test statistic directly. we use the test statistic:
$$
t\left(\delta_{0}\right) = \frac{\left(\bar{X} - \bar{Y}\right) - \delta_{0}}{\sqrt{\frac{s^{2}_{X}}{n} + \frac{s^{2}_{Y}}{n} - 2\frac{s^{2}_{XY}}{n}}}
$$

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
sampMean.A <- mean(newSamp.A) # Sample mean X
sampMean.B <- mean(newSamp.B) # Sample mean Y
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
sampMean.A <- mean(newSamp.A)
sampMean.B <- mean(newSamp.B)
sampMean.A - sampMean.B
sampMean.diffs
# We can see the difference is indeed the same (good)
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Calculating sample variances and covariance
sampVar.A <- var(newSamp.A)
sampVar.B <- var(newSamp.B)
sampCov.AB <- cov(newSamp.A, newSamp.B)
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Difference variance estimate (denominator of test stat squared)
diffVarEst <- sampVar.A / n + sampVar.B / n - 2 * sampCov.AB / n
diffVarEst
sampVar.diffs / n
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Test stat
tstat.pair <- (sampMean.A - sampMean.B) / sqrt(diffVarEst)
tstat.pair
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We see it matches the other method's test stat (good)
tstat.pair
tstat.diffs
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We have R to do all this for us though
t.test(newSamp.A - newSamp.B)
t.test(newSamp.diffs)
```

## Paired t-Test {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# Perform a paired two-sample t-test directly
t.test(newSamp.A, newSamp.B, paired = TRUE)
```

## [Pearson Chi-Squared Test for Two-Sample Binomial Proportions]{.r-fit-text} {.evensmaller auto-animate="true" auto-animate-easing="ease-in-out"}
Suppose we sampled \(m = 25\) people from Eastern Oregon and \(n = 25\) from Western Oregon, classifying each as vegetarian or non-vegetarian:

\begin{array}{lccc}
    & \text{Western Oregon} & \text{Eastern Oregon} & \text{Totals} \\
    \hline
    \text{Vegetarian} & 15 & 5 & 20 \\
    \text{Non-vegetarian} & 10 & 20 & 30 \\
    \text{Totals} & 25 & 25 & 50 \\
\end{array}

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
tabA <- cbind(c(15, 10), c(5, 20))
tabA
```
:::

## [Pearson Chi-Squared Test for Two-Sample Binomial Proportions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
testRslt <- chisq.test(tabA, correct = FALSE)
testRslt
```

## [Pearson Chi-Squared Test for Two-Sample Binomial Proportions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
names(testRslt)
```

## [Pearson Chi-Squared Test for Two-Sample Binomial Proportions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
testRslt2 <- prop.test(tabA, correct = FALSE)
testRslt2
```