---
title: "(ST 551) Lab 8 <br> Contingency Tables"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
    code-line-numbers: TRUE
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
---

<style>
.slightlysmaller {
  font-size: 0.85em;   /* Slightly smaller than normal */
}

.evensmaller {
  font-size: 0.666em;   /* Slightly smaller than normal */
}

.bigger {
  font-size: 1.15em;   /* Slightly bigger than normal */
}

.biggest {
  font-size: 1.3em;    /* Larger, if needed */
}

.footnote {
  position: absolute;
  bottom: 1em;
  font-size: 0.8em;
  color: #555;
}
</style>

## No Lab Next Week
* Questions, Comments, Concerns?


## Last Episode... {.evensmaller}
* Last week we talked about the Chi-squared test of homogeneity or independence with the following scenario:
  * Suppose we sampled $m = 25$ people from Eastern Oregon and $n = 25$ from Western Oregon, classifying each as vegetarian or non-vegetarian:

::: {.fragment}
\begin{array}{lccc}
    & \text{Western Oregon} & \text{Eastern Oregon} & \text{Totals} \\
    \hline
    \text{Vegetarian} & 15 & 5 & 20 \\
    \text{Non-vegetarian} & 10 & 20 & 30 \\
    \text{Totals} & 25 & 25 & 50 \\
\end{array}
:::

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
tabA <- cbind(c(15, 10), c(5, 20))
tabA
```
:::

## Last Episode... {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: fragment
testRslt <- chisq.test(tabA, correct = FALSE)
testRslt
```

## Last Episode... {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
testRslt <- chisq.test(tabA, correct = FALSE)
testRslt
testRslt2 <- prop.test(tabA, correct = FALSE)
testRslt2
```

## [This Episode - Fisher's Exact Test]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
fisherRslt <- fisher.test(tabA)
```

## [This Episode - Fisher's Exact Test]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
fisherRslt <- fisher.test(tabA)
fisherRslt
```

## [This Episode - Fisher's Exact Test]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
names(fisherRslt)
```


## [Fisher's Exact Test... Manually :(]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Pros
  - As long as there is no pairing (our groups are independent) we can always use Fisher's Exact test because we can always estimate an odds ratio
- Cons
  - It's a total pain in the butt to calculate by hand (essentially impossible in most cases)
  - There are just better / easier alternatives (chi-squared test, log-odds tests, difference in proportions test) in most cases
- So...
  - In an exam settings, I would pretty much never use this test unless I have to (aka the question tells me I have to)

## [Fisher's Exact Test... Manually :(]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- What we testing:
  - $H_{0}: p_{X} = p_{Y}$ (No association between response variable $X$ and grouping variable $G$)
- Test Statistic: 
  - Probability of observed table conditioning on margins
- p-value: 
  - Sum of probability of all tables as or more extreme than observed table
  
  
## The Hypergeometric Distribution {.smaller}
- The `dhyper()` function computes probabilities from the **hypergeometric distribution**:
  - ``dhyper(x, m, n, k)``

- Use it when sampling **without replacement** and counting the number of “successes.”

- In a 2×2 table, the key is:
  - `x` = the count in the **cell of interest**  
  - `m` = total count of the corresponding **row or column** containing that cell  
  - `n = N - m`  
  - `k` = total of the **other dimension** (row or column) containing that cell  

- It does *not* matter whether you treat `m` and `n` as row totals or column totals, but `m` and `k` must correspond to the row/column of the cell in question (here, the first-row, first-column cell).

---

## Our Example {.evensmaller}
We observed:

$$
\begin{array}{lccc}
    & \text{Western Oregon} & \text{Eastern Oregon} & \text{Totals} \\
    \hline
    \text{Vegetarian} & 15 & 5 & 20 \\
    \text{Non-vegetarian} & 10 & 20 & 30 \\
    \text{Totals} & 25 & 25 & 50 \\
\end{array}
$$

The observed value in cell $[1,1]$ is:

$$
x = 15
$$

There are two equivalent ways to parameterize the hypergeometric distribution:

- Treat **row totals** (Vegetarian vs Non-vegetarian) as `m` and `n`  
- Treat **column totals** (Western vs Eastern Oregon) as `m` and `n`
- Both yield the same probability.

---

## [Computing the Probability of the Observed Table]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
dhyper(tabA[1,1], m = 25, n = 25, k = 20)
```

## [Computing the Probability of the Observed Table]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
dhyper(tabA[1,1], m = 25, n = 25, k = 20)
dhyper(tabA[1,1], m = 20, n = 30, k = 25)
```

## [Computing the Probability of the Observed Table]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
dhyper(tabA[1,1], m = 25, n = 25, k = 20)
obsTabProb <- dhyper(tabA[1,1], m = 25, n = 25, k = 20)
```

## [Computing the Probability of the Observed Table]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
Either case shown computes:

$$
P\left(X = 15\right) = \frac{\binom{m}{15} \binom{n}{k - 15}}{\binom{50}{k}} = \frac{\binom{25}{15} \binom{25}{5}}{\binom{50}{20}}
$$

## [Computing the Probability of the Observed Table]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
One could also always just do this ``even more manually'' by:

$$
P\left(X = 15\right)= \frac{\binom{25}{15} \binom{25}{5}}{\binom{50}{20}}
$$

::: {.fragment}
```{r}
#| echo: TRUE
#| eval: TRUE
choose(25, 15) * choose(25, 5) / choose(50, 20)
dhyper(15, m = 25, n = 25, k = 20)
```
:::



## [Probabilities of All Possible Tables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
The possible values in cell $[1, 1]$ range from 0 to 20:
$$
\begin{array}{|c|c|}
    \hline 
    k & 20 - k \\
    \hline
    25 - k & 25 - (20 - k) \\
    \hline
\end{array}
    \qquad \text{for } k = 0, 1, \ldots, 20
$$

::: {.fragment}
To compute hypergeometric probabilities for each
```{r}
#| echo: TRUE
#| eval: TRUE
dhyper(0:20, m = 25, n = 25, k = 20)
allTabsProbs <- dhyper(0:20, m = 25, n = 25, k = 20)
```
:::

## [p-value Time]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# The indices of probabilities ≤ the observed table probability
which(allTabsProbs <= obsTabProb) 
```

## [p-value Time]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# The indices of probabilities ≤ the observed table probability
which(allTabsProbs <= obsTabProb)

# The corresponding values of k producing such tables
(0:20)[allTabsProbs <= obsTabProb]
```

## [p-value Time]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
allTabsProbs[allTabsProbs <= obsTabProb]
```

## [p-value Time]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
allTabsProbs[allTabsProbs <= obsTabProb]
sum(allTabsProbs[allTabsProbs <= obsTabProb])
```

## [p-value Time]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# We can check it matches the R function
sum(allTabsProbs[allTabsProbs <= obsTabProb])
fisher.test(tabA)[["p.value"]]
```

## [p-value Time]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\sum\limits_{i \in \mathcal{T}}P\left(\text{table}_{i}\right) \qquad \text{where } \mathcal{T} \equiv \left\{P\left(\text{table}_{i}\right) \le P\left(\text{observed table}\right)\right\}
$$

## Paired Binary Data
- What happens when we don't have independence between groups?
- We need to account for that correlation structure somehow
- As an example we can generate some paired binary data and try out our Paired $t$-test and McNemar's test

## Generating Paired Binary Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
We want to generate **two paired binary samples**  
$$
(X_1, Y_1), (X_2, Y_2), \ldots, (X_n, Y_n)
$$
that may be **correlated**. One approach is:

1. Generate each $X_i$ first
2. Generate $Y_i$ **conditionally** on $X_i$

- This allows the success probability of $Y_i$ to depend on the value of $X_i$ (thus not independent)

---

## Generating Paired Binary Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Step 1: Generate the X Sample 
  - We begin by choosing a success probability for $X_i$:

::: {.fragment}
$$
P(X_i = 1) = p_X
$$
:::

---

## Generating Paired Binary Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Step 2: Conditional Success Probability for $Y$
  - To induce correlation, we generate $Y_i$ using:

::: {.fragment}
$$
P(Y_i = 1 \mid X_i = 1) = p_{Y|1},
\qquad
P(Y_i = 1 \mid X_i = 0) = p_{Y|0}
$$
:::


::: {.fragment}
Then the marginal success probability for $Y_i$ is:

$$
\begin{aligned}
P(Y_i = 1)
&= p_{Y|1} \cdot p_X + p_{Y|0} \cdot (1 - p_X)
\end{aligned}
$$

:::

- This ensures that the distribution of $Y$ depends on the value of $X$.

---

## Generating Paired Binary Data {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
Matching or Violating the Null Hypothesis, if we want the null hypothesis

$$
H_0 : p_X = p_Y
$$

to hold, we must choose $p_{Y|1}$ and $p_{Y|0}$ such that:

$$
p_X = p_{Y|1}\,p_X + p_{Y|0}(1 - p_X)
$$

Otherwise the null hypothesis is false. This gives control over whether the paired samples have equal or unequal marginal probabilities.

---

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
Lastly, we solve for $p_{Y|0}$. If we **fix** $p_{Y|1}$, we can solve for $p_{Y|0}$:

$$
p_{Y|0}
= \frac{p_X - p_{Y|1}p_X}{1 - p_X}
= p_X \cdot \frac{1 - p_{Y|1}}{1 - p_X}
$$

This lets us construct any desired $p_Y$.

---

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
pX <- 0.5      # P(X = 1)
pY1 <- 0.7     # P(Y = 1 | X = 1)

# Solve for pY0 so that p_Y has desired value
pY0 <- pX * (1 - pY1) / (1 - pX)

# Marginal P(Y = 1)
pY <- pY1 * pX + pY0 * (1 - pX)
pY
```

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
That's the basic idea. With our marginal probabilities worked out. We can generate some data:

- We first set a sample size, 
- then generating the $X$ sample,
- then generating the $Y$ sample conditional on the $X$ sample.


## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 50
```

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 50
sampX <- rbinom(n, 1, pX)
```

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 50
sampX <- rbinom(n, 1, pX)
table(sampX)
```

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 50
sampX <- rbinom(n, 1, pX)
rbind(sampX, pY1 * (sampX == 1) + pY0 * (sampX == 0)) # Each pair displayed
```

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 50
sampX <- rbinom(n, 1, pX)
sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
```

## Generating Paired Binary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 50
sampX <- rbinom(n, 1, pX)
sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
table(sampX, sampY)
```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
t.test(sampX, sampY, paired = TRUE)
```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
t.test(sampX, sampY, paired = TRUE)
mcnemar.test(sampX, sampY, correct = FALSE)
```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
pval.mc <- pval.t <- rep(0, nsim)
```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nsim <- 10000
pval.mc <- pval.t <- rep(0, nsim)
for (i in 1:nsim) {
  sampX <- rbinom(n, 1, pX)
  sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
}
```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
nsim <- 10000
pval.mc <- pval.t <- rep(0, nsim)
for (i in 1:nsim) {
  sampX <- rbinom(n, 1, pX)
  sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
  pval.mc[i] <- mcnemar.test(sampX, sampY, correct = FALSE)$p.val
  pval.t[i] <- t.test(sampX, sampY, paired = TRUE)$p.val
}

```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
pval.mc <- pval.t <- rep(0, nsim)
for (i in 1:nsim) {
  sampX <- rbinom(n, 1, pX)
  sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
  pval.mc[i] <- mcnemar.test(sampX, sampY, correct = FALSE)$p.val
  pval.t[i] <- t.test(sampX, sampY, paired = TRUE)$p.val
}
mean(pval.mc < 0.05)
mean(pval.t < 0.05)
```

## Testing Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
plot(pval.mc, pval.t, 
     col = "orchid", 
     pch = 17)
abline(0, 1)
```

## Testing Time - Part 2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
pX <- 1 / 3      # P(X = 1)
pY1 <- 0.50      # P(Y = 1 | X = 1)

# Solve for pY0 so that p_Y has desired value
pY0 <- pX * (1 - pY1) / (1 - pX)

# Marginal P(Y = 1)
pY <- pY1 * pX + pY0 * (1 - pX)
pY
```

## Testing Time - Part 2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
n <- 25
sampX <- rbinom(n, 1, pX)
sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
table(sampX, sampY)
```


## Testing Time - Part 2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
nsim <- 10000
pval.mc <- pval.t <- rep(0, nsim)
for (i in 1:nsim) {
  sampX <- rbinom(n, 1, pX)
  sampY <- rbinom(n, 1, pY1 * (sampX == 1) + pY0 * (sampX == 0))
  pval.mc[i] <- mcnemar.test(sampX, sampY, correct = FALSE)$p.val
  pval.t[i] <- t.test(sampX, sampY, paired = TRUE)$p.val
}
mean(pval.mc < 0.05)
mean(pval.t < 0.05)
```

## Testing Time - Part 2 {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
plot(pval.mc, pval.t, 
     col = "lawngreen", 
     pch = 17)
abline(0, 1)
```
