---
title: "(ST 551) Lab 1 <br> Welcome to R"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

```{r}
#| echo: FALSE
# Here if I want to load something in the background
```

## Hello! {.smaller}
* Who am I?
  * A deep philosophical question
  * So instead I will answer any questions anyone may have about my identity
  * If no questions I am content to be anonymous for the term and we can continue
* Office Hours
  * Wednesday 3:00pm - 4:00pm (right before this)
    * All the good spots were taken so fast! :(
* Lab Structure
  * I'll yap for a while
  * After I have put everyone to sleep, I can answer any homework (or whatever) questions
  
## Let's Clear Some Things Up
* Sarah challenged ME to a dual
  * I don't play Tetris all day, just a couple weeks before we played
* Next, I should all congratulate you all
  * You answering of the three questions was perfect! -- Don't let her corrupt your minds

## R We Ready To Go?
* Hopefully everyone has R and RStudio installed on their system

## What's Happening Today?
* Kinda boring day today (if you've used R before), we're just covering some basic commands in R
* There is an R script on Canvas you're welcome to use to follow along, `RLab1.R`
* These slides go in a slightly different order than the script and lab PDF
  * Let me know if you find typos and such in the Lab PDFs :|
  
## Today's Dataset {.smaller}
* `case0102` (Sex Discrimination in Employment)
* **Description**: The data are the beginning salaries for all 32 male and all 61 female skilled, entry–level clerical employees hired by a bank between 1969 and 1977.
* 93 observations
* 2 variables:
  * `Salary`: starting salaries (in US$)
  * `Sex`: sex of the clerical employee, with levels "Female" and "Male"

## Loading Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# If package is not installed: install.packages("Sleuth3")
library(Sleuth3)
```

## Dimensions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
nrow(case0102)
```

## Dimensions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
nrow(case0102)
ncol(case0102)
```

## Dimensions {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
nrow(case0102)
ncol(case0102)
dim(case0102)
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
head(case0102) # I guess not technically subsetting...
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0102[1:10, ] # First 10 observations
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
case0102[1:15, 1] # First 15 observations and first column
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# If we want to look at the first column (all observations)
case0102[, 1]
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Since we're working with a dataframe (a list) we can also subset by
case0102$Salary
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Since we're working with a dataframe (a list) we can also subset by
case0102$Salary
case0102[[1]]
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Or when the naming attribute is present
case0102[, "Salary"]
```

## Subsetting {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Or when the naming attribute is present
case0102[, "Salary"]
case0102[["Salary"]]
```

## Structure and Attributes {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# How to check if we have the `names' attribute
names(case0102)
```

## Structure and Attributes {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# How to check if we have the `names' attribute
names(case0102)
colnames(case0102) # Equivalent to names() on a dataframe
```

## Structure and Attributes {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# The overall structure of an object
str(case0102)
```

## How Many Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Which elements of this vector are `male'
case0102$Sex == "Male"
```

## How Many Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Which elements of this vector are `male'
case0102$Sex == "Male"

# Count how many `TRUE's (1s) there are
sum(case0102$Sex == "Male")
```

## How Many Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have also done:
table(case0102$Sex)
```

## How Many Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Or even:
summary(case0102$Sex)
```

## Average Salary? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# 5 number summary, but no mean
summary(case0102$Salary)
```

## Average Salary? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# 5 number summary, but no mean
summary(case0102$Salary)

mean(case0102$Salary) # Easy fix
```

## Average Salary for Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Salary for just male observations
case0102$Salary[case0102$Sex == "Male"]
```

## Average Salary for Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Salary for just male observations
case0102$Salary[case0102$Sex == "Male"]

# Summing up all `TRUE's (1s)
sum(case0102$Salary[case0102$Sex == "Male"])
```

## Average Salary for Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have used the `subset' function too
case0102.males <- subset(case0102, Sex == "Male")
```

## Average Salary for Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have used the `subset' function too
case0102.males <- subset(case0102, Sex == "Male")

# Returns a subsetted dataframe
head(case0102.males)
```

## Average Salary for Males? {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Could have used the `subset' function too
case0102.males <- subset(case0102, Sex == "Male")

# Returns a subsetted dataframe
head(case0102.males)

# Can now take the mean of the `Salary' variable (column)
mean(case0102.males$Salary)
```

## Histogram of Salary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Default plotting (no additional arguments)
hist(case0102$Salary)
```

## Histogram of Salary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Can specify bin width
hist(case0102$Salary, breaks = seq(3750, 8250, 250))
```

## Sequence {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# How does the seq() function work?
seq(1, 4, 0.5)
```

## Sequence {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# How does the seq() function work?
seq(1, 4, 0.5)
seq(-4,-6, -1)
```

## Sequence {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# How does the seq() function work?
seq(1, 4, 0.5)
seq(-4,-6, -1)
seq(1, 4, length = 10) 
```

## Sequence {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# How does the seq() function work?
seq(1, 4, 0.5)
seq(-4,-6, -1)
seq(1, 4, length = 10) 
seq(-4,-6, length = 5)
```

## Histogram of Salary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding a title
hist(case0102$Salary, 
     breaks = seq(3750, 8250, 250), 
     main = "Histogram of Salaries")
```

## Histogram of Salary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding a label
hist(case0102$Salary,
     breaks = seq(3750, 8250, 250),
     main = "Histogram of Salaries",
     xlab = "Salary")
```

## Histogram of Salary Data {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Adding some color
hist(case0102$Salary,
     breaks = seq(3750, 8250, 250),
     main = "Histogram of Salaries",
     xlab = "Salary",
     col = "lightblue")
```

## [Histogram of Salary Data (For The Boys)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Histogram of just the guys
hist(case0102$Salary[case0102$Sex == "Male"])
```

## [Histogram of Salary Data (For The Boys)]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Histogram of just the guys -- another way
hist(case0102.males$Salary)
```

## Boxplot Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
boxplot(Salary ~ Sex, data = case0102)
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
samp.A <- rexp(30)
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
samp.A
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
mean(samp.A) # Mean
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
var(samp.A) # Variance
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
median(samp.A) # Median
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
sd(samp.A) # Standard Deviation
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
range(samp.A) # Range
```

## Summary Statistics {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Each run of this code will generate another (random) sample
samp.B <- rexp(30)
samp.B; mean(samp.B); var(samp.B); median(samp.B); sd(samp.B); range(samp.B)
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Basic `for' loop
for (i in 1:10) {
  print(i)
}
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Basic `for' loop
for (i in c(13, 17, 32)) {
  print(i)
}
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Basic `for' loop
for (i in c(1/10, 1/100, 1/1000)) {
  print(i^2)
}
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Basic `for' loop
rslt <- 0
for (i in 4:12) {
  rslt <- rslt + i
}
rslt
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Basic `for' loop
rsltVec <- rep(0, 12)
for (i in 1:12) {
  rsltVec[i] <- i^2
}
rsltVec
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Basic `for' loop simulation template
nSamp <- 10000
sampMedVec <- rep(0, nSamp)
for (i in 1:nSamp) {
  newSamp <- rexp(30)
  sampMedVec[i] <- median(newSamp)
}
```

## Sampling Distribution Time {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Histogram of results
hist(sampMedVec, 
     main = "Histogram of Sample Median (n = 30)", 
     xlab = "Sample Median")
```