---
title: "(ST 551) Lab 3 <br> t vs z & Binomial Stuff"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: concave
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## Welcome Back Kiddos!
* Back Again... Woohoo...
* Questions, Comments, Concerns?

## Basic R Program {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
```{.asm}
INCLUDE Irvine32.inc

.data
  array       DWORD 10 DUP(?)        ; Reserve space for 10 random integers (4 bytes each)
  msg1        BYTE "Random numbers:", 0
  msg2        BYTE "Mean = ", 0
  msg3        BYTE "Variance = ", 0
  count       DWORD 10               ; Total number of random values to generate
  countMinus1 DWORD ?                ; Will store (count - 1) for variance calculation
  mean        REAL4 ?                ; To store mean as a floating-point value
  
.code
main PROC
  CALL Randomize          ; Intialize seed

  MOV  EDX, OFFSET msg1
  CALL WriteString
  CALL Crlf

  ; Generate random numbers
  MOV  ECX, count         ; Number of variables to draw
  MOV  ESI, OFFSET array  ; Point ESI to first element
generate_loop:            ; Counted loop
  MOV EAX, 100            ; Produce random integer within 
  CALL RandomRange        ; Random integer [0, 100)
  MOV [ESI], EAX          ; Store in array
  CALL WriteDec
  CALL Crlf
  ADD  ESI, 4             ; Advance pointer to next DWORD (next element)
  LOOP generate_loop      ; Decrement ECX, Loop if ECX != 0

  ; Sum generated numbers
  MOV  ECX, count         ; Number of variables to sum
  MOV  ESI, OFFSET array  ; Point ESI to first element
  MOV  EAX, 0             ; Set EAX = 0
sum_loop:
  ADD  EAX, [ESI]         ; Add current element to running total in EAX
  ADD  ESI, 4             ; Advance pointer to next DWORD (next element)
  loop sum_loop           ; Decrement ECX, Loop if ECX != 0

  ; Compute mean
  FILD DWORD PTR count    ; Load count (as integer) onto FPU stack
  PUSH EAX                ; Push total sum onto CPU stack (as integer)
  FILD DWORD PTR [ESP]    ; Load sum (as integer) onto FPU stack
  ADD  ESP, 4             ; Clean up CPU stack (remove pushed value)
  FDIV                    ; Divide sum / count then store result in ST(0)
  FSTP mean               ; Store result (mean) in memory
  FLD  mean               ; Reload mean into FPU for later use

  MOV  EDX, OFFSET msg2              
  CALL WriteString
  CALL WriteFloat                    
  CALL Crlf

  ; Compute variance
  MOV  ECX, count         ; Loop counter = number of elements
  MOV  ESI, OFFSET array  ; Point ESI to first array element
  FLDZ                    ; Initialize FPU register ST(0) = 0 (sum of squares)

var_loop:
  FILD DWORD PTR [ESI]    ; Load current element as integer to float
  FSUB mean               ; Subtract mean
  FMUL ST(0), ST(0)       ; Square the difference
  FADDP ST(1), ST(0)      ; Add to running sum of squares
  ADD  ESI, 4             ; Move to next element
  LOOP var_loop           ; Repeat for all elements

  MOV  EAX, count
  DEC  EAX                ; Subtract 1 to get our (n - 1)
  MOV  countMinus1, EAX   ; Store as countMinus1 for denominator

  FILD countMinus1        ; Load (n - 1) onto FPU stack
  FDIV                    ; Divide sum of squares / (n - 1)
  
  MOV  EDX, OFFSET msg3
  CALL WriteString
  CALL WriteFloat
  CALL Crlf

  EXIT
main ENDP
END main
```

## Where The CLT Doesn't Apply {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# What if we don't have a finite variance?
nsim <- 10000
n <- 1000
```

## Where The CLT Doesn't Apply {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# What if we don't have a finite variance?
nsim <- 10000
n <- 1000

t1.sampMeans <- rep(0, nsim)       # Initialize vector for sample means 
```

## Where The CLT Doesn't Apply {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# What if we don't have a finite variance?
nsim <- 10000
n <- 1000

t1.sampMeans <- rep(0, nsim)       # Initialize vector for sample means 
for (i in 1:nsim) {                # Perform the following nsim times
  t1.samp <- rt(n, 1)              # Draw a sample from a Cauchy distribution
}
```

## Where The CLT Doesn't Apply {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
# What if we don't have a finite variance?
nsim <- 10000
n <- 1000

t1.sampMeans <- rep(0, nsim)       # Initialize vector for sample means 
for (i in 1:nsim) {                # Perform the following nsim times
  t1.samp <- rt(n, 1)              # Draw a sample from a Cauchy distribution
  t1.sampMeans[i] <- mean(t1.samp) # Take and store mean of sample
}
```

## Where The CLT Doesn't Apply {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Plot it!
hist(t1.sampMeans,
     col = "orange",
     border = "black",
     main = "Histogram of Cauchy Sample Means (n = 1000)",
     xlab = "Sample Means")
```

## Where The CLT Doesn't Apply {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Why does it look weird?
range(t1.sampMeans)
```

* If $X \sim t_{(1)}$
  * $\operatorname{E}(X)$ is undefined

## t-tests {auto-animate="true" auto-animate-easing="ease-in-out"}
* Listen up! -- About to give you the secret to Statistics
  * Everything (pretty much) is a $t$-test
  * It's also the best test (fact)

## t-tests {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create a sequence representing the degrees of freedom
t.df <- seq(1, 50, 1)
t.df
```

## t-tests {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Create a sequence representing the degrees of freedom
t.df <- seq(1, 50, 1)

# Create critical values for each element in the sequence
t.crit.05 <- qt(1 - 0.05, t.df)
t.crit.05
```

## t-tests {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot critical value corresponding to their df
plot(t.df, t.crit.05, 
     ylim = c(0, 7), 
     xlab = "Degrees of Freedom (df)",
     ylab = "t Critical Value (0.05)")
```

## t-tests {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Plot critical value corresponding to their df
plot(t.df, t.crit.05, 
     ylim = c(0, 7), 
     xlab = "Degrees of Freedom (df)",
     ylab = "t Critical Value (0.05)")
abline(h = qnorm(1 - 0.05), lty = 3)
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
# If you change the population distribution, you need to change your null hypothesis
mu0 <- 4
t.stat <- rep(0, nsim) # ``Empty'' vector for storing t-tests
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
# If you change the population distribution, you need to change your null hypothesis
mu0 <- 4
t.stat <- rep(0, nsim) # ``Empty'' vector for storing t-tests
for (i in 1:nsim) {
  dataSamp <- rchisq(n, 4)
  t.stat[i] <- (mean(dataSamp) - mu0) / sqrt(var(dataSamp) / n)
}
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
# If you change the population distribution, you need to change your null hypothesis
mu0 <- 4
t.stat <- rep(0, nsim) # ``Empty'' vector for storing t-tests
for (i in 1:nsim) {
  dataSamp <- rchisq(n, 4)
  t.stat[i] <- (mean(dataSamp) - mu0) / sqrt(var(dataSamp) / n)
}
pvals.t <- 2 * (1 - pt(abs(t.stat), n - 1))
pvals.z <- 2 * (1 - pnorm(abs(t.stat)))
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
mean(pvals.t < 0.05)
mean(pvals.z < 0.05)
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
### Comparing the t-stat on a t vs z reference distribution
par(mfrow = c(1, 2))
hist(pvals.t, 
     breaks = seq(0, 1, 0.05),
     col = "lightgreen",
     border = "black",
     main = "Histogram of p-values \n from t-Distribution")
hist(pvals.z, 
     breaks = seq(0, 1, 0.05),
     col = "lightblue",
     border = "black",
     main = "Histogram of p-values \n from Normal Distribution")
```
## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
* Wait... what are we even looking for in these histograms?
* Why do I care about this proportion?
* What proportion of $p$-values would we expect to be below $\alpha = 0.05$ if the null hypothesis is true

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
mu0 <- 9               # Notice we had to change the null 
t.stat <- rep(0, nsim)
for (i in 1:nsim) {
  dataSamp <- rgamma(n, shape = 3, scale = 3)
  t.stat[i] <- (mean(dataSamp) - mu0) / sqrt(var(dataSamp) / n)
}
pvals.t <- 2 * (1 - pt(abs(t.stat), n - 1))
pvals.z <- 2 * (1 - pnorm(abs(t.stat)))
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "4"
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
mu0 <- 9               # Notice we had to change the null 
t.stat <- rep(0, nsim)
for (i in 1:nsim) {
  dataSamp <- rgamma(n, shape = 3, scale = 3)
  t.stat[i] <- (mean(dataSamp) - mu0) / sqrt(var(dataSamp) / n)
}
pvals.t <- 2 * (1 - pt(abs(t.stat), n - 1))
pvals.z <- 2 * (1 - pnorm(abs(t.stat)))
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| code-line-numbers: "7"
### Comparing the t-stat on a t vs z reference distribution
n <- 30
nsim <- 10000
mu0 <- 9               # Notice we had to change the null 
t.stat <- rep(0, nsim)
for (i in 1:nsim) {
  dataSamp <- rgamma(n, shape = 3, scale = 3)
  t.stat[i] <- (mean(dataSamp) - mu0) / sqrt(var(dataSamp) / n)
}
pvals.t <- 2 * (1 - pt(abs(t.stat), n - 1))
pvals.z <- 2 * (1 - pnorm(abs(t.stat)))
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
### Comparing the t-stat on a t vs z reference distribution
par(mfrow = c(1, 2))
hist(pvals.t, 
     breaks = seq(0, 1, 0.05),
     col = "lightgreen",
     border = "black",
     main = "Histogram of p-values \n from t-Distribution")
hist(pvals.z, 
     breaks = seq(0, 1, 0.05),
     col = "lightblue",
     border = "black",
     main = "Histogram of p-values \n from Normal Distribution")
```

## t vs z {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
mean(pvals.t < 0.05)
mean(pvals.z < 0.05)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
n <- 20
p <- 0.5
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
n <- 20
p <- 0.5
x <- 0:n
pmf.x <- dbinom(x, n, p)
cdf.x <- pbinom(x, n, p)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
par(mfrow = c(2, 2))
plot(x, pmf.x, 
     type = "h",
     col = "lightpink",
     main = paste("Binomial(", n, ", ", p, ") pmf", sep = ""),
     ylab = "p(x)", xlab = "x")
plot(x, cdf.x, 
     type = "h",
     col = "lightblue",
     main = paste("Binomial(", n, ", ", p, ") cdf", sep = ""),
     ylab = "F(x)", 
     xlab = "x")
barplot(pmf.x, 
     names = x,
     col = "lightpink",
     main = paste("Binomial(", n, ", ", p, ") pmf", sep = ""),
     ylab = "p(x)", 
     xlab = "x")
barplot(cdf.x, 
     names = x,
     col = "lightblue",
     main = paste("Binomial(", n, ", ", p, ") cdf", sep = ""),
     ylab = "F(x)", 
     xlab = "x")
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# We can try our CLT approximation
approx.cdf.x <- pnorm(x, n*p, sqrt(n*p*(1 - p)))
cbind(cdf.x, approx.cdf.x)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
par(mfrow = c(1, 1)) # Return plotting window back to normal
plot(x, cdf.x, 
     type = "h",
     main = paste("Binomial(", n, ", ", p, ") cdf", sep = ""),
     ylab = "F(x)", 
     xlab = "x", 
     lwd = 3)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot(x, cdf.x, 
     type = "h",
     main = paste("Binomial(", n, ", ", p, ") cdf", sep = ""),
     ylab = "F(x)", 
     xlab = "x", 
     lwd = 3)
points(x, approx.cdf.x, type = "h", col = 2, lwd = 3)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot(x, cdf.x, 
     type = "h",
     main = paste("Binomial(", n, ", ", p, ") cdf", sep = ""),
     ylab = "F(x)", 
     xlab = "x", 
     lwd = 3)
points(x + 0.25, approx.cdf.x, type = "h", col = 2, lwd = 3)
```


## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Not bad, but consistently a tad low
# To see why, consider
pbinom(3, n, p)
pbinom(3.5, n, p)
pbinom(3.9, n, p)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Not bad, but consistently a tad low
# To see why, consider
pbinom(3, n, p)
pbinom(3.5, n, p)
pbinom(3.9, n, p)
# Since our data is discrete, the cdf is the same for x in [3.0, 4)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Not the case for the Normal distribution (since continuous)
pnorm(3, n * p, sqrt(n * p * (1 - p)))
pnorm(3.5, n * p, sqrt(n * p * (1 - p)))
pnorm(3.9, n * p, sqrt(n * p * (1 - p)))
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# We can try a ``continuity correction''
approx.cdf.cc.x <- pnorm(x + 0.5, n * p, sqrt(n * p * (1 - p)))
cbind(cdf.x, approx.cdf.x, approx.cdf.cc.x)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot(x, cdf.x, 
     type = "h",
     main = paste("Binomial(", n, ", ", p, ") cdf", sep = ""),
     ylab = "F(x)", 
     xlab = "x", 
     lwd = 3)
points(x + 0.25, approx.cdf.x, type = "h", col = 2, lwd = 3)
points(x + 0.5, approx.cdf.cc.x, type = "h", col = 3, lwd = 3)
legend(0, 1, 
       legend = c("Exact CDF", "Normal Approx", "Normal Approx w/ CC"), 
       col = c(1, 2, 3), 
       lwd = c(3, 3, 3))
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| output-location: fragment
# Intuition behind continuity correction:
# Generate a lot of samples to get an accurate sampling distribution
rbinom(1000000, n, p) 
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Intuition behind continuity correction:
hist(rbinom(1000000, n, p),
     breaks = seq(-0.5, n + 0.5, 1),
     col = "#eda3ff",
     main = "Displaying Continuity Correction Reason",
     xlab = "X ~ Binomial(n, p)",
     freq = FALSE)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Intuition behind continuity correction:
hist(rbinom(1000000, n, p),
     breaks = seq(-0.5, n + 0.5, 1),
     col = "#eda3ff",
     main = "Displaying Continuity Correction Reason",
     xlab = "X ~ Binomial(n, p)",
     freq = FALSE)
lines(seq(0, n, 0.01), dnorm(seq(0, n, 0.01), n * p, sqrt(n * p * (1 - p))), col = "#42f578", lwd = 2)
```

## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Intuition behind continuity correction:
hist(rbinom(1000000, n, p),
     breaks = seq(-0.5, n + 0.5, 1),
     col = "#eda3ff",
     main = "Displaying Continuity Correction Reason",
     xlab = "X ~ Binomial(n, p)",
     freq = FALSE)
lines(seq(0, n, 0.01), dnorm(seq(0, n, 0.01), n * p, sqrt(n * p * (1 - p))), col = "#42f578", lwd = 2)
abline(v = 11, lwd = 2)
```


## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
pbinom(11, n, p)                                           # Exact Probability
pnorm(11, mean = n * p, sd = sqrt(n * p * (1 - p)))        # No CC      
pnorm(11 + 0.50, mean = n * p, sd = sqrt(n * p * (1 - p))) # CC
```
## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
pbinom(11, n, p)                                           # Exact Probability
pnorm(11, mean = n * p, sd = sqrt(n * p * (1 - p)))        # No CC      
pnorm(11 + 0.50, mean = n * p, sd = sqrt(n * p * (1 - p))) # CC
```
## Binomial Stuff {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
pbinom(11, n, p)                                           # Exact Probability
pnorm(11, mean = n * p, sd = sqrt(n * p * (1 - p)))        # No CC      
pnorm(11 + 0.50, mean = n * p, sd = sqrt(n * p * (1 - p))) # CC
```

## Binomial Stuff {.smaller}
Let 
$$
Y \sim \text{Binomial}(n = `r n`, p = `r p`)
$$
Using the Normal approximation to the Binomial distribution via the Central limit Theorem here. Our standardized quantity will be
$$
\frac{Y - \operatorname{E}(X)}{\sqrt{\operatorname{Var}(X)}}  = \frac{Y - np}{\sqrt{np(1-p)}} \overset{d}{\to} N(0, 1)
$$
Let's say the probability of interest is $P(Y \leq 11)$. Using the normal approximation with continuity correction:
\begin{align*}
    P\left(Y \le 11\right) & = P\left(Y \le 11.5\right) \\[0.25cm]
    & = P\left(\frac{Y - np}{\sqrt{np(1-p)}} \le \frac{11.5 - np}{\sqrt{np(1-p)}}\right) \\[0.25cm]
    & = P\left(\frac{Y - `r n * p`}{`r sqrt(n * p * (1 - p))`} \le \frac{11.5 - `r n * p`}{`r sqrt(n * p * (1 - p))`}\right) \\[0.25cm]
    & \approx P\left(Z \le `r (11.5 - n * p) / sqrt(n * p * (1 - p))`\right) \\[0.25cm]
    & = `r pnorm((11.5 - n * p) / sqrt(n * p * (1 - p)))`
\end{align*}

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
The following is very similar to Question 4, Homework 2:

* Exploring the performance (exactness) of the $t$-test for data from different distributions.
* Consider $n = 25$
* Let $X \sim \text{Gamma}\left(\text{shape} = 3, \text{scale} = 3\right)$
* We want to generate a sample many times from this distribution and for each, calculate the $t$-statistic at the end we'll see (when the null is true) the proportion of $p$-values under 

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
}
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
  sampMean <- mean(data)
  sampSD <- sd(data)
}
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
  sampMean <- mean(data)
  sampSD <- sd(data)
  tStat[i] <- sqrt(n) * (sampMean - trueMean) / sampSD
}
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
  sampMean <- mean(data)
  sampSD <- sd(data)
  tStat[i] <- sqrt(n) * (sampMean - trueMean) / sampSD
}
pVals <- 2 * (1 - pt(abs(tStat), n - 1))
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "3"
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
  sampMean <- mean(data)
  sampSD <- sd(data)
  tStat[i] <- sqrt(n) * (sampMean - trueMean) / sampSD
}
pVals <- 2 * (1 - pt(abs(tStat), n - 1))
mean(pVals < 0.05)
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "6-7"
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
  sampMean <- mean(data)
  sampSD <- sd(data)
  tStat[i] <- sqrt(n) * (sampMean - trueMean) / sampSD
}
pVals <- 2 * (1 - pt(abs(tStat), n - 1))
mean(pVals < 0.05)
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| code-line-numbers: "13"
# Homework 2 Question 4 Simulation Setup
nsim <- 10000
n <- 25
tStat <- rep(0, nsim)
for (i in 1:nsim) {
  data <- rgamma(n, shape = 3, scale = 3)
  trueMean <- 9
  sampMean <- mean(data)
  sampSD <- sd(data)
  tStat[i] <- sqrt(n) * (sampMean - trueMean) / sampSD
}
pVals <- 2 * (1 - pt(abs(tStat), n - 1))
mean(pVals < 0.05)
```

## Homework 2 Example {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: TRUE
#| output-location: fragment
# Our p-values should be Unif(0, 1) if exact
hist(pVals, 
     breaks = seq(0, 1, 0.05),
     col = "purple",
     border = "black",
     main = "Histogram of p-values \n from Gamma(3, 3) Distribution",
     xlab = "p-values")
```