---
title: "(ST 352) Lab 9 <br> The Chi-Square Test of Association"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## Objectives {transition="zoom" transition-speed="slow"}
- Use R to perform a Test of Association for two categorical variables using the Chi-square methods

## Today's Example {.smaller}
Using the data collected from the “memory” activity in which you tried to remember as many items as possible that were displayed for 30 seconds, we will investigate if an association exists between the type of person students identified as (morning or evening) and whether or not at least 60% of the items were remembered.

- `pctremember`
  - "yes" if person remembered at least 60% of the items
  - "no" if a person remembered less than 60% of the items
- `type`
  - "evening" for a self-proclaimed `evening' person
  - "morning" for a self-proclaimed `morning' person

## Today's Data {transition="zoom" transition-speed="slow"}
```{r}
memorygame <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Spring 2024/ST 352 (TA)/Lab 9/memorygame.txt")
memorygame
```

## Hypotheses Defined (Notation) {.smaller}
$$
\begin{align*}
\text{Notation:} &  \\[0.25cm]
  & H_{0}: p_{\text{evening}} = p_{\text{morning}} \quad\quad\Leftrightarrow\quad\quad H_{0}: p_{\text{evening}} - p_{\text{morning}} = 0 \\[0.25cm]
  & H_{A}: p_{\text{evening}} \neq p_{\text{morning}} \quad\quad\Leftrightarrow\quad\quad H_{A}: p_{\text{evening}} - p_{\text{morning}} \neq 0 \\[0.25cm]
\end{align*}
$$

- Where $p_{\text{evening}}$ is the proportion of evening people who remembered at least 60% of the items
- Where $p_{\text{morning}}$ is the proportion of morning people who remembered at least 60% of the items

## Hypotheses Defined (In Words) {.smaller}
- Null Hypothesis $H_{0}$
  - There is no association between type of person (morning or evening) and whether or not they remembered at least 60% of the items
  - There is no difference in proportions between evening and morning people remembering at least 60% of the items
- Alternative Hypothesis $H_{A}$
  - There is an association between type of person (morning or evening) and whether or not they remembered at least 60% of the items
  - There is a difference in the proportions between evening and morning people remembering at least 60% of the items

:::aside
Notice, either of each hypothesis (or similar variant) can be used. You don't need two null and two alternative hypothesis in words.
:::


## Table of Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# To make your life easier I would suggest putting the response variable first
memory.tbl <- table(memorygame$pctremember, memorygame$type)
```

## Table of Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# To make your life easier I would suggest putting the response variable first
memory.tbl <- table(memorygame$pctremember, memorygame$type)
memory.tbl
```

## Table of Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
addmargins(memory.tbl)
```

## Proportions Table {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
prop.tbl <- prop.table(memory.tbl, margin = 2)
```

## Proportions Table {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
prop.tbl <- prop.table(memory.tbl, margin = 2)
prop.tbl
```

## Proportions Table {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
round(prop.tbl, 3)
```

## [Example Format for Question (2) Table]{.r-fit-text}
|       	| Evening     	| Morning     	| Total 	|
|-------	|-------------	|-------------	|-------	|
| No    	| 72 (0.6154) 	| 36 (0.6429) 	| 108   	|
| Yes   	| 45 (0.3846) 	| 20 (0.3571) 	| 65    	|
| Total 	| 117         	| 56          	| 173   	|


:::aside
Question 2 can be answered at this time
:::

## "Expected" Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
memory.test <- chisq.test(memory.tbl)
```

## "Expected" Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
memory.test <- chisq.test(memory.tbl)
memory.test
```

## "Expected" Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
names(memory.test)
```

## "Expected" Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
memory.test[1:length(memory.test)]
```

## "Expected" Counts {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
memory.test$expected
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl)
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE)
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"))
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"),
        main = "Memory Game Activity",
        xlab = "Type of Person",
        ylab = "Proportion")
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"),
        main = "Memory Game Activity",
        xlab = "Type of Person",
        ylab = "Proportion",
        ylim = c(0,1))
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"),
        main = "Memory Game Activity",
        xlab = "Type of Person",
        ylab = "Proportion",
        ylim = c(0,1))
legend("topright", legend = c("Less than 60%", "At least 60%"))
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"),
        main = "Memory Game Activity",
        xlab = "Type of Person",
        ylab = "Proportion",
        ylim = c(0,1))
legend("topright", legend = c("Less than 60%", "At least 60%"),
       fill = c("springgreen2", "plum"))
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"),
        main = "Memory Game Activity",
        xlab = "Type of Person",
        ylab = "Proportion",
        ylim = c(0,1))
legend("topright", legend = c("Less than 60%", "At least 60%"),
       fill = c("springgreen2", "plum"),
       title = "Percentage (%) Remembered")
```

## Side-by-Side Bar Chart {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
barplot(prop.tbl, beside = TRUE,
        col = c("springgreen2", "plum"),
        main = "Memory Game Activity",
        xlab = "Type of Person",
        ylab = "Proportion",
        ylim = c(0,1))
legend("topright", legend = c("Less than 60%", "At least 60%"),
       fill = c("springgreen2", "plum"),
       title = "Percentage (%) Remembered")
```

:::aside
Question 3 can be answered at this time
:::

## [Extracting $p$-value from Chi-Squared Test]{.r-fit-text}{auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
### Multiple ways to see the p-value
# Printing out default object output
memory.test
```

## [Extracting $p$-value from Chi-Squared Test]{.r-fit-text}{auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Printing only p-value
memory.test$p.value
```

## [Extracting $p$-value from Chi-Squared Test]{.r-fit-text}{auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Re-running chi-squared test again
chisq.test(memory.tbl)
```

## [Extracting $p$-value from Chi-Squared Test]{.r-fit-text}{auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Printing out default object output
memory.test

# Printing only p-value
memory.test$p.value

# Re-running chi-squared test again
chisq.test(memory.tbl)
```

:::aside
All lab questions can be answered at this time
:::