---
title: "(ST 352) Lab 3 - Inference using the Two-Sample t-methods"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    chalkboard: TRUE
---
## Objectives {transition="zoom" transition-speed="slow"}
- Recognize when the two-sample methods can be used for inference
- Recognize when the pooled two-sample t-methods can be used.
- Use R to perform a hypothesis test and construct a confidence interval for the difference in population means using the two-sample t-methods (including pooled two-sample t-methods)
- Interpret the output from the two-sample t-methods.

## Today's Example {transition="zoom" transition-speed="slow"}
**Question:** Is there a difference in the average amount of time spent exercising each day between OSU students who have a job and OSU students who are not employed?

## Data {transition="zoom" transition-speed="slow"}
- The `studentsurvey` data set on Canvas are the results of a survey of ST 352 students several years ago. The first two variables in the data set are
  - `work` (employed or notemployed (which is a one-word name that is used for students who were not employed))
  - `exercise` (daily exercise in minutes)
  
```{r}
#| echo: FALSE
studentsurvey <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Spring 2024/ST 352 (TA)/Lab 3/studentsurvey.txt")
```

## Data {transition="zoom" transition-speed="slow"}
```{r}
#| echo: TRUE
studentsurvey
```

## Data {transition="zoom" transition-speed="slow"}
```{r}
#| echo: TRUE
studentsurvey[,1:2]
```

## [Usual Preliminary Steps]{.r-fit-text} {transition="zoom" transition-speed="slow"}
- Step 1: Identify the variable of interest and population
- Step 2: Assess whether the samples are representative of their respective populations
- Step 3: Understand why a particular example is an estimation example or hypothesis test example
- Step 4: Define null and alternative hypotheses

## Defining Hypotheses
- $H_{0}: \mu_{\text{employed}} - \mu_{\text{notemployed}} = 0$
- $H_{A}: \mu_{\text{employed}} - \mu_{\text{notemployed}} \neq 0$
  - $\mu_{\text{employed}}$ is the mean time spent exercising each day for employed OSU students
  - $\mu_{\text{notemployed}}$ is the mean time spent exercising each day for unemployed (not employed) OSU students

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(exercise ~ work, data = studentsurvey)
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE)
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between Employed and Non-Employed Students")
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between 
        Employed and Non-Employed Students")
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students")
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status")
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: fragment
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status",
        cex.main = 0.8) # Title font size
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: fragment
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status",
        cex.main = 0.8) 
axis(2, at = c(1,2), labels = c("Employed", "Not Employed")) # Adds custom y axis labels
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: fragment
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status",
        cex.main = 0.8,
        yaxt = "n") # Removes default y labels
axis(2, at = c(1,2), labels = c("Employed", "Not Employed"))
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: fragment
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status",
        cex.main = 0.8,
        yaxt = "n")
axis(2, at = c(1,2), labels = c("Employed", "Not Employed"),
     las = 2) # Turns labels to be horizontal
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: fragment
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status",
        cex.main = 0.8,
        yaxt = "n")
axis(2, at = c(1,2), labels = c("Employed", "Not Employed"),
     las = 2,
     cex.axis = 0.6) # Shrinks y label font size
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: fragment
boxplot(exercise ~ work, data = studentsurvey, horizontal = TRUE,
        main = "Comparison of Time Spent Exercising Between \n Employed and Non-Employed Students",
        xlab = "Time Spent Exercising per day (minutes)",
        ylab = "Work Status",
        cex.main = 0.8,
        yaxt = "n",
        col = "lawngreen")
axis(2, at = c(1,2), labels = c("Employed", "Not Employed"),
     las = 2,
     cex.axis = 0.6)
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: slide
library(ggplot2)
ggplot(data = studentsurvey) +
  geom_boxplot(aes(x = work, y = exercise, fill = work), show.legend = FALSE) +
  coord_flip() +
  labs(y = "Time Spent Exercising per day (minutes)", x = "Work Status",
       title = "Comparison of Time Spent Exercising Between Employed and Non-Employed Students",
       caption = "Another visualization of the same plot using 'ggplot' package") +
  scale_x_discrete(limits = c("notemployed", "employed"), labels = c("Not Employed", "Employed")) +
  theme_bw() +
  theme(plot.caption = element_text(color = "purple", face = "italic", hjust = 0),
        panel.grid.major.x = element_blank(), # Remove vertical lines
        panel.grid.minor.x = element_blank(), # Remove vertical lines
        axis.title.y = element_text(hjust = 0.5, vjust = -5), 
        axis.title.x = element_text(hjust = 0.5, vjust = -0.25), # Slightly move down x axis label
        axis.title = element_text(size = 12.5)) # Increase x and y titles
```

## Data Exploration {auto-animate=true}
```{r}
#| echo: TRUE
#| output-location: slide
ggplot(data = studentsurvey) +
  geom_violin(aes(x = work, y = exercise, fill = work), show.legend = FALSE) +
  coord_flip() +
  labs(y = "Time Spent Exercising per day (minutes)", x = "Work Status",
       title = "Comparison of Time Spent Exercising Between Employed and Non-Employed Students",
       caption = "Another visualization of the same plot using 'ggplot' package") +
  scale_x_discrete(limits = c("notemployed", "employed"), labels = c("Not Employed", "Employed")) +
  theme_bw() +
  theme(plot.caption = element_text(color = "purple", face = "italic", hjust = 0),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.title.y = element_text(hjust = 0.5, vjust = -5), 
        axis.title.x = element_text(hjust = 0.5, vjust = -0.25),
        axis.title = element_text(size = 12.5))
```

## Data Exploration - Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
library(mosaic)
favstats(exercise ~ work, data = studentsurvey)
```

## Data Exploration - Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative Code
summary(studentsurvey$exercise[studentsurvey$work == "employed"])
summary(studentsurvey$exercise[studentsurvey$work == "notemployed"])
```

## Data Exploration - Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative Code
summary(studentsurvey$exercise[studentsurvey$work == "employed"])
summary(studentsurvey$exercise[studentsurvey$work == "notemployed"])

length(studentsurvey$exercise[studentsurvey$work == "employed"])
length(studentsurvey$exercise[studentsurvey$work == "notemployed"])
```

## Data Exploration - Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative Code
summary(studentsurvey$exercise[studentsurvey$work == "employed"])
summary(studentsurvey$exercise[studentsurvey$work == "notemployed"])

length(studentsurvey$exercise[studentsurvey$work == "employed"])
length(studentsurvey$exercise[studentsurvey$work == "notemployed"])

sd(studentsurvey$exercise[studentsurvey$work == "employed"])
sd(studentsurvey$exercise[studentsurvey$work == "notemployed"])
```

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t_{U}(\mu_{0}) & = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}} \\[0.25cm]
\end{aligned}
$$

:::aside
$\bar{X}$ and $\bar{Y}$ are sample mean estimations for $\mu_{X}$ and $\mu_{Y}$ respectively
:::

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t_{U}(\mu_{0}) & = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}} \\[0.25cm]
\text{We know,} \; \mu_{0} & = 0, \; \text{therefore:} \\[0.25cm]
\end{aligned}
$$

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t_{U}(\mu_{0}) = \dfrac{(\bar{X} - \bar{Y})}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}}
\end{aligned}
$$

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
**Welch-Satterthwaite Approximation**: to the null distribution of $t(\mu_{0})$
$$
\nu = \dfrac{\left(\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}\right)^{2}}{\dfrac{s^{4}_{X}}{m^{2}(m-1)} + \dfrac{s^{4}_{Y}}{n^{2}(n-1)}}
$$
Note: $\min(m-1, n-1) \le \nu \le m+n-2$

:::aside
Note: $\nu$ are the degrees of freedom
:::

## [Behind the Scenes of 't.test' function (Unequal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
**Confidence Interval:**
$$
(\bar{X} - \bar{Y}) \pm t_{(\nu)(1-\frac{\alpha}{2})}\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}
$$

:::aside
Remember, unless specified otherwise, R will make group 1 the one that comes alphabetically first (`employed` in our example)
:::

## [Behind the Scenes of 't.test' function (Equal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
$$
t_{E}(\mu_{0}) = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s_{p}^{2}}{m} + \dfrac{s_{p}^{2}}{n}}} = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{s_{p}^{2}\left(\dfrac{1}{m} + \dfrac{1}{n}\right)}}
$$

## [Behind the Scenes of 't.test' function (Equal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
$$
\begin{align*}
s_{p}^{2} & = \dfrac{\sum\limits_{i=1}^{m}(X_{i} - \bar{X})^{2} + \sum\limits_{i=1}^{n}(Y_{i} - \bar{Y})^{2}}{(m - 1) + (n - 1)} \\[0.25cm]
& = \dfrac{(m-1)s^{2}_{X} + (n-1)s^{2}_{Y}}{(m+n-2)}
\end{align*}
$$

:::aside
We use the pooled methods if we assume the two groups have equal variance (equal standard deviations)
:::

## [Behind the Scenes of 't.test' function (Equal Variance)]{.r-fit-text} {transition="slide" transition-speed="slow"}
$$
(\bar{X} - \bar{Y}) \pm t_{(m+n-2)(1-\frac{\alpha}{2})}\sqrt{s_{p}^{2}\left(\dfrac{1}{m} + \dfrac{1}{n}\right)}
$$

:::aside
Notice the degrees of freedom = $m + n - 2$
:::

## [Behind the Scenes of 't.test' function (Summary)]{.r-fit-text} {transition="slide" transition-speed="slow"}

- **Unequal Variance (un-pooled)**
  - No assumption of equal variance
  - Degrees of freedom calculated with complicated formula (R will do this for you)
  - `var.equal = FALSE` (R's default)
- **Equal Variance (pooled)**
  - Assumption of equal variance
  - Degrees of freedom: $m + n - 2$ 
    - = Group 1 sample size + Group 2 sample size - 2
  - `var.equal = TRUE`

## 't.test' time (Unequal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
#| output: FALSE
# Unequal Variance (non-pooled) code
t.test(exercise ~ work, data = studentsurvey)
```

## 't.test' time (Unequal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
#| output: FALSE
# Unequal Variance (non-pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0)
```

## 't.test' time (Unequal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
#| output: FALSE
# Unequal Variance (non-pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = FALSE)
```

## 't.test' time (Unequal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
#| output: FALSE
# Unequal Variance (non-pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = FALSE,
       alternative = "two.sided")
```

## 't.test' time (Unequal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
# Unequal Variance (non-pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = FALSE,
       alternative = "two.sided",
       conf.level = 0.95)
```

## 't.test' time (Equal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
# Equal Variance (pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = TRUE,
       alternative = "two.sided",
       conf.level = 0.95)
```

## 't.test' time (Equal Variance) {transition="slide" transition-speed="slow"}
```{r}
#| echo: TRUE
#| code-line-numbers: "4"
# Equal Variance (pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = TRUE,
       alternative = "two.sided",
       conf.level = 0.95)
```

## 't.test' time (Unequal vs. Equal) {transition="slide" transition-speed="slow" .smaller}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
# Unequal Variance (non-pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = FALSE,
       alternative = "two.sided",
       conf.level = 0.95)
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
# Equal Variance (pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = TRUE,
       alternative = "two.sided",
       conf.level = 0.95)
```
:::

::::

:::aside
Notice we have our confidence intervals given here too
:::

## [When to use which t.test (pooled vs. unpooled)]{.r-fit-text}
Standard deviation rule (Learning Path 3, part 2 lecture notes)

$$
\begin{align*}
\frac{s_{1}}{s_{2}} < 1.5 \hspace{2cm} \text{where } s_{1} > s_{2}
\end{align*}
$$
Less 'mathy' version:

$$
\frac{\text{larger standard deviation}}{\text{smaller standard deviation}} < 1.5 \hspace{0.5cm}(\text{or so})
$$

:::aside
This is not the only criteria one can use, but it is what is recommended by your professor
:::

## [When to use which t.test (pooled vs. unpooled)]{.r-fit-text}
```{r}
#| echo: TRUE
favstats(exercise ~ work, data = studentsurvey)

# Alternative code
sd(studentsurvey$exercise[studentsurvey$work == "employed"])
sd(studentsurvey$exercise[studentsurvey$work == "notemployed"])

# Ratio of largest over smallest standard deviation
sd.ratio <- sd(studentsurvey$exercise[studentsurvey$work == "notemployed"]) / sd(studentsurvey$exercise[studentsurvey$work == "employed"])
sd.ratio

# Is this ratio less than 1.5?
sd.ratio < 1.5
```

## [t.test of choice (based on sd rule)]{.r-fit-text}
```{r}
#| echo: TRUE
# Equal Variance (pooled) code
t.test(exercise ~ work, data = studentsurvey,
       mu = 0,
       var.equal = TRUE,
       alternative = "two.sided",
       conf.level = 0.95)
```

## Answer Question of Interest {.smaller}
- There is not enough evidence to indicate the mean time spent exercising each day for employed OSU students is different than the mean time spent exercising each day for unemployed (not employed) OSU students (p-value = 0.5974).
- Confidence interval for $\mu_{\text{employed}} - \mu_{\text{notemployed}} = (-12.03719 \text{ minutes}, \, 20.81971 \text{ minutes})$
- We’re 95% confident that the mean time spent exercising each day for employed OSU students is between 12.04 minutes *less* to 21.82 minutes *more* than the mean time spent exercising each day for unemployed (not employed) OSU students.
  - We’re 95% confident that the mean time spent exercising each day is between 12.04 minutes *less* to 21.82 minutes *more* for OSU students who work compared to those who do not work.

:::aside
Our p-value agrees with our confidence interval (since the C.I. contains 0)
:::

## I'll stop talking (for now)
- Let me know if you have any questions
- I'll provide some hints a little later

## Lab Hints :)
```{r}
#| echo: TRUE
facebook <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Spring 2024/ST 352 (TA)/Lab 3/facebook.txt")

# Structure of data
str(facebook)

# Sample Standard Deviations
sd(facebook$GPA[facebook$user == "Yes"])
sd(facebook$GPA[facebook$user == "No"])

# Ratio
sd(facebook$GPA[facebook$user == "Yes"]) / sd(facebook$GPA[facebook$user == "No"])

# Pooled t-test
t.test(GPA ~ user, data = facebook,
       mu = 0,
       var.equal = TRUE,
       alternative = "less")

# Copying data set for modification
facebook.2 <- facebook

# Structure of data
str(facebook)

# Re-leveling the groups in user
facebook.2$user <- factor(facebook.2$user, levels = c("Yes", "No"), ordered = TRUE)

# What should we expected to happen
t.test(GPA ~ user, data = facebook.2,
       mu = 0,
       var.equal = TRUE,
       alternative = "less")

```

