---
title: "(ST 352) Lab 5 - Multiple Linear Regression Analysis"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    chalkboard: TRUE
---

## Objectives {transition="zoom" transition-speed="slow"}
- Using R, obtain and interpret appropriate graphical displays to assess conditions for inference in multiple linear regression. 
- Using R, obtain output from a Multiple Linear Regression analysis that best satisfies the conditions.
- Use R to obtain a predicted value and a prediction interval. Interpret the prediction interval in the context of the problem

## Today's Example
Researchers were interested in what characteristics of a person may explain their body fat (measured as a percentage). Factors researchers thought might be significant predictors of percent body fat were age, height, weight, and various body measurements (such as neck, abdomen, knee, and ankle circumference, just to name a few). In this example, in addition to age, height, and weight as explanatory variables, we will use just one of the body measurements (chest circumference) as our last explanatory variable. A full data set (not available for your use) contains these variables and other body measurements. The `bodyfat` data set on Canvas only contains the following variables:

## Today's Data {transition="zoom" transition-speed="slow"}
```{r}
bodyfat <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Spring 2024/ST 352 (TA)/Lab 5/bodyfat.txt")
bodyfat
```

## Goal
Use the `bodyfat` data set to perform a multiple linear regression analysis to determine which (if any) of age, height, weight, and/or chest help explain percent body fat.

:::aside
This is the same dataset that is used in your lab assignment this week
:::

## [Step 1: Is the sample representative of the population?]{.r-fit-text}
- Who knows

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text}
- Obtain a scatterplot matrix to assess the correlation between explanatory variables
- Obtain a correlation matrix showing the correlation coefficients between pairs of
explanatory variables

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19)
```


## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables")
```


## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables",
      # Labels are applied left to right, should match first input argument order
      labels = c("Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"))
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables",
      labels = c("Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables",
      labels = c("Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL,
      gap = 2)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables",
      labels = c("Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL,
      gap = 0)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables",
      labels = c("Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL,
      gap = 1,
      font.labels = 2)
```

## Slight Aside
- `1` Default Font
  - `2` Bold
  - `3` Italicized
  - `4` Bold and Italicized
  - `5` Greek Letters
- `6` Times New Roman Font
  - `7` Bold Times New Roman
  - `8` Italicized Times New Roman
  - `9` Bold and Italicized Times New Roman
- `10` Typewriter Font
  - `11` Bold Typewriter 
  - `12` Italicized Typewriter 
  - `13` Bold and Italicized Typewriter
- Goes up to 20 different `font.labels` in total (everything past 20 is the same)

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Explanatory Variables",
      labels = c("Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL,
      gap = 0.75,
      font.labels = 2)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- Uses for explanatory variable scatterplot matrix
  - Checking for high correlation between explanatory variables
  - Checking for a linear relationship between response variable and each explanatory variable
  - Checking for outliers
- What do we see?

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
cor(bodyfat)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
round(cor(bodyfat), digits = 4)
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
cor(bodyfat)[2:5,2:5]

# Accomplishes the same as the above
cor(bodyfat)[-1,-1]
```

## [Step 2 - Assessing correlation among the explanatory variables]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- Uses for correlation matrix
- Gives the correlation coefficient between each pair of explanatory variables. This can be used in
conjunction with the scatterplot matrix to check for high correlation between explanatory
variables.

:::aside
Take a minute to answer question 1 on the lab assignment
:::

## [Strategy in determining which of two highly correlated variables to remove]{.r-fit-text}
- Generate a linear model with just the two highly correlated explanatory variables. 
- Look at the p-values from the t-tests – the one with the highest p-value would be removed as it would be deemed the lesser significant variable
  - Note that the one with the t-statistic closer to 0 will have the higher p-value.
  
## [Strategy in determining which of two highly correlated variables to remove]{.r-fit-text}
```{r}
lmod <- lm(fat ~ weight + chest, data = bodyfat)
summary(lmod)
```

:::aside
Take a minute to answer question 2 on the lab assignment
:::

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
- Scatterplot matrix
  - We need to generate a new scatterplot matrix with the response variable included

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
pairs(fat ~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Variables",
      labels = c("Body Fat Percentage", "Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL,
      gap = 0.75,
      font.labels = 2)
```

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Create model (remember to exclude variable that was take out earlier)
lmod2 <- lm(fat ~ age + height + chest, data = bodyfat)

# Construct residual plot
plot(lmod2$residuals ~ lmod2$fitted)
```

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Construct residual plot
plot(lmod2$residuals ~ lmod2$fitted,
     pch = 19,
     col = "pink")
```

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot(lmod2$residuals ~ lmod2$fitted,
     pch = 19,
     col = "pink",
     main = "Residual vs. Predicted Values",
     ylab = "Residuals",
     xlab = "Predicted Body Fat %")
```

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
plot(lmod2$residuals ~ lmod2$fitted,
     pch = 19,
     col = "pink",
     main = "Residual vs. Predicted Values",
     ylab = "Residuals",
     xlab = "Predicted Body Fat %")
abline(h = 0)
```

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Residual plot (residuals vs. explanatory variable)
plot(lmod2$residuals ~ bodyfat$age,
     pch = 19,
     col = "orange",
     main = "Residual vs. Age",
     ylab = "Residuals",
     xlab = "Age (years)")
abline(h = 0)
```

## Step 3 - Checking for outliers {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| code-line-numbers: "2"
#| output-location: fragment
# Residual plot (residuals vs. explanatory variable)
plot(lmod2$residuals ~ bodyfat$age,
     pch = 19,
     col = "orange",
     main = "Residual vs. Age",
     ylab = "Residuals",
     xlab = "Age (years)")
abline(h = 0)
```

## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- Linearity condition
  - Scatterplot matrix (pairs plot)
  - Residual plot
- Constant variance
  - Residual plot
- Normality condition
  - Normal probability plot of residuals
  
## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
pairs(fat ~ age + weight + height + chest, data = bodyfat,
      col = "lawngreen",
      pch = 19,
      main = "Body Fat Variables",
      labels = c("Body Fat Percentage", "Age (years)", "Weight (lbs)", "Height (inches)", "Chest Circumference (cm)"),
      lower.panel = NULL,
      gap = 0.75,
      font.labels = 2)
```

## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
plot(lmod2$residuals ~ lmod2$fitted,
     pch = 19,
     col = "pink",
     main = "Residual vs. Predicted Values",
     ylab = "Residuals",
     xlab = "Predicted Body Fat %")
abline(h = 0)
```

## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qqnorm(lmod2$residuals)
```

## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qqnorm(lmod2$residuals,
       pch = 19, 
       col = "purple",
       main = "Normal Probability Plot of Residuals")
```

## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
qqnorm(lmod2$residuals,
       pch = 19, 
       col = "purple",
       main = "Normal Probability Plot of Residuals")
qqline(lmod2$residuals)
```

## [Step 4 - Assess the linearity, constant variation, and normality conditions]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
qqnorm(lmod2$residuals,
       pch = 19, 
       col = "purple",
       main = "Normal Probability Plot of Residuals")
qqline(lmod2$residuals)
```

:::aside
Take a couple minutes to answer questions 3 through 6 on the lab assignment
:::

## [Steps 5-7 - Analysis - Use your model to make predictions and obtain a prediction interval]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output-location: fragment
# Previously created model
lmod2 <- lm(fat ~ age + height + chest, data = bodyfat)
summary(lmod2)
```

## [Steps 5-7 - Analysis - Use your model to make predictions and obtain a prediction interval]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| output: FALSE
predict(lmod2, data.frame(age = 30, height = 72, chest = 105))
```

## [Steps 5-7 - Analysis - Use your model to make predictions and obtain a prediction interval]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
predict(lmod2, data.frame(age = 30, height = 72, chest = 105),
        interval = "prediction")
```

## [Steps 5-7 - Analysis - Use your model to make predictions and obtain a prediction interval]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
predict(lmod2, data.frame(age = 30, height = 72, chest = 105),
        interval = "prediction",
        level = 0.95)
```

