---
title: "(ST 352) Lab 2 - Paired t-methods"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    chalkboard: TRUE
---
## Office Hours {transition="zoom" transition-speed="slow"}
When are my office hours?

- Wednesday, 10:00am - 11:00am
- Friday, 8:00am - 9:00am
  - Why are they so early?
    - All the midday slots were taken :(
- Where?
  - Weniger, Room 205 
- Why would you want to stop by?
  - Questions on lab assignments
  - Questions regarding Acrobatiq content
  - To share a funny joke
  - Questions about R
  - Any miscellaneous statistical queries
  - Etc.

## The Q-Q Plot {transition="concave" transition-speed="slow"}
- What is a QQ plot
  - A graphical method used to compare two probability distributions via their quantiles.
  - What we often see in practice is our sample (empirical probability distribution) against the theoretical (often Normal probability distribution)
  - If the two are similar we expect points to land on and / or close to the reference line
- Why do we construct them?
  - We typically want to verify if a normality assumption is reasonable.

## The Q-Q Plot {transition="concave" transition-speed="slow"}
- How does one tell if the QQ plot suggests normality or not
  - If the points land on and / or around the reference line
- What if we have a large sample size?
  - The more skewed our data is, the larger the sample size needed to reach approximate normality.
- How do I typically approach these QQ plot interpretations?
  - One can simulate normal data of equal sample size and construct a QQ plot

## The Q-Q Plot {transition="concave" transition-speed="slow"}
```{r}
#| echo: FALSE
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Spring 2024/ST 352 (TA)/Lab 2")
television <- read.delim("television.txt")
```

```{r}
#| echo: FALSE
par(mfrow = c(1, 2))
# Simulated Normal Data
Normal.Data <- rnorm(183)
qqnorm(Normal.Data,
       main = "Simulated Normal Data",
       pch = 19)
qqline(Normal.Data)

# Lab 1 Data
qqnorm(scale(television$amount),
       main = "Lab Assignment 1 Data",
       pch = 19,
       col = "lawngreen")
qqline(scale(television$amount))

```

## The Q-Q Plot {transition="concave" transition-speed="slow"}
Notice this is the same thing but displayed row-wise instead of column-wise
```{r}
#| echo: FALSE
par(mfrow = c(2, 1))
# Simulated Normal Data
Normal.Data <- rnorm(183)
qqnorm(Normal.Data,
       main = "Simulated Normal Data",
       pch = 19)
qqline(Normal.Data)

# Lab 1 Data
qqnorm(scale(television$amount),
       main = "Lab Assignment 1 Data",
       pch = 19,
       col = "lawngreen")
qqline(scale(television$amount))
par(mfrow = c(1, 1))
```

## The Q-Q Plot {transition="concave" transition-speed="slow"}
Notice how depending on the plot window size, one can draw differing conclusions.

- I recommend creating a large window to view plots to avoid unintended distortions.

## Objectives {transition="zoom" transition-speed="slow"}
- Recognize when the paired methods can be used
- Use R to perform a hypothesis test and construct a confidence interval for the population mean of the differences using the paired t-methods
- Interpret the output from analysis

## Today's Example {transition="zoom" transition-speed="slow"}
**Scenario:**
Weights of 9 students (in pounds) before and after a month-long training schedule are given below. 

**Question:**
Is there evidence to indicate a mean weight loss after this month-long training program?

## Loading in Data
```{r}
#| echo: TRUE
training <- read.delim("training.txt")
```

## Why is this a matched-pairs design?
- Because...
  - We are interested in a difference between two variables of the same observation / subject.

## Obtaining Difference {auto-animate=true}
```{r}
#| echo: TRUE
training$change <- training$before - training$after
```

## Obtaining Difference {auto-animate=true}
```{r}
#| echo: TRUE
training$change <- training$before - training$after
training
```

## Obtaining Difference {auto-animate=true}
```{r}
#| echo: TRUE
# To match the lab notes
change <- training$before - training$after
```

## [Usual Preliminary Steps]{.r-fit-text} {transition="zoom" transition-speed="slow"}
- Step 1: Identify the variable of interest and population
- Step 2: Assess whether the samples are representative of their respective populations
- Step 3: Understand why a particular example is an estimation example or hypothesis test example
- Step 4: Define null and alternative hypotheses

## Defining Hypothesis
- $H_{0}: \mu_{\text{change}} = 0$
- $H_{A}: \mu_{\text{change}} > 0$
  - Where $\mu_{\text{change}}$ is the mean difference between start and end weight for student partaking in the month-long training program
  - Where change = before - after

:::aside
Make sure on lab assignment question 2, to define how you subtracted to obtain the difference
:::

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
require(mosaic) # library() and require() are essentially the same
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
dotPlot(change)
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
dotPlot(change,
        main = "Change in Weight")
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
dotPlot(change,
        main = "Change in Weight",
        xlab = "Weight Change in lbs (before - after)")
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
dotPlot(change,
        main = "Change in Weight",
        xlab = "Weight Change in lbs (before - after)",
        col = "pink")
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
hist(change)
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
hist(change,
     main = "Change in Weight")
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
hist(change,
     main = "Change in Weight",
     xlab = "Weight Change in lbs (before - after)")
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
hist(change,
     main = "Change in Weight",
     xlab = "Weight Change in lbs (before - after)",
     col = "pink")
```

## Exploring Data {auto-animate=true .smaller}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
# Using 'mosaic' package
dotPlot(change,
        main = "Change in Weight",
        xlab = "Weight Change in lbs (before - after)",
        col = "pink")
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
# No packages needed
hist(change,
     main = "Change in Weight",
     xlab = "Weight Change in lbs (before - after)",
     col = "pink")
```
:::

::::

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(change)
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(change,
       pch = 19)
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(change,
       pch = 19,
       col = "lawngreen")
```

## Exploring Data {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(change,
       pch = 19,
       col = "lawngreen")
qqline(change)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
favstats(change) # Within 'mosaic' package
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
# No package needed
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
summary(change)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
summary(change)
length(change)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
summary(change)
length(change)
sd(change)
```

## Descriptive Statistics {auto-animate=true .smaller}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
favstats(change)
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
# Alternative code
summary(change)
length(change)
sd(change)
```
:::

::::

## Hypotheses
- $H_{0}: \mu_{\text{change}} = 0$
- $H_{A}: \mu_{\text{change}} > 0$
  - Where $\mu_{\text{change}}$ is the mean difference between start and end weight for student partaking in the month-long training program
  - Where change = before - after

:::aside
Same reminder: make sure on lab assignment question 2, to define how you subtracted to obtain the difference
:::

## Paired t-test (Option 1) {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after)
```

## Paired t-test (Option 1) {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after,
       paired = TRUE)
```

## Paired t-test (Option 1) {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0)
```

## Paired t-test (Option 1) {auto-animate=true}
```{r}
#| echo: TRUE
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0,
       alternative = "greater")
```

## Paired t-test (Option 2) {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(change)
```

## Paired t-test (Option 2) {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(change,
       mu = 0)
```

## Paired t-test (Option 2) {auto-animate=true}
```{r}
#| echo: TRUE
t.test(change,
       mu = 0,
       alternative = "greater")
```

## [Paired t-test (Option 1 vs. Option 2)]{.r-fit-text} {auto-animate=true}
:::: {.columns}

::: {.column width="56%"}
```{r}
#| echo: TRUE
# Option 1
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0,
       alternative = "greater")
```
:::

::: {.column width="44%"}
```{r}
#| echo: TRUE
# Option 2
t.test(change,
       mu = 0,
       alternative = "greater")
```
:::

::::

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after)
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after,
       paired = TRUE)
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0)
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0,
       alternative = "two.sided")
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.95)
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
#| code-line-numbers: "4"
t.test(training$before, training$after,
       paired = TRUE,
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.95)
```

## Lab Assignment Hints
- Appropriate graphs to support t-method claim should at least include one of the following:
  - dotplot
  - histogram
  - Q-Q plot
  - box-and-whisker plot of the *differences in prices*
- Some conclusion components to remember to include:
  - Amount of evidence
  - Parameter of interest
  - Variable
  - Populations
  - Alternative hypothesis context