---
title: "(ST 352) Lab 1 - Single Quantitative Variable Inference"
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    chalkboard: TRUE
---

## Objectives {transition="zoom" transition-speed="slow"}
- Review exploratory tools for a quantitative variable
- Use R to perform a hypothesis test and construct a confidence interval for a mean using the one-sample t-methods

## Today's Example {transition="zoom" transition-speed="slow"}
The American Heart Association recommends that adults get an average of 30 minutes of moderate-intensity exercise 5 days a week AND participate in moderate to high intensity muscle-strengthening activities at least 2 days per week. Assuming the moderate to high intensity muscle-strengthening exercises take an average of 30 minutes per day, let's say that adults should exercise more than 30 minutes per day every day.

## Question of Interest {transition="zoom" transition-speed="slow"}
- **Question:** Do college students exercise more than 30 minutes per day, on average?
- **Sample Data:** A sample of 34 OSU students from the ST 352 class several years ago was taken. Their daily amount of exercise (in minutes) is recorded in the `exercise` data set posted on Canvas

## [Some Usual Preliminary Steps]{.r-fit-text} {transition="zoom" transition-speed="slow"}
- Step 1: Identify the variable of interest and population
- Step 2: Assess whether the sample is representative of the population
- Step 3: Understand why a particular example is an estimation example or hypothesis test example
- Step 4: Define null and alternative hypotheses

## Defining Hypothesis
- $H_{0}: \mu = 30$
- $H_{A}: \mu > 30$
  - Where $\mu$ is the average daily exercise (in minutes) for college students

## Data Importation {transition="zoom" transition-speed="slow"}
```{mermaid}
flowchart TD
A{RStudio} -->|Recommended| B{R Script}
B -->|"Environment Panel (Top Right Panel)"| C1[Import Dataset Button]
--> C2("From Text (base)...") 
--> C3(Wa-La)
B --> D1["File Button (Top Left)"] 
--> D2(Import Dataset) 
--> D3("From Text (base)...") 
--> D4(Ta-Da)
B -->|Directly Coding it| E1[Set Working Directory] 
--> E2("'read.csv' function") 
--> E3(Woo-Hoo)
```

## Loading in Data
```{r}
#| echo: TRUE
#| results: hide
exercise <- read.csv("C:/Users/Ghcto/OneDrive/Desktop/School/Spring 2024/ST 352 (TA)/Lab 1/exercise.txt", sep="")
```

:::aside
If you haven't already, install the 'mosaic' packages
:::

## Viewing Data {auto-animate=true}
```{r}
#| echo: TRUE
head(exercise)
```

## Viewing Data {auto-animate=true}
```{r}
#| echo: TRUE
exercise
```

## Histogram {auto-animate=true}
```{r}
#| echo: TRUE
hist(exercise$time)
```

## Histogram {auto-animate=true}
```{r}
#| echo: TRUE
hist(exercise$time,
     main = "Histogram of Exercise Time")
```

## Histogram {auto-animate=true}
```{r}
#| echo: TRUE
hist(exercise$time,
     main = "Histogram of Exercise Time",
     xlab = "Time (Minutes)")
```

## Histogram {auto-animate=true}
```{r}
#| echo: TRUE
hist(exercise$time,
     main = "Histogram of Exercise Time",
     xlab = "Time (Minutes)",
     col = "pink")
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(exercise$time)
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(exercise$time)
qqline(exercise$time)
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(exercise$time,
       main = "Normal Probability Plot",
       pch = 19)
qqline(exercise$time)
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
qqnorm(exercise$time,
       main = "Normal Probability Plot",
       pch = 19,
       cex = 1.2)
qqline(exercise$time)
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
# One can always spice it up with some color
qqnorm(exercise$time,
       main = "Normal Probability Plot",
       pch = 19,
       cex = 1.2,
       col = rainbow(nrow(exercise)))
qqline(exercise$time)
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: FALSE
# One can always spice it up with some color
qqnorm(exercise$time,
       main = "Normal Probability Plot",
       pch = 19,
       cex = 1.2,
       col = rainbow(nrow(exercise)))
qqline(exercise$time)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
library(mosaic) # library() and require() are essentially the same
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
favstats(exercise$time)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
# No package needed
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
summary(exercise$time)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
summary(exercise$time)
length(exercise$time)
```

## Descriptive Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative code
summary(exercise$time)
length(exercise$time)
sd(exercise$time)
```

## Descriptive Statistics {auto-animate=true .smaller}
:::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: TRUE
favstats(exercise$time)
```
:::

::: {.column width="50%"}
```{r}
#| echo: TRUE
summary(exercise$time)
length(exercise$time)
sd(exercise$time)
```
:::

::::

## Table Example {.smaller}
```{r}
library(knitr)
kable(favstats(exercise$time), 
      format = "pipe",
      col.names = c("Min.", "Q1", "Median", "Q3", "Max.", "Mean", "Std.Dev.", "Sample Size (n)", "Missing"),
      digits = 3,
      align = "c",
      label = "Table 1",
      caption = "Example of Summary Statistic Table")
```

## Hypothesis Test
**Reminder:**

- $H_{0}: \mu = 30$
- $H_{A}: \mu > 30$
  - Where $\mu$ is the average daily exercise (in minutes) for college students

## Hypothesis Test {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(exercise$time)
```

## Hypothesis Test {auto-animate=true}
```{r}
#| echo: TRUE
#| output: FALSE
t.test(exercise$time, alternative = "greater")
```

## Hypothesis Test {auto-animate=true}
```{r}
#| echo: TRUE
t.test(exercise$time, alternative = "greater", mu = 30)
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
t.test(exercise$time, alternative = "greater", mu = 30)
```

:::aside
Notice this has the confidence interval, but it is not two sided, hence the $\infty$ upper bound
:::

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
# Option 1
t.test(exercise$time, alternative = "two.sided", mu = 30)
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
# Option 2
confint(t.test(exercise$time, conf.level = 0.95))
```

## Confidence Interval {auto-animate=true}
```{r}
#| echo: TRUE
# Option 3
t.test(exercise$time, alternative = "two.sided", mu = 30)$conf.int
```
