---
title: "(ST 351) Lab 8 Two-Sample t-test"
format: 
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dark
    transition: slide
    chalkboard: TRUE
---

## [(Unequal Variance) Two-sample t-test Assumptions]{.r-fit-text}
- Independent observations in each sample
- Samples independent of each other
- Normally distributed data **OR** Large sample sizes

:::aside
Also called: Two-sample t-test (Welch-Satterthwaite)
:::

## [(Unequal Variance) Two-sample t-test - Other Conditions]{.r-fit-text}
- Random sample from each population
- Quantitative response variable
- Two-category categorical explanatory variable

## When to **NOT** use the t-test
- If at least one sample size is “small” AND the sample data are not symmetric
- The sample data are heavily skewed and/or there are extreme outliers

## Central Limit Theorem (My Hero) {auto-animate=true}
```{r}
set.seed(11302023)

z.data <- rnorm(100)
t.data.df3 <- rt(30, 3)
t.data.df20 <- rt(30, 20)
t.data.df60 <- rt(30, 60)

curve(dnorm(x, mean = mean(z.data), sd = sd(z.data)),
      main = "Distribution Comparisons",
      xlab = "",
      ylab = "Density",
      from = -3, to = 3,
      ylim = c(0,0.45),
      col = "blue", 
      lwd = 2)
legend("topright", 
       legend = "Normal Distribution", 
       fill = "blue")
```

## Central Limit Theorem (My Hero)
```{r}
curve(dnorm(x, mean = mean(z.data), sd = sd(z.data)),
      main = "Distribution Comparisons",
      xlab = "",
      ylab = "Density",
      from = -3, to = 3,
      ylim = c(0,0.45),
      col = "blue", 
      lwd = 2)
curve(dnorm(x, mean = mean(t.data.df3), sd = sd(t.data.df3)),
      col = "seagreen", 
      lwd = 2,
      add = TRUE)
legend("topright", 
       legend = c("t Distribution (df = 3)", "Normal Distribution"), 
       fill = c("seagreen", "blue"))
```

## Central Limit Theorem (My Hero)
```{r}
curve(dnorm(x, mean = mean(z.data), sd = sd(z.data)),
      main = "Distribution Comparisons",
      xlab = "",
      ylab = "Density",
      from = -3, to = 3,
      ylim = c(0,0.45),
      col = "blue", 
      lwd = 2)
curve(dnorm(x, mean = mean(t.data.df3), sd = sd(t.data.df3)),
      col = "seagreen", 
      lwd = 2,
      add = TRUE)
curve(dnorm(x, mean = mean(t.data.df20), sd = sd(t.data.df20)),
      col = "lawngreen", 
      lwd = 2,
      add = TRUE)
legend("topright", 
       legend = c("t Distribution (df = 3)", "t Distribution (df = 20)", "Normal Distribution"), 
       fill = c("seagreen", "lawngreen", "blue"))
```

## Central Limit Theorem (My Hero)
```{r}
curve(dnorm(x, mean = mean(z.data), sd = sd(z.data)),
      main = "Distribution Comparisons",
      xlab = "",
      ylab = "Density",
      from = -3, to = 3,
      ylim = c(0,0.45),
      col = "blue", 
      lwd = 2)
curve(dnorm(x, mean = mean(t.data.df3), sd = sd(t.data.df3)),
      col = "seagreen", 
      lwd = 2,
      add = TRUE)
curve(dnorm(x, mean = mean(t.data.df20), sd = sd(t.data.df20)),
      col = "lawngreen", 
      lwd = 2,
      add = TRUE)
curve(dnorm(x, mean = mean(t.data.df60), sd = sd(t.data.df60)),
      col = "lightgreen", 
      lwd = 2,
      add = TRUE)
legend("topright", 
       legend = c("t Distribution (df = 3)", "t Distribution (df = 20)", 
                  "t Distribution (df = 60)", "Normal Distribution"), 
       fill = c("seagreen", "lawngreen", "lightgreen", "blue"))
```

## Example
We want to determine if there is a *difference* in the average speed of roller coasters between steel and wooden track roller coasters

:::aside
The coasters data set contains many variables on a sample of wooden track roller coasters and a sample of steel track roller coasters, including speed (speed of roller coaster in miles per hour) and track (wooden or steel).
:::

## [Identify the Variable of Interest and Population of Interest]{.r-fit-text}
- Variable of interest: speed of roller coasters (in mph)
- Populations: roller coasters with wooden tracks and steel tracks

## [Assess Sample Representativeness]{.r-fit-text}
- Doesn't say anything about randomization. It *could* be. 

:::aside
We'll just pretend it is and proceed
:::

## [State Null and Alternative Hypotheses]{.r-fit-text}
- Null Hypothesis:
  - $H_{0}: \mu_{steel} = \mu_{wood} \equiv \mu_{steel} - \mu_{wood} = 0$
    - There is **NOT** a difference in the average speed of roller coasters between steel and wooden track
roller coasters
- Alternative Hypothesis:
  - $H_{A}: \mu_{steel} \neq \mu_{wood} \equiv \mu_{steel} - \mu_{wood} \neq 0$
    - There is a difference in the average speed of roller coasters between steel and wooden track
roller coasters
- Where $\mu_{steel}$ = Population average (mean) speed of steel track roller coasters
- Where $\mu_{wood}$ = Population average (mean) speed of wooden track roller coasters

## Importing & Viewing Data {auto-animate=true}
```{r}
#| echo: TRUE
#| results: hide
# setwd("C:/Users/Ghcto/OneDrive/Desktop/Fall 2023/ST 351 (TA)")
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8")
```
## Importing & Viewing Data {auto-animate=true}
```{r}
#| echo: TRUE
#| results: hide
# setwd("C:/Users/Ghcto/OneDrive/Desktop/Fall 2023/ST 351 (TA)")
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8")

coasters <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8/coasters.txt")
```
## Importing & Viewing Data {auto-animate=true}
```{r}
#| echo: TRUE
#| results: hide
# setwd("C:/Users/Ghcto/OneDrive/Desktop/Fall 2023/ST 351 (TA)")
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8")

coasters <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8/coasters.txt")

head(coasters)
```
## Importing & Viewing Data {auto-animate=true}
```{r}
#| echo: TRUE
# setwd("C:/Users/Ghcto/OneDrive/Desktop/Fall 2023/ST 351 (TA)")
setwd("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8")

coasters <- read.delim("C:/Users/Ghcto/OneDrive/Desktop/School/Fall 2023/ST 351 (TA)/Lab 8/coasters.txt")

head(coasters)
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
#| warning: false
#| message: false
#| results: hide
library(mosaic)
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
#| warning: false
#| message: false
#| results: hide
library(mosaic)
favstats(speed ~ track, data = coasters)
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
#| warning: false
#| message: false
library(mosaic)
favstats(speed ~ track, data = coasters)
```

:::aside
Remember not to just copy and paste your summary statistics results directly from R
:::

:::aside
Units in MPH
:::

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
#| results: hide
# Alternative Code
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative Code
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
#| results: hide
# Alternative Code
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
sd(coasters$speed[coasters$track=="Steel"])
sd(coasters$speed[coasters$track=="Wood"])
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative Code
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
sd(coasters$speed[coasters$track=="Steel"])
sd(coasters$speed[coasters$track=="Wood"])
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
#| results: hide
# Alternative Code
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
sd(coasters$speed[coasters$track=="Steel"])
sd(coasters$speed[coasters$track=="Wood"])
length(coasters$speed[coasters$track=="Steel"])
length(coasters$speed[coasters$track=="Wood"])
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Alternative Code
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
sd(coasters$speed[coasters$track=="Steel"])
sd(coasters$speed[coasters$track=="Wood"])
length(coasters$speed[coasters$track=="Steel"])
length(coasters$speed[coasters$track=="Wood"])
```

## Summary Statistics {auto-animate=true}
```{r}
#| echo: TRUE
# Comparing the two methods
favstats(speed ~ track, data = coasters)
summary(coasters$speed[coasters$track=="Steel"])
summary(coasters$speed[coasters$track=="Wood"])
sd(coasters$speed[coasters$track=="Steel"])
sd(coasters$speed[coasters$track=="Wood"])
length(coasters$speed[coasters$track=="Steel"])
length(coasters$speed[coasters$track=="Wood"])
```

:::aside
Remember not to just copy and paste your summary statistics results directly from R
:::

:::aside
Units in MPH
:::

## [Side-by-Side Box-and-Whisker Plots]{.r-fit-text} {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(speed ~ track, data = coasters)
```

## [Side-by-Side Box-and-Whisker Plots]{.r-fit-text} {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(speed ~ track, data = coasters, horizontal = TRUE)
```

## [Side-by-Side Box-and-Whisker Plots]{.r-fit-text} {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(speed ~ track, data = coasters, horizontal = TRUE,
 col = "lightblue")
```

## [Side-by-Side Box-and-Whisker Plots]{.r-fit-text} {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(speed ~ track, data = coasters, horizontal = TRUE,
 col = "lightgreen",
 main = "Speed of World Roller Coasters")
```

## [Side-by-Side Box-and-Whisker Plots]{.r-fit-text} {auto-animate=true}
```{r}
#| echo: TRUE
boxplot(speed ~ track, data = coasters, horizontal = TRUE,
 col = "pink",
 main = "Speed of World Roller Coasters",
 xlab = "speed (miles per hour)",
 ylab = "type of track")
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
# Normal probability plot of the speeds for the steel tracks
qqnorm(coasters$speed[coasters$track=="Steel"])
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
# Normal probability plot of the speeds for the steel tracks
qqnorm(coasters$speed[coasters$track=="Steel"])
qqline(coasters$speed[coasters$track=="Steel"])
```


## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
# Normal probability plot of the speeds for the wooden tracks
qqnorm(coasters$speed[coasters$track=="Wood"])
```

## Normal Probability Plot {auto-animate=true}
```{r}
#| echo: TRUE
# Normal probability plot of the speeds for the wooden tracks
qqnorm(coasters$speed[coasters$track=="Wood"])
qqline(coasters$speed[coasters$track=="Wood"])
```

## [Side-by-Side Comparison of Normal Probability Plots]{.r-fit-text} {auto-animate=true}
```{r}
# Side-by-Side Comparison
par(mfrow = c(1,2))
qqnorm(coasters$speed[coasters$track=="Steel"])
qqline(coasters$speed[coasters$track=="Steel"])
qqnorm(coasters$speed[coasters$track=="Wood"])
qqline(coasters$speed[coasters$track=="Wood"])
par(mfrow = c(1,1))
```


## [Behind the Scenes of 't.test' function]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t(\mu_{0}) & = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}} \\[0.25cm]
\end{aligned}
$$

## [Behind the Scenes of 't.test' function]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t(\mu_{0}) & = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}} \\[0.25cm]
\text{We know,} \; \mu_{0} & = 0, \; \text{therefore:} \\[0.25cm]
\end{aligned}
$$

## [Behind the Scenes of 't.test' function]{.r-fit-text} {auto-animate=true}
**Test Statistic:**
$$
\begin{aligned}
t(\mu_{0}) & = \dfrac{(\bar{X} - \bar{Y}) - \mu_{0}}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}} \\[0.25cm]
\text{We know,} \; \mu_{0} & = 0, \; \text{therefore:} \\[0.25cm]
& = \dfrac{(\bar{X} - \bar{Y})}{\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}}
\end{aligned}
$$

## [Behind the Scenes of 't.test' function]{.r-fit-text} {transition="zoom" transition-speed="slow"}
**Welch-Satterthwaite Approximation**: to the null distribution of $t(\mu_{0})$
$$
\nu = \dfrac{\left(\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}\right)^{2}}{\dfrac{s^{4}_{X}}{m^{2}(m-1)} + \dfrac{s^{4}_{Y}}{n^{2}(n-1)}}
$$
Note: $\min(m-1, n-1) \le v \le m+n-2$

:::aside
Note: $\nu$ are the degrees of freedom
:::

## [Behind the Scenes of 't.test' function]{.r-fit-text} {transition="zoom" transition-speed="slow"}
**Confidence Interval:**
$$
(\bar{X} - \bar{Y}) \pm t_{(v)(1-\frac{\alpha}{2})}\sqrt{\dfrac{s^{2}_{X}}{m} + \dfrac{s^{2}_{Y}}{n}}
$$

:::aside
Remember, unless specified otherwise, R will make group 1 the one that comes alphabetically first (steel tracks in our case)
:::

## Using 't.test' function {auto-animate=true}
```{r}
#| echo: true
#| results: hide
t.test(speed ~ track,
       data = coasters)
```

## Using 't.test' function {auto-animate=true}
```{r}
#| echo: true
t.test(speed ~ track,
       data = coasters)
```

## Using 't.test' function {auto-animate=true}
```{r}
#| echo: true
#| results: hide
t.test(speed ~ track,
       data = coasters,
       mu = 0)
```

## Using 't.test' function {auto-animate=true}
```{r}
#| echo: true
#| results: hide
t.test(speed ~ track,
       data = coasters,
       mu = 0,
       alternative = "two.sided")
```

## Using 't.test' function {auto-animate=true}
```{r}
#| echo: true
#| results: hide
t.test(speed ~ track,
       data = coasters,
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.95)
```

## Using 't.test' function {auto-animate=true}
```{r}
#| echo: true
t.test(speed ~ track,
       data = coasters,
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.95)
```

## Using 't.test' function {auto-animate=true}
:::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: true
# Outputs are the same!
t.test(speed ~ track,
       data = coasters)
```
:::
::: {.column width="50%"}
```{r}
#| echo: true
# Outputs are the same!
t.test(speed ~ track,
       data = coasters,
       mu = 0,
       alternative = "two.sided",
       conf.level = 0.95)
```
:::
::::

## Answer Question of Interest
- There is not enough evidence to indicate the mean speed of roller coasters on steel tracks is different than the mean speed of roller coasters on wooden tracks (p-value = 0.3098).
- Confidence interval for $\mu_{steel} - \mu_{wood} = (-7.07 \; MPH, 2.28 \; MPH)$
- We’re 95% confident that the mean speed of roller coasters on steel tracks is between 7.07 MPH less to 2.28 MPH more than the mean speed of roller coasters on wooden tracks.
