---
title: "Transform with Caution: <br> When Statistical Magic Can Go Wrong"
author: The Guy From Lab
format:
  revealjs:
    incremental: TRUE
    scrollable: TRUE
    theme: dracula
    transition: slide
    transition-speed: slow
    auto-animate: TRUE
    auto-animate-easing: ease-in-out
    chalkboard: TRUE
    plugins:
      - reveal.js/plugin/zoom-js/zoom.js
      - reveal.js/plugin/highlight/highlight.js
    dependencies:
      - r: kableExtra
---

<style>
.slightlysmaller {
  font-size: 0.85em;   /* Slightly smaller than normal */
}

.evensmaller {
  font-size: 0.666em;   /* Slightly smaller than normal */
}

.bigger {
  font-size: 1.15em;   /* Slightly bigger than normal */
}

.biggest {
  font-size: 1.3em;    /* Larger, if needed */
}

.footnote {
  position: absolute;
  bottom: 1em;
  font-size: 0.8em;
  color: #555;
}
</style>


## Imagine This {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r, echo = FALSE}
set.seed(588190)
library(tidyverse)
library(gridExtra)
library(e1071)
library(knitr)
library(kableExtra)

# Need to generate at start so tests line up like I want
# Set Sample Sizes
n1 <- 1000
n2 <- 1000

# Generate samples
sampX <- 100 * rgamma(n1, shape = 0.58, scale = 2 / 0.58)
sampY <- 100 * exp(rnorm(n2, log(2) - 1, sqrt(2)))
```

- You're walking down the street and you see some commotion in your way, people are gathered around.
- As you approach you hear whispering, crying, some are even angry.
- Deciding to engage you speak to the calmest one available and inquire on what is going on, you come to find they're all discussing performing a t-test on some data.
- Then before you can say anything else you see the following lying, lifelessly on the ground...

## \*Gasp\* {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
ggplot(data = data.frame(vals = 100*rgamma(100, shape = 0.58, scale = 2/0.58)), aes(x = vals)) +
  geom_histogram(bins = 50, fill = "#f56056", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "A Skewed Histogram",
       subtitle = "A Tragic Tale") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
```

## Everyone Wants to be Normal
- Many methods require this Normality assumption
  - $t$-tests
  - ANOVA
  - Linear Regression, etc.
- Without Normality these methods are not exact (but they are asymptotically exact because they're based on sample means)
- The level of inexactness will depend on the population distribution(s) skewness, kurtosis, etc.

## Transforming Data {.smaller}
- To remedy situations of asymmetry some people (perhaps too many) suggest employing data transformations
  - A common data transformation is the log transformation ($\ln(x)$)
    - Often it's suggested for right-skewed positive data
    - One of the most abused data transformations
  - Some other transformations include
    - Square root transformation ($\sqrt{x}$)
    - Cube root transformation ($\sqrt[3]{x}$)
    - Logit transformation (sometimes for percent data) $\log\!\left(\frac{p}{1 - p}\right)$

## [Transforming Data - The Punchline]{.r-fit-text} {.smaller}
- Why caution on data transformations?
- Once transformation is performed we are almost never answering the original question
- If we are interested in say, population means on the original scale, performing a hypothesis test on population means of transformed data is going to answer a different question

::: {.fragment}
$$
E(g(T)) \neq g(E(T))
$$
:::

::: {.fragment}
Transforming inference back to the original scale is very challenging to interpret unless strong assumptions are made (stay tuned for more!)
:::

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
To illustrate my point, let's generate some data:
```{r}
#| echo: TRUE
#| eval: FALSE
# Set Sample Sizes
n1 <- 1000
n2 <- 1000

# Generate samples
sampX <- 100 * rgamma(n1, shape = 0.58, scale = 2 / 0.58)
sampY <- 100 * exp(rnorm(n2, log(2) - 1, sqrt(2)))
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
hist_sampX <- ggplot(data.frame(sampX), aes(x = sampX)) +
  geom_histogram(binwidth = 20, fill = "#87CEEB", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Sample X", x = "Sample X", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_sampY <- ggplot(data.frame(sampY), aes(x = sampY)) +
  geom_histogram(binwidth = 20, fill = "#98FB98", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Sample Y", x = "Sample Y", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
grid.arrange(hist_sampX, hist_sampY, ncol = 2)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
t.test(sampX, sampY)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.test(sampX, sampY) # Not significant: means are plausibly equal
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
hist_log_sampX <- ggplot(data.frame(log_sampX = log(sampX)), aes(x = log_sampX)) +
  geom_histogram(binwidth = 0.5, fill = "#1E90FF", color = "black", alpha = 0.7) +
  labs(title = "Histogram of log(Sample X)", x = "log(Sample X)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_log_sampY <- ggplot(data.frame(log_sampY = log(sampY)), aes(x = log_sampY)) +
  geom_histogram(binwidth = 0.5, fill = "#3CB371", color = "black", alpha = 0.7) +
  labs(title = "Histogram of log(Sample Y)", x = "log(Sample Y)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
grid.arrange(hist_log_sampX, hist_log_sampY, ncol = 2)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
t.test(log(sampX), log(sampY))
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.test(log(sampX), log(sampY)) # Significant: mean(log(X)) < mean(log(Y))
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
hist_sqrt_sampX <- ggplot(data.frame(sqrt_sampX = sqrt(sampX)), aes(x = sqrt_sampX)) +
  geom_histogram(binwidth = 2, fill = "#4682B4", color = "black", alpha = 0.7) +
  labs(title = "Histogram of sqrt(Sample X)", x = "sqrt(Sample X)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_sqrt_sampY <- ggplot(data.frame(sqrt_sampY = sqrt(sampY)), aes(x = sqrt_sampY)) +
  geom_histogram(binwidth = 2, fill = "#228B22", color = "black", alpha = 0.7) +
  labs(title = "Histogram of sqrt(Sample Y)", x = "sqrt(Sample Y)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))
grid.arrange(hist_sqrt_sampX, hist_sqrt_sampY, ncol = 2)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
#| eval: FALSE
t.test(sqrt(sampX), sqrt(sampY))
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
t.test(sqrt(sampX), sqrt(sampY)) # Significant: mean(sqrt(X)) > mean(sqrt(Y))
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Create the individual histograms
hist_sampX <- ggplot(data.frame(sampX), aes(x = sampX)) +
  geom_histogram(binwidth = 20, fill = "#87CEEB", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Sample X", x = "Sample X", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_sampY <- ggplot(data.frame(sampY), aes(x = sampY)) +
  geom_histogram(binwidth = 20, fill = "#98FB98", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Sample Y", x = "Sample Y", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_log_sampX <- ggplot(data.frame(log_sampX = log(sampX)), aes(x = log_sampX)) +
  geom_histogram(binwidth = 0.5, fill = "#1E90FF", color = "black", alpha = 0.7) +
  labs(title = "Histogram of log(Sample X)", x = "log(Sample X)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_log_sampY <- ggplot(data.frame(log_sampY = log(sampY)), aes(x = log_sampY)) +
  geom_histogram(binwidth = 0.5, fill = "#3CB371", color = "black", alpha = 0.7) +
  labs(title = "Histogram of log(Sample Y)", x = "log(Sample Y)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_sqrt_sampX <- ggplot(data.frame(sqrt_sampX = sqrt(sampX)), aes(x = sqrt_sampX)) +
  geom_histogram(binwidth = 2, fill = "#4682B4", color = "black", alpha = 0.7) +
  labs(title = "Histogram of sqrt(Sample X)", x = "sqrt(Sample X)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

hist_sqrt_sampY <- ggplot(data.frame(sqrt_sampY = sqrt(sampY)), aes(x = sqrt_sampY)) +
  geom_histogram(binwidth = 2, fill = "#228B22", color = "black", alpha = 0.7) +
  labs(title = "Histogram of sqrt(Sample Y)", x = "sqrt(Sample Y)", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

# Combine all six histograms into a single layout
grid.arrange(hist_sampX, hist_log_sampX, hist_sqrt_sampX,
             hist_sampY, hist_log_sampY, hist_sqrt_sampY, 
             nrow = 2)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- Our tests disagree... but why the underlying data is not being changed?
  - These three tests are addressing fundamentally different questions. The results need not coincide: it is entirely possible that:
    - the original scale means are equal, 
    - the mean of the log-transformed data suggests Group $Y$ have larger values than Group $X$,
    - the mean of the square-root–transformed data suggests Group $X$ have larger values than Group $Y$.
  - Tests on transformed data provide no direct information about comparisons on the original scale, and vice versa.
  - These tests are targeting distinct population parameters.

## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- I've started with discussing people's typical response to asymmetry / skewness. Let's clarify what we're talking about.
  - Suppose there is some random variable $X$ with mean $\mu$ and standard deviation $\sigma$
  - Skewness is defined as the standardized third central moment of a distribution:
  

::: {.fragment}
$$
\text{Population Skewness} = \frac{\kappa_{3}}{\kappa_{2}^{3/2}}
$$
:::

## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
$$
\text{Population Skewness} = \frac{\kappa_{3}}{\kappa_{2}^{3/2}}
$$


- Whoa, what are these funky symbols...
  - $\kappa_{3}$ is the third cumulant
  - $\kappa_{2}$ is the second cumulant
  - Lucky for us the first three cumulants are the same as the central moments, that is
  
::: {.fragment}
$$
\begin{align*}
  \kappa_{1} & = \mu = \operatorname{E}(X) \\[0.20cm]
  \kappa_{2} & = \operatorname{E}\left[(X - \mu)^{2}\right] = \sigma^{2} \\[0.20cm]
  \kappa_{3} & = \operatorname{E}\left[(X - \mu)^{3}\right] = \mu_{3} \\[0.20cm]
\end{align*}
$$
:::

- So I could have written it in a less scary way:

::: {.fragment}
$$
\text{Population Skewness} = \frac{\mu_{3}}{\sigma^{3}} = \frac{\operatorname{E}\left[(X - \mu)^{3}\right]}{\sigma^{3}}
$$
:::

## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
I also mentioned **kurtosis**, which measures how heavily the tails of a distribution differ from those of a normal distribution.

- It summarizes the **concentration of probability in the tails versus the center**.
- In practical terms:
  - **High kurtosis** → more extreme values (heavy tails)
  - **Low kurtosis** → fewer extreme values (light tails)
  
::: {.fragment}
Similar to skewness, kurtosis is the standardized fourth central moment
$$
\text{Population Kurtosis} = \frac{\mu_{4}}{\sigma^{4}} = \frac{\operatorname{E}\left[(X - \mu)^{4}\right]}{\sigma^{4}}
$$
:::

- It's a little more complicated so I'll refrain from going further into it... Continuing with skewness... (after some plots)


## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: FALSE
#| eval: TRUE
set.seed(131)
# High skewness: Exponential distribution
samp_skew <- rexp(5000, rate = 0.05)
hist_skew <- ggplot(data.frame(samp_skew), aes(x = samp_skew)) +
  geom_histogram(binwidth = 3, fill = "#93f542", color = "black", alpha = 0.7) +
  labs(title = "Histogram of High Skewness", x = "Value", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

# High kurtosis: t-distribution with low df (heavy tails)
samp_kurt <- rt(5000, df = 2)
hist_kurt <- ggplot(data.frame(samp_kurt), aes(x = samp_kurt)) +
  geom_histogram(binwidth = 3, fill = "#f5425a", color = "black", alpha = 0.7) +
  labs(title = "Histogram of High Kurtosis", x = "Value", y = "Frequency") +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

# Combine plots side by side
grid.arrange(hist_skew, hist_kurt, ncol = 2)
```


## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
There are many different sample skewness formulas, one of the most simple being:
  
::: {.fragment}
$$
\text{Sample Skewness} = \frac{1}{n} \sum_{i=1}^{n} \left( \frac{x_i - \bar{x}}{s} \right)^3
$$
:::

- where $s^{2} = \frac{1}{n - 1}\sum\limits_{i=1}^{n}\left(x_{i} - \bar{x}\right)^{2}$

- **Positive skew**: The right tail is longer or fatter than the left.
- **Negative skew**: The left tail is longer or fatter than the right.
- **Zero skew**: The distribution is symmetric.


## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r, fig.align = 'center'}
# Calculate skewness for each transformation
skewness_values <- data.frame(
  Transformation = c("Sample X", "Sample Y", "log(Sample X)", 
                     "log(Sample Y)", "sqrt(Sample X)", "sqrt(Sample Y)"),
  Skewness = c(skewness(sampX), 
               skewness(sampY),
               skewness(log(sampX)),
               skewness(log(sampY)),
               skewness(sqrt(sampX)),
               skewness(sqrt(sampY))))
skewness_values %>%
  kable(format = "markdown") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center",
    font_size = 30
  ) %>%
  column_spec(1, color = "white", background = "#333333") %>%
  column_spec(2, color = "lightblue", background = "#444444", bold = TRUE) %>%
  row_spec(0, background = "#222222", color = "white", bold = TRUE) %>%
  row_spec(1:nrow(skewness_values), color = "white", background = "#333333") %>%
  add_header_above(c("Skewness Values Before and After Transformations" = 2))
```

<div class="footnote fragment" style="position: absolute; bottom: 1em; font-size: 0.8em;">
   We see that absolute skewness is indeed smaller after transformation... but is this always the case?
</div>

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}

```{r}
#| echo: TRUE
# Simulate right-skewed data (Gamma distribution)
right_skewed_data <- rgamma(1000, shape = 2, scale = 2)

# Simulate left-skewed data (Inverse Gamma distribution)
left_skewed_data <- 1 / rgamma(1000, shape = 2, scale = 2)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Simulate right-skewed data (Gamma distribution)
right_skewed_data <- rgamma(1000, shape = 2, scale = 2)

# Simulate left-skewed data (Inverse Gamma distribution)
left_skewed_data <- 1 / rgamma(1000, shape = 2, scale = 2)

# Apply concave transformation (log) to both datasets
right_skewed_log <- log(right_skewed_data)
left_skewed_log <- log(left_skewed_data)
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
# Simulate right-skewed data (Gamma distribution)
right_skewed_data <- rgamma(1000, shape = 2, scale = 2)

# Simulate left-skewed data (Inverse Gamma distribution)
left_skewed_data <- 1 / rgamma(1000, shape = 2, scale = 2)

# Apply concave transformation (log) to both datasets
right_skewed_log <- log(right_skewed_data)
left_skewed_log <- log(left_skewed_data)

# Apply convex transformation (square) to both datasets
right_skewed_square <- right_skewed_data^2
left_skewed_square <- left_skewed_data^2
```

## [Disagreement Among the Little Ones]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Calculate skewness for each dataset before and after transformations
shooskew <- data.frame(
  Transformation = c("Right Skewed", "Left Skewed", "Log (Right Skewed)", 
                     "Log (Left Skewed)", "Square (Right Skewed)", "Square (Left Skewed)"),
  Skewness = c(
    skewness(right_skewed_data), 
    skewness(left_skewed_data),
    skewness(right_skewed_log),
    skewness(left_skewed_log),
    skewness(right_skewed_square),
    skewness(left_skewed_square)
  )
)

shooskew %>%
  kable(format = "markdown") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center",
    font_size = 30
  ) %>%
  column_spec(1, color = "white", background = "#333333") %>%
  column_spec(2, color = "lightblue", background = "#444444", bold = TRUE) %>%
  row_spec(0, background = "#222222", color = "white", bold = TRUE) %>%
  row_spec(1:nrow(shooskew), color = "white", background = "#333333") %>%
  add_header_above(c("Skewness Values Before and After Transformations" = 2))
```

## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- In General
- Concave functions (e.g., log, square root, cube root) tend to reduce skewness by compressing the extreme values
  - This in turn often results in a less skewed distribution
  - This is in line with Jensen's inequality, which states for a concave function $f$

::: {.fragment}
$$
f\left(E[X]\right) \ge E\left[f(X)\right]
$$
:::

## [Disagreement Among the Little Ones]{.r-fit-text} {.smaller auto-animate="true" auto-animate-easing="ease-in-out"}
- In General
- Convex functions (e.g., squaring, exponentiating) generally amplify skewness because they increase the effect of extreme values, 
  - For right-skewed data, this increases the positive skew, and for left-skewed data, it may reverse the skew (creating positive skew by squaring negative values)
    - An example of this follows
  - Again, this comes from Jensen's, which states for a convex function $f$

::: {.fragment}
$$
f\left(E[X]\right) \le E\left[f(X)\right]
$$
:::


## You had one job... {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Step 1: Create data with small skew
n <- 1000
samp <- rnorm(n)
samp <- samp - 2.5 * abs(rnorm(n))  # Normal distribution slightly negatively skewed

# Step 2: Apply a transformation (squaring the data to induce positive skew)
transformed_samp <- samp^2

# Create the original and transformed histograms
plot1 <- ggplot(data.frame(samp), aes(x = samp)) +
  geom_histogram(binwidth = 0.75, fill = "skyblue", color = "white") +
  labs(title = "Original Sample (Slightly Negative Skew)", x = "Values", y = "Frequency") +
  theme_minimal()

plot2 <- ggplot(data.frame(transformed_samp), aes(x = transformed_samp)) +
  geom_histogram(binwidth = 2, fill = "lightcoral", color = "white") +
  labs(title = "Transformed Sample (Very Positive Skew)", x = "Values", y = "Frequency") +
  theme_minimal()

# Arrange both plots side by side
grid.arrange(plot1, plot2, ncol = 2)
```

## You had one job... {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r, fig.align = 'center'}
# Step 3: Calculate skewness before and after transformation
Skewed <- data.frame(
  Type = c("Orginal", "Transformed"),
  Skewness = c(skewness(samp), skewness(transformed_samp)))
Skewed %>%
  kable(format = "markdown") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "responsive"),
    full_width = FALSE,
    position = "center",
    font_size = 30
  ) %>%
  column_spec(1, color = "white", background = "#333333") %>%
  column_spec(2, color = "lightblue", background = "#444444", bold = TRUE) %>%
  row_spec(0, background = "#222222", color = "white", bold = TRUE) %>%
  row_spec(1:nrow(Skewed), color = "white", background = "#333333") %>%
  add_header_above(c("Skewness Values Before and After Transformation" = 2))
```

- We see that the direction of skewness can change and it can even be increased by transformation as shown here

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text}
- As mentioned in the beginning, data transformations are often suggested for skewed data, data that "doesn't look Normal"
- A common transformation is the natural log
  - But, when we are interested in population means (as is often the case)...
- Testing equality of two population means on log-transformed data is equivalent to testing equality of the population geometric means on the original scale

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
\begin{align*}
\text{Let } & y_{i} = \log(x_{i}) \\[0.25cm]
\bar{y} & = \frac{1}{n} \sum_{i=1}^{n} y_{i}
\end{align*}

::: {.fragment}
\begin{align*}
            & = \frac{1}{n} \sum_{i=1}^{n} \log(x_{i})
\end{align*}
:::

::: {.fragment}
\begin{align*}
            & = \frac{1}{n} \left(\log(x_{1}), \log(x_{2}), \cdots, \log(x_{n})\right)
\end{align*}
:::

::: {.fragment}
\begin{align*}
            & = \frac{1}{n} \log \left( \prod_{i=1}^{n} x_{i}\right) && \Big(\log(a) + \log(b) = \log(ab)\Big)
\end{align*}
:::

::: {.fragment}
\begin{align*}
            & = \log\left(\left(\prod_{i=1}^{n} x_{i}\right)^{\frac{1}{n}}\right) && \Big(x\log(a) = \log(a^x)\Big)
\end{align*}
:::

::: {.fragment}
\begin{align*}
            & = \log\left(\text{GM}\right)
\end{align*}
:::

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
\begin{align*}
\text{Let } & y_{i} = \log(x_{i}) \\[0.25cm]
    \bar{y} & = \frac{1}{n} \sum_{i=1}^{n} y_{i} \\[0.25cm]
            & = \frac{1}{n} \sum_{i=1}^{n} \log(x_{i}) \\[0.25cm]
            & = \frac{1}{n} \left(\log(x_{1}), \log(x_{2}), \cdots, \log(x_{n})\right) \\[0.25cm]
            & = \frac{1}{n} \log \left( \prod_{i=1}^{n} x_{i}\right) && \Big(\log(a) + \log(b) = \log(ab)\Big) \\[0.25cm]
            & = \log\left(\left(\prod_{i=1}^{n} x_{i}\right)^{\frac{1}{n}}\right) && \Big(x\log(a) = \log(a^x)\Big) \\[0.25cm]
            & = \log\left(\text{GM}\right)
\end{align*}

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
- So we see that the arithmetic mean of the log-transformed data is the log of the geometric mean of the original data.
- Notice we can't just exponentiate to interpret sample mean on original scale

::: {.fragment}
\begin{align*}
    e^{\log\left(\left(\prod\limits_{i=1}^{n} x_{i}\right)^{\frac{1}{n}}\right)} = \left(\prod_{i=1}^{n} x_{i}\right)^{\frac{1}{n}} \neq \frac{1}{n} \sum_{i=1}^{n} x_{i}
\end{align*}
:::

- Exponentiating the sample mean on the log-scale doesn't give the mean on the original scale

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
- So, since

::: {.fragment}
\begin{align*}
    \frac{1}{n} \sum_{i=1}^{n} \log(x_{i}) \neq \log\left(\frac{1}{n} \sum_{i=1}^{n} x_{i}\right)
\end{align*}
:::

- By Jensen's inequality again


## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
- Suppose we are testing: $H_{0}: \delta_{X} = \delta_{Y}$, where
  - $\delta_{X}$ is the population mean for the natural log of $X$
  - $\delta_{Y}$ is the population mean for the natural log of $Y$
- this isn't going to tell us about the population means of $X$ and $Y$
  - Meaning it's not going to answer $H_{0}: \mu_{X} = \mu_{Y}$
  - What it tests is...
    - $H_{0}: \delta_{X} = \delta_{Y}$

- Some people try to get around this by interpreting a different quantity, the *median*
  - Why the median?

<div class="footnote fragment" style="position: absolute; bottom: 1em; font-size: 0.8em;">
   Notice, I didn't say $H_{0}: \ln\left(\mu_{X}\right) = \ln\left(\mu_{Y}\right)$, subtle difference
</div>

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
To see why, let's generate some data... again!
```{r}
#| echo: TRUE
some_data <- runif(9, 0, 100)
log_some_data <- log(some_data)
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
some_data <- runif(9, 0, 100)
log_some_data <- log(some_data)

Median_x <- median(some_data)
Median_x
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Median_x <- median(some_data)
Median_x
Median_log <- median(log_some_data)
Median_log
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
Median_x
Median_log

log(Median_x)
exp(Median_log)
```

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
#| echo: TRUE
c(Median_log, log(Median_x))
c(Median_x, exp(Median_log))
```

- We see that the median of the logs is the log of the medians

## [Interpreting Inference on Log-Transformed Data]{.r-fit-text} {.smaller}
- What is taught in ST 411/511 is the interpretation of a median on the original scale
- To make this interpretation we've got to make some big leaps in assumption land
  - If we assume the log-transformed data have a symmetric distribution, we know

::: {.fragment}
$$
\text{Mean}\left[\log(Y)\right] = \text{Median}\left[\log(Y)\right]
$$
:::
- And, as we have just demonstrated, the log preserves ordering meaning

::: {.fragment}
$$
\text{Median}\left[\log(Y)\right] = \log\left[\text{Median}(Y)\right]
$$
:::

## [Assuming Symmetry: The Statistical Shortcut We All Wish Worked]{.r-fit-text} {.evensmaller}
- So the logic goes as such:

::: {.fragment}
$$
\frac{1}{n}\sum\limits_{i=1}^{n}\log(X_{i}) - \frac{1}{m}\sum\limits_{i=1}^{m}\log(Y_{i}) = \bar{U} - \bar{W}
$$
:::

::: {.fragment}
Estimates
$$
\text{Mean}\left[\log(X)\right] - \text{Mean}\left[\log(Y)\right]
$$
:::

::: {.fragment}
If we assume the log-transformed data have a symmetric distribution
$$
\text{Median}\left[\log(X)\right] - \text{Median}\left[\log(Y)\right]
$$
:::

::: {.fragment}
Log preserves ordering 
$$
\log\left[\text{Median}(X)\right] - \log\left[\text{Median}(Y)\right] = \log\left(\frac{\text{Median}(X)}{\text{Median}(Y)}\right)
$$
:::

::: {.fragment}
Exponentiating we have
$$
e^{\bar{U} - \bar{W}} \quad \text{ estimates } \quad \frac{\text{Median}(X)}{\text{Median}(Y)}
$$
:::

## [Assuming Symmetry: The Statistical Shortcut We All Wish Worked]{.r-fit-text} {.slightlysmaller}
- Notice that symmetry is not technically required, we just need a setting where 
$$
\text{Mean}\left[\log(Y)\right] = \text{Median}\left[\log(Y)\right]
$$
symmetry just guarantees this property.
- To demonstrate this assumption, I will show the best case and a seemingly indistinguishable case to that of the best that does not work
- Assumption: If we have right skewed positive data that once transformed has a **symmetric distribution**

## [Let’s Pretend Everything is Symmetric: A Statistical Fantasy]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating right-skewed positive data (log-normal data)
da.data <- data.frame(vals = rlnorm(10000, mean = 0, sd = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Log-Normal Data)") +
  xlim(0, 20) +
  theme_bw()
```

## [Let’s Pretend Everything is Symmetric: A Statistical Fantasy]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Natural log of that data
ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
```

## [Let’s Pretend Everything is Symmetric: A Statistical Fantasy]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
p1 <- ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Log-Normal Data)") +
  xlim(0, 20) +
  theme_bw()

p2 <- ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#eb34de", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## [Assuming Symmetry: The Statistical Version of ‘Fake It Till You Make It]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- The right skewed data shown was log-normal so unless you assume that your data is log-normal you won't have completely Normal data (symmetric) once transformed
- Aka, just because it is positive and right-skewed, doesn't mean everything is golden
  - The couple following of slides will illustrate this
  
## [Let’s Just Assume Symmetry: What Could Go Wrong?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Generating right-skewed positive data (exponential data)
da.data <- data.frame(vals = rexp(10000, rate = 1))
ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Exponential)") +
  xlim(0, 20) +
  theme_bw()
```

## [Let’s Just Assume Symmetry: What Could Go Wrong?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Natural log of that data
ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
```

## [Let’s Just Assume Symmetry: What Could Go Wrong?]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
```{r}
# Plotting side-by-side
p3 <- ggplot(data = da.data, aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(X),
       y = "Frequency",
       title = "Right-Skewed Positive Data (Exponential)") +
  xlim(0, 20) +
  theme_bw()
p4 <- ggplot(data = log(da.data), aes(x = vals)) +
  geom_histogram(binwidth = 0.25, fill = "#ed6728", col = "white", alpha = 0.8) +
  labs(x = expression(ln(X)),
       y = "Frequency",
       title = "Exact Same Data, Log Transformed") +
  theme_bw()
grid.arrange(p3, p4, ncol = 2)
```

## [Symmetry in Statistics: It’s Like a Unicorn... We Hope It Exists]{.r-fit-text} {auto-animate="true" auto-animate-easing="ease-in-out"}
- We can see that since our original data was not log-normal we don't end up with a symmetric distribution

## Shift and Forget? {auto-animate="true" auto-animate-easing="ease-in-out"}
- Notice, I mentioned positively skewed data, we cannot have negative values before we log-transform (or the world will explode)
- Some people try to circumvent this by shifting the data beforehand but... 
  - then interpretation becomes challenging / hopeless and dependent on a perhaps arbitrary shift parameter

## [Same Cat, Different Hat for Linear Regression]{.r-fit-text} {.evensmaller}
- For linear regression, people often want to log transform the response if they suspect some non-normality of their residuals or want to address some nonlinearity

::: {.fragment}
Simple linear regression (SLR) setting
$$
E\left(Y \,|\, X\right) =\beta_{0} + \beta_{1}X
$$
:::

::: {.fragment}
Log transformation on $Y$
$$
E\left(\log(Y) \,|\, X\right) =\beta_{0} + \beta_{1}X
$$
:::

::: {.fragment}
We can exponentiate to interpret $Y$ on the original scale
$$
e^{E\left(\log(Y) \,|\, X\right)} = e^{\beta_{0} + \beta_{1}X}
$$
:::

::: {.fragment}
Back to our geometric mean on original scale
$$
\text{Geometric Mean}\left(Y \,|\, X\right) = e^{\beta_{0}}e^{\beta_{1}X}
$$
:::

## [Same Cat, Different Hat for Linear Regression]{.r-fit-text} {.smaller}
- Again we are faced with the mean of the log-transformed variable equal to the log of the geometric mean on the original scale
  - Thus a one unit increase in $X$ corresponds to a multiplicative increase of $e^{\beta_{1}}$ in the geometric mean of $Y$
- The *big assumption* of our data being symmetric after log transforming is sometimes made (although it probably shouldn't be) which leads to a slightly different interpretation
  - A one unit increase in $X$ corresponds to a multiplicative increase of $e^{\beta_{1}}$ in the median of $Y$

## [Same Cat, Different Hat for Linear Regression]{.r-fit-text} {.smaller}
- If instead we log transform our predictor:

::: {.fragment}
$$
E\left(Y \,|\, X\right) =\beta_{0} + \beta_{1}\log(X)
$$
:::
- We can interpret a one unit increase in $\log(X)$ as an associated increase in $Y$ by $\beta_{1}$
  - A one unit increase in $\log(X)$ is equivalent to a multiplicative increase of $e^{1}$ in $X$

## Summary {.smaller}
- Transformations change the question one is answering
  - Interpreting on original scale comes with complications of dealing with the geometric mean or make big assumptions about the data
- Transformations don't always fix the skewness problems
- Also note:
  - If plots and transformations are gone through prior to hypothesis testing:
    - A form of data snooping
    - A form of model selection
      - Need to be accounted for or forgone all-together

## Recommendations {.smaller}
- A data transformation is not always inappropriate...
  - If you're not interested in inference, only prediction
  - There are situations in which one might be instead in the transformed scale
    - Sound measurement in decibels follow a natural log scale
    - Richter scale (quantification of earthquake magnitude) is on a base-10 logarithmic scale
- I would caution making that symmetry assumption since it can't reliably be tested
- If you have an adequate sample size, I probably wouldn't worry about skewness 
  - What is considered "adequate" will depend on the skewness encountered
  - CLT is our secret sauce

## [From Any Shape to Bell Curve: The Magic of the Central Limit Theorem]{.r-fit-text} {.smaller}
```{r}
# Parameters
n_simulations <- 1000  # Number of simulations for each sample size
sample_sizes <- c(10, 30, 100)  # Sample sizes to test

# Generalized function to simulate sample means for any distribution
simulate_sample_means <- function(sample_size, dist_name, dist_params = list()) {
  dist_fun <- match.fun(paste0("r", dist_name))  # Find the R function for the distribution
  sample_means <- replicate(n_simulations, {
    mean(do.call(dist_fun, c(list(n = sample_size), dist_params)))  # Generate and calculate mean
  })
  data.frame(sample_means = sample_means, sample_size = as.factor(sample_size), dist_type = dist_name)
}

# Define the distributions and their parameters
distributions <- list(
  list(name = "unif", params = list(min = 0, max = 1)),
  list(name = "exp", params = list(rate = 1)),
  list(name = "norm", params = list(mean = 0, sd = 1)),
  list(name = "lnorm", params = list(meanlog = 0, sdlog = 1))
)

# Run simulations for each sample size and distribution type
sample_means_data <- bind_rows(
  lapply(distributions, function(dist) {
    bind_rows(lapply(sample_sizes, simulate_sample_means, dist_name = dist$name, dist_params = dist$params))
  })
)
# # Plot
# ggplot(sample_means_data, aes(x = sample_means, fill = sample_size)) +
#   geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.6, color = "black", position = "identity") +
#   geom_density(color = "black", size = 0.7, linetype = "dashed", alpha = 0.60) +
#   facet_grid(dist_type ~ sample_size, scales = "free",  # Allow x and y to scale freely
#              labeller = labeller(dist_type = c(
#                "unif" = "Unif",
#                "exp" = "Exp",
#                "norm" = "Normal",
#                "lnorm" = "Log-Norml"
#              ))) +
#   scale_fill_manual(values = c("#4B9CD3", "#74C476", "#f2715a")) +  # Custom colors for sample sizes
#   labs(
#     title = "Central Limit Theorem Demonstration across Distributions",
#     subtitle = "Effect of Sample Size on Sample Means",
#     x = "Sample Mean",
#     y = "Density",
#     fill = "Sample Size"
#   ) +
#   theme_minimal(base_size = 14) +
#   theme(
#     legend.position = "bottom",
#     strip.text = element_text(size = 12, face = "bold"),
#     plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5, size = 12),
#     panel.spacing = unit(1, "lines")
#   )

```

```{r}
ggplot(sample_means_data, aes(x = sample_means, fill = sample_size)) +
  geom_histogram(aes(y = ..density..), bins = 30, alpha = 0.4, color = "black", position = "identity") +
  geom_density(color = "black", size = 0.6, linetype = "dashed", alpha = 0.3) +
  facet_wrap(~ dist_type, scales = "free", strip.position = "left", 
             labeller = labeller(dist_type = c(
               "unif" = "Uniform",
               "exp" = "Exponential",
               "norm" = "Normal",
               "lnorm" = "Log-Normal"
             ))) +
  scale_fill_manual(values = c("#4B9CD3", "#74C476", "#f2715a")) +  # Custom colors for sample sizes
  labs(
    title = "Central Limit Theorem Demonstration across Distributions",
    subtitle = "Effect of Sample Size on Sample Means",
    x = "Sample Mean",
    y = "Density",
    fill = "Sample Size"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold", hjust = 1),  # Make distribution names bold
    strip.placement = "outside",
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    panel.spacing = unit(1, "lines"),
    panel.grid.major = element_line(color = "gray90")
  )
```

## The End
- Homework 5 is posted and when you will be able to show off your transformation knowledge
- I've got office hours 3:00 - 4:00pm Wednesday

## References {.smaller}
- Pek, J., Wong, O. & Wong, A. C., (2017) “Data Transformations for Inference with Linear Regression: Clarifications and Recommendations”, Practical Assessment, Research, and Evaluation 22(1): 9. doi: https://doi.org/10.7275/2w3n-0f07
- Pek, J., Wong, A.C.M. and Wong, O.C.Y. (2017) Confidence Intervals for the Mean of Non-Normal Distribution: Transform or Not to Transform. Open Journal of Statistics, 7, 405-421. https://doi.org/10.4236/ojs.2017.73029
- **The Real Deal Sarah Emerson (aka The Original)**